### 2023-06-05 slides/main.tex
sparse-condensed start

% \input{preamble}

\title{{\bfseries Sparse Condensed-Space Interior-Point Methods with Elastic Relaxations on GPUs: Will it Work?}}

\graphicspath{}

\author[Sungho Shin]{
  Sungho Shin\\
  {\normalfont\footnotesize\url{sshin@anl.gov}}
}
\subtitle{}
\institute[ANL/MCS]{
  Argonne National Laboratory
}

\begin{document}

{
  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{TitleANLBlue}}
  \frame[noframenumbering,plain]{
    \titlepage
    \vspace{-.7in}
  }
}

\begin{frame}{A Naive Approach Doesn't Work.}
  We solve the KKT system within IPM using sparse LU solver in \texttt{CUSOLVERRF}.

  \begin{itemize}
  \item Symbolic factorization needs to be performed every 
  \end{itemize}  
\end{frame}

\begin{frame}{Key Idea}
  Consider an NLP of the following form: 
  \begin{align*}
    \mathop{\text{min}}_{x^L\leq x\leq x^U}\;& f(x)\;\\
    \st\;&c(x) = 0.
  \end{align*}
  The problem is relaxed via {\it elastic relaxation}:
  \begin{align*}
    \mathop{\text{min}}_{x^L\leq x\leq x^U, s}\;& f(x)\;\\
    \st\;& 0\leq c(x) + s\leq \epsilon_{\text{ER}},
  \end{align*}
  where $\epsilon_{\text{ER}}\approx \sqrt{\epsilon_{\text{mach}}}$.

  We now observe that the problem has {\it inequality constriants only}. This allows us to apply the {\it condensation strategy}:
  \begin{align*}
    \begin{bmatrix}
      H + \Sigma_x & &J^\top \\
      & \Sigma_s &I \\
      J & I &\\
    \end{bmatrix}
    \begin{bmatrix}
      p^u \\ p^s \\ p^\lambda
    \end{bmatrix}  = -
    \begin{bmatrix}
      r_1 \\ r_2 \\ r_3 ,
    \end{bmatrix}
    \iff \left(H + \Sigma_x+ J^\top \Sigma_s J \right) p^u = -r_1 + J r_2 - J^\top \Sigma_s r_3.
  \end{align*}
  Of course, $J^\top \Sigma_s J$ can be arbitrarily dense, but our favorite problems will still be sparse.
\end{frame}

\begin{frame}{What do we need for Portabiltiy?}
  \begin{itemize}
  \item All we need is a sparse Cholesky solver (preferrably, written in {\tt KernelAbstractions.jl}).
  \end{itemize}
\end{frame}

\begin{frame}{Schur Complement as ``Bandwidth Thickening''}

\end{frame}

{
  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{TitleANLBlue}}
  \frame[noframenumbering,plain]{
    \titlepage
    \vspace{-.7in}
  }
}

\end{document}

### 2023-06-05 example.jl

@@ -1,0 +1,35 @@
+using PowerModels, MadNLP, MadNLPHSL, NLPModels, JuMP, SparseArrays, LinearAlgebra
+PowerModels.silence()
+
+pm = instantiate_model(
+    # "/home/sshin/git/pglib-opf/pglib_opf_case30_ieee.m",
+    # "/home/sshin/git/pglib-opf/pglib_opf_case300_ieee.m",
+    "/home/sshin/git/pglib-opf/pglib_opf_case1354_pegase.m",
+    # "/home/sshin/git/pglib-opf/pglib_opf_case2869_pegase.m",
+    # "/home/sshin/git/pglib-opf/pglib_opf_case9241_pegase.m",
+    # "/home/sshin/git/pglib-opf/pglib_opf_case2000_goc.m",
+    ACPPowerModel,
+    PowerModels.build_opf
+)
+
+set_optimizer(
+    pm.model,
+    () -> MadNLP.Optimizer(
+        linear_solver = Ma57Solver
+    )
+)
+optimize!(pm.model)
+
+nlp = pm.model.moi_backend.optimizer.model.nlp
+
+nlp.meta.ucon .+= 1e-8
+
+tol = 1e-6
+
+solver = MadNLPSolver(
+    nlp;
+    # linear_solver=Ma57Solver,
+    linear_solver = MadNLPCUSOLVER.RFSolver
+)
+
+solve!(solver)

### 2023-06-27 abstract/main.tex
pscc abstract added

\input{preamble} 
\title{\Large Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}

\author{Sungho Shin$^{1}$, Francois Pacaud$^2$, and Mihai Anitescu$^{1}$}
\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\thispagestyle{empty}

This work presents an accelerated solution method for solving AC optimal power flow (OPF) problems on graphics processing units (GPUs). While GPUs have demonstrated impressive accelerated capabilities in various computing domains, thus far they have shown limited capabilities in large-scale, constrained nonlinear programming regime, due to the challenging nature of parallel factorization of indefinite sparse matrices commonly encountered within interior-point methods (IPMs) \cite{anitescu2021targeting}. Although GPU computation can accelerate various other components of the optimization process, such as automatic differentiation and sparse matrix-vector multiplications, slow data transfer between host and device memory impedes the ad-hoc implementation of GPU accelerations. Thus, to fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework, including automatic differentiation, linear algebra, and optimization, on the GPU, minimizing data transfers to/from host memory.

In this work, we present a computational framework and associated software packages—MadNLP.jl\footnote{\url{https://github.com/MadNLP/MadNLP.jl}} and SIMDiff.jl\footnote{\url{https://github.com/sshin23/SIMDiff.jl}}—for solving AC OPF problems. Our approach leverages (i) condensed-space IPMs, (ii) inequality relaxation strategy, (iii) sparse matrix factorization with a fixed pivot sequence, and (iv) SIMD abstraction of nonlinear programs. In particular, our method relaxes power flow equality constraints by allowing small violations, which allows expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space. While this condensation strategy is not a new concept \cite{nocedal2006numerical}, it has been considered less efficient compared to the standard full-space method, largely due to the significant increase in the nonzero entries. However, when implemented on GPUs, it has the key advantage of a guaranteed positive definite condensed KKT system upon the application of standard regularization techniques. This allows for the application of linear solvers with a fixed numerical pivot sequence (an efficient implementation available in CUDA), enabling efficient solution of KKT systems on GPUs. Although this method is prone to ill-conditioning, our results indicate that the solver is robust enough to solve problems up to a relative accuracy of $10^{-6}$. Furthermore, the model functions and derivative evaluations can be parallelized by exploiting the single-instruction, multiple-data (SIMD) abstraction of nonlinear programs, which aims to preserve the parallelizable structure in the model to facilitate the evaluations on the GPU. We demonstrate that the AC power flow model is particularly suitable for such an abstraction, as it involves repetitive expressions for each component of the models (e.g., buses, lines, generators), and the number of computational patterns does not grow with the network's size. Exploiting this structure provides significant acceleration of model function and derivative evaluations.

We implement the entire process in our optimization solver, MadNLP.jl, and automatic differentiation tool, SIMDiff.jl, while the solution of the condensed KKT system is carried out using the external CUSOLVER library. Our numerical experiments reveal that the proposed framework has the potential to accelerate the solution of AC OPF problems by an order of magnitude compared to existing tools, such as Ipopt interfaced with MATPOWER or PowerModels.jl, especially for large-scale instances. We will present comprehensive numerical results demonstrating the effectiveness of our accelerated OPF framework. Our method can be extended further and help solve more complex variants of OPF problems, such as multi-period, security-constrained, joint transmission-distribution problems.

\pagebreak
\thispagestyle{empty}
\bibliographystyle{plain}
\bibliography{main}

\end{document}

### 2023-06-27 slides/main.tex
pscc abstract added

@@
 
-\title{{\bfseries Sparse Condensed-Space Interior-Point Methods with Elastic Relaxations on GPUs: Will it Work?}}
+\title{{\bfseries Sparse Condensed-Space Interior-Point Methods with Inequality Relaxations on GPUs: Will it Work?}}
 
@@
 
-\begin{frame}{A Naive Approach Doesn't Work.}
-  We solve the KKT system within IPM using sparse LU solver in \texttt{CUSOLVERRF}.
-
+\begin{frame}{Executive Summary}
   \begin{itemize}
-  \item Symbolic factorization needs to be performed every 
+  \item We exmaine the potential benefit of sparse condensed-space interior-point methods with inequality relaxations, with the goal of solving general large-scale NLPs on GPUs.
+  \item The proposed method relaxes the equality constraints as inequality constraints by replacing $c(x) =0$ by $c(x)+ s=0$ and $0\leq s\leq\epsilon_{\text{IR}}$.
+  \item The resulting inequality-constrained NLP is solved with sparse condensed-space interior-point method, which requires solving a sparse positive definite system.
+  \item While the LU solver in CUSOLVERRF cannot handle the sparse indefinite systems wihtin IPM, it can handle the sparse PD systems up to a certain accuracy ($\epsilon_{\text{IR}}=\texttt{tol} = 10^{-3}$).
+  \item For portability, we will still need a sparse Cholesky solver running on GPU.
   \end{itemize}
@@
 
-\begin{frame}{Key Idea}
+\begin{frame}{A Naive Approach Doesn't Work.}
+  \begin{itemize}
+  \item Can we solve the indefinite KKT systems using sparse LU solver in \texttt{CUSOLVERRF}? No.
+    \begin{center}
+      \includegraphics[width=.5\textwidth]{../fig/linear_solvers.pdf}  
+    \end{center}   
+  \item Symbolic factorization needed every several iterations (every 3 for above case).
+  \item Numeric factorization is not fast.
+  \end{itemize}  
+\end{frame}
+
+\begin{frame}{Condensed Approach}
   Consider an NLP of the following form: 
@@
   \end{align*}
-  The problem is relaxed via {\it elastic relaxation}:
+  The problem is relaxed via {\it inequality relaxation}:
   \begin{align*}
-    \mathop{\text{min}}_{x^L\leq x\leq x^U, s}\;& f(x)\;\\
-    \st\;& 0\leq c(x) + s\leq \epsilon_{\text{ER}},
+    \mathop{\text{min}}_{x^L\leq x\leq x^U, 0\leq s \leq \epsilon_{\text{IR}}}\;& f(x)\;\\
+    \st\;& c(x) + s= 0,
   \end{align*}
-  where $\epsilon_{\text{ER}}\approx \sqrt{\epsilon_{\text{mach}}}$.
+  where $\epsilon_{\text{IR}}\approx\texttt{tol}$.
 
@@
   \end{align*}
-  Of course, $J^\top \Sigma_s J$ can be arbitrarily dense, but our favorite problems will still be sparse.
+  Of course, $J^\top \Sigma_s J$ can be arbitrarily dense, but our favorite problems (e.g., OPF) will still be sparse.
+\end{frame}
+
+\begin{frame}{Effects of Inequality Relaxation}
+  \includegraphics[width=.32\textwidth]{../fig/iter.pdf}
+  \includegraphics[width=.32\textwidth]{../fig/objval.pdf}
+  \includegraphics[width=.32\textwidth]{../fig/cvio.pdf}
+  \begin{itemize}
+  \item $\epsilon_{\text{IR}}=10^{-8}$
+  \item Number of iterations increased (5--50\%).
+  \item Not much difference in the quality of solution (obj value/constraint violation).
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Will it be Efficient?}
+  \begin{itemize}
+  \item We do this experiment to better evaluate the potential benefit of the proposed strategy.
+  \item We take the KKT system from the iterates obtained from MadNLP+Ma57 and solve it with CUSOLVER. We aim to check:
+    \begin{itemize}
+    \item Symbolic factorization can be reused between iterations.
+    \item Numeric factorization/backsolve is fast/precise enough. 
+    \end{itemize}
+    \begin{center}
+      \includegraphics[width=.5\textwidth]{../fig/hypo.pdf}  
+    \end{center}
+  \item Observation: factorization/backsolve is precise/efficient enough for $\epsilon_{\text{IR}}=\texttt{tol} = 10^{-3}$.
+  \end{itemize}
 \end{frame}
@@
   \begin{itemize}
-  \item All we need is a sparse Cholesky solver (preferrably, written in {\tt KernelAbstractions.jl}).
+  \item For now, the sparse LU solver in CUSOLVERRF can do the job.
+  \item For portability, all we need is a sparse Cholesky solver (preferrably, written in {\tt KernelAbstractions.jl}).
   \end{itemize}
@@
 
-\begin{frame}{Schur Complement as ``Bandwidth Thickening''}
+% \begin{frame}{Schur Complement as ``Bandwidth Thickening''}
+%   \begin{itemize}
+%   \item Directly solving large-scale 
+%   \item Can be applied to any graph-structured sparse systems.
+%   \end{itemize}
+% \end{frame}
 
+% \def\checkmark{$\text{\rlap{$\checkmark$}}\square$} 
+\begin{frame}{Current Status (6/27/2023)}
+  \begin{itemize}
+  \item SparseCondensedKKTSystem works on CPU.
+  \item With inequality relaxation, can solve {\tt case9241} up to ${\tt tol} = 10^{-7}$, but requires long iterative refinement. Convergence up to $10^{-5}$ is reliable/efficient.
+  \item Numerical results (w/ linear algebra/AD on GPU) for {\tt case9241}:
+    \begin{center}
+      \includegraphics[width=.8\textwidth]{result.png}
+    \end{center}
+    \begin{itemize}
+    \item Linear Sovler: only 0.708 secs out of 4.408 secs are spent on GPU.
+    \item AD: only 0.104 secs out of 0.660 secs are spent on GPU.
+    \item MadNLP: operations like $y \leftarrow y + ax$ are $\times 10$ faster on GPUs.
+    \end{itemize}  
+  \item By running everything on GPUs, we may solve {\tt case9241} in less than 4 secs (JuMP+Ipopt takes 40 secs).
+  \item The group in LLNL may be looking at the same problem (based on their activities on Github).
+  \item Implement the GPU version (2-3 weeks of work) and submit this to for PSCC next year?
+\end{itemize}
 \end{frame}

### 2023-06-27 example.jl

@@ -1,35 +1,0 @@
-using PowerModels, MadNLP, MadNLPHSL, NLPModels, JuMP, SparseArrays, LinearAlgebra
-PowerModels.silence()
-
-pm = instantiate_model(
-    # "/home/sshin/git/pglib-opf/pglib_opf_case30_ieee.m",
-    # "/home/sshin/git/pglib-opf/pglib_opf_case300_ieee.m",
-    "/home/sshin/git/pglib-opf/pglib_opf_case1354_pegase.m",
-    # "/home/sshin/git/pglib-opf/pglib_opf_case2869_pegase.m",
-    # "/home/sshin/git/pglib-opf/pglib_opf_case9241_pegase.m",
-    # "/home/sshin/git/pglib-opf/pglib_opf_case2000_goc.m",
-    ACPPowerModel,
-    PowerModels.build_opf
-)
-
-set_optimizer(
-    pm.model,
-    () -> MadNLP.Optimizer(
-        linear_solver = Ma57Solver
-    )
-)
-optimize!(pm.model)
-
-nlp = pm.model.moi_backend.optimizer.model.nlp
-
-nlp.meta.ucon .+= 1e-8
-
-tol = 1e-6
-
-solver = MadNLPSolver(
-    nlp;
-    # linear_solver=Ma57Solver,
-    linear_solver = MadNLPCUSOLVER.RFSolver
-)
-
-solve!(solver)

### 2023-06-28 abstract/main.tex
add PSCC template

@@
 
-\author{Sungho Shin$^{1}$, Francois Pacaud$^2$, and Mihai Anitescu$^{1}$}
+% \author{Sungho Shin$^{1}$, Francois Pacaud$^2$, and Mihai Anitescu$^{1}$}
+\author{\IEEEauthorblockN{Sungho Shin\IEEEauthorrefmark{1},
+François Pacaud\IEEEauthorrefmark{2},
+Mihai Anitescu\IEEEauthorrefmark{1},
+\IEEEauthorblockA{\IEEEauthorrefmark{1} Mathematics and Computer Science Division,
+  Argonne National Laboratory \\
+}
+\IEEEauthorblockA{\IEEEauthorrefmark{2} Centre Automatique et Systèmes, Mines Paris - PSL \\
+}
+}
+}
 \date{\small
@@
 \maketitle
-\thispagestyle{empty}
 
+\begin{abstract}
 This work presents an accelerated solution method for solving AC optimal power flow (OPF) problems on graphics processing units (GPUs).
@@
 Although GPU computation can accelerate various other components of the optimization process, such as automatic differentiation and sparse matrix-vector multiplications, slow data transfer between host and device memory impedes the ad-hoc implementation of GPU accelerations.
-Thus, to fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework, including automatic differentiation, linear algebra, and optimization, on the GPU, minimizing data transfers to/from host memory.
+Thus, to fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU --- including automatic differentiation, linear algebra, and optimization --- minimizing data transfers from/to host memory.
+\end{abstract}
 
-In this work, we present a computational framework and associated software packages—MadNLP.jl\footnote{\url{https://github.com/MadNLP/MadNLP.jl}} and SIMDiff.jl\footnote{\url{https://github.com/sshin23/SIMDiff.jl}}—for solving AC OPF problems.
+In this work, we present a computational framework with its associated software packages—MadNLP.jl\footnote{\url{https://github.com/MadNLP/MadNLP.jl}} for the IPM solver and SIMDiff.jl\footnote{\url{https://github.com/sshin23/SIMDiff.jl}} for the automatic differentiation—for solving AC OPF problems.
 Our approach leverages (i) condensed-space IPMs, (ii) inequality relaxation strategy, (iii) sparse matrix factorization with a fixed pivot sequence, and (iv) SIMD abstraction of nonlinear programs.
-In particular, our method relaxes power flow equality constraints by allowing small violations, which allows expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space.
-While this condensation strategy is not a new concept \cite{nocedal2006numerical}, it has been considered less efficient compared to the standard full-space method, largely due to the significant increase in the nonzero entries.
-However, when implemented on GPUs, it has the key advantage of a guaranteed positive definite condensed KKT system upon the application of standard regularization techniques.
+Notably, our method relaxes power flow equality constraints by allowing small violations, which allows expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space.
+While this condensation strategy is not a new concept \cite{nocedal2006numerical}, it has been considered less efficient compared to the standard full-space method, largely due to increased fill-in in the sparse factorization.
+When implemented on GPUs, however, it has the key advantage of guaranteeing a positive definite condensed KKT system upon the application of standard regularization techniques.
 This allows for the application of linear solvers with a fixed numerical pivot sequence (an efficient implementation available in CUDA), enabling efficient solution of KKT systems on GPUs.
@@
 
-We implement the entire process in our optimization solver, MadNLP.jl, and automatic differentiation tool, SIMDiff.jl, while the solution of the condensed KKT system is carried out using the external CUSOLVER library.
-Our numerical experiments reveal that the proposed framework has the potential to accelerate the solution of AC OPF problems by an order of magnitude compared to existing tools, such as Ipopt interfaced with MATPOWER or PowerModels.jl, especially for large-scale instances.
+We implement the entire process in our optimization solver, MadNLP.jl, and automatic differentiation tool, SIMDiff.jl.
+Only the solution of the condensed KKT system is carried out using the external CUSOLVER library.
+Our numerical experiments reveal that the proposed framework has the potential to accelerate the solution of AC OPF problems by an order of magnitude compared to existing tools, such as Ipopt (interfaced with MATPOWER or PowerModels.jl) especially for large-scale instances.
 We will present comprehensive numerical results demonstrating the effectiveness of our accelerated OPF framework.
@@
 
-\pagebreak
-\thispagestyle{empty}
 \bibliographystyle{plain}

### 2023-06-29 abstract/main.tex
abstract edited

--- 2023-06-27 abstract/main.tex
@@
 \input{preamble}
-\title{\Large Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}
+\title{Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}
 
-\author{Sungho Shin$^{1}$, Francois Pacaud$^2$, and Mihai Anitescu$^{1}$}
+\author{\IEEEauthorblockN{Sungho Shin\IEEEauthorrefmark{1},
+François Pacaud\IEEEauthorrefmark{2},
+Mihai Anitescu\IEEEauthorrefmark{1},
+\IEEEauthorblockA{\IEEEauthorrefmark{1}Mathematics and Computer Science Division,
+  Argonne National Laboratory \\
+}
+\IEEEauthorblockA{\IEEEauthorrefmark{2}Centre Automatique et Systèmes, Mines Paris - PSL \\
+}
+}
+}
 \date{\small
@@
 \maketitle
-\thispagestyle{empty}
 
-This work presents an accelerated solution method for solving AC optimal power flow (OPF) problems on graphics processing units (GPUs).
-While GPUs have demonstrated impressive accelerated capabilities in various computing domains, thus far they have shown limited capabilities in large-scale, constrained nonlinear programming regime, due to the challenging nature of parallel factorization of indefinite sparse matrices commonly encountered within interior-point methods (IPMs) \cite{anitescu2021targeting}.
-Although GPU computation can accelerate various other components of the optimization process, such as automatic differentiation and sparse matrix-vector multiplications, slow data transfer between host and device memory impedes the ad-hoc implementation of GPU accelerations.
-Thus, to fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework, including automatic differentiation, linear algebra, and optimization, on the GPU, minimizing data transfers to/from host memory.
+While GPUs have demonstrated impressive capabilities in various computing domains, their effectiveness in large-scale, constrained nonlinear optimization regimes, such as AC OPFs, has been limited due to the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms \cite{anitescu2021targeting}.
+Although GPU computation can accelerate various other parts of the optimization process in a straightforward manner, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations.
+Thus, to fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU, incorporating AD, linear algebra, and optimization, while minimizing data transfers to and from host memory.
 
-In this work, we present a computational framework and associated software packages—MadNLP.jl\footnote{\url{https://github.com/MadNLP/MadNLP.jl}} and SIMDiff.jl\footnote{\url{https://github.com/sshin23/SIMDiff.jl}}—for solving AC OPF problems.
-Our approach leverages (i) condensed-space IPMs, (ii) inequality relaxation strategy, (iii) sparse matrix factorization with a fixed pivot sequence, and (iv) SIMD abstraction of nonlinear programs.
-In particular, our method relaxes power flow equality constraints by allowing small violations, which allows expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space.
-While this condensation strategy is not a new concept \cite{nocedal2006numerical}, it has been considered less efficient compared to the standard full-space method, largely due to the significant increase in the nonzero entries.
-However, when implemented on GPUs, it has the key advantage of a guaranteed positive definite condensed KKT system upon the application of standard regularization techniques.
-This allows for the application of linear solvers with a fixed numerical pivot sequence (an efficient implementation available in CUDA), enabling efficient solution of KKT systems on GPUs.
-Although this method is prone to ill-conditioning, our results indicate that the solver is robust enough to solve problems up to a relative accuracy of $10^{-6}$.
-Furthermore, the model functions and derivative evaluations can be parallelized by exploiting the single-instruction, multiple-data (SIMD) abstraction of nonlinear programs, which aims to preserve the parallelizable structure in the model to facilitate the evaluations on the GPU.
-We demonstrate that the AC power flow model is particularly suitable for such an abstraction, as it involves repetitive expressions for each component of the models (e.g., buses, lines, generators), and the number of computational patterns does not grow with the network's size.
-Exploiting this structure provides significant acceleration of model function and derivative evaluations.
+We present a comprehensive computational framework and the associated software implementations---an optimization solver MadNLP.jl and an AD tool SIMDiff.jl---for solving AC OPF problems on GPUs \cite{madnlp, simdiff}.
+Our approach utilizes the following techniques: (i) condensed-space interior-point methods (IPMs), (ii) an inequality relaxation strategy, (iii) sparse matrix factorization with a fixed pivot sequence, and (iv) a single-instruction, multiple-data (SIMD) abstraction of nonlinear programs.
+Notably, our method relaxes power flow equality constraints by allowing small violations, which enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space (referred to as the condensation strategy).
+Although this strategy is not new \cite{nocedal2006numerical}, it has traditionally been considered less efficient than the standard full-space method due to increased fill-in in the sparse factorization.
+However, when implemented on GPUs, it offers the key advantage of guaranteeing positive definiteness in the condensed KKT system through the application of standard regularization techniques.
+This allows for the utilization of linear solvers with a fixed numerical pivot sequence (an efficient implementation available as part of CUDA library), facilitating efficient KKT system solutions on GPUs.
+Although this method is susceptible to ill-conditioning, our results demonstrate that the solver is robust enough to solve problems with a relative accuracy of up to $10^{-6}$.
 
-We implement the entire process in our optimization solver, MadNLP.jl, and automatic differentiation tool, SIMDiff.jl, while the solution of the condensed KKT system is carried out using the external CUSOLVER library.
-Our numerical experiments reveal that the proposed framework has the potential to accelerate the solution of AC OPF problems by an order of magnitude compared to existing tools, such as Ipopt interfaced with MATPOWER or PowerModels.jl, especially for large-scale instances.
-We will present comprehensive numerical results demonstrating the effectiveness of our accelerated OPF framework.
-Our method can be extended further and help solve more complex variants of OPF problems, such as multi-period, security-constrained, joint transmission-distribution problems.
+Furthermore, by leveraging the SIMD abstraction of nonlinear programs, which preserves the parallelizable structure in the model, the model functions and derivative evaluations can be parallelized, facilitating evaluations on the GPU.
+We show that the AC power flow model is particularly well-suited for this abstraction, as it involves repetitive expressions for each component of the model (e.g., buses, lines, generators), and the number of computational patterns does not increase with the network's size.
+As observed by the impressive performance of Gravity \cite{hijazi2018gravity}, exploiting this structure significantly accelerates model function and derivative evaluations.
+
+We will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach.
+The proposed method is implemented in our packages, MadNLP.jl and SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library.
+We compare our method against the standard CPU approach \cite{wachter2006implementation} and the recently developed reduced-space interior-point method on GPUs \cite{pacaud2022condensed} using the standard benchmark library \cite{pglib}.
+Our initial results suggest that the proposed framework has the potential to accelerate the solution of AC OPF problems by an order of magnitude compared to existing tools, such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances, up to moderate tolerances ($10^{-6}$).
+We also discuss the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.
 
 \pagebreak
-\thispagestyle{empty}
+
 \bibliographystyle{plain}

### 2023-06-30 abstract/main.tex
minor formatting

@@
 The proposed method is implemented in our packages, MadNLP.jl and SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library.
-We compare our method against the standard CPU approach \cite{wachter2006implementation} and the recently developed reduced-space interior-point method on GPUs \cite{pacaud2022condensed} using the standard benchmark library \cite{pglib}.
+We compare our method using the standard benchmark library \cite{pglib} against the standard CPU approach \cite{wachter2006implementation} and the recently developed reduced-space interior-point method on GPUs \cite{pacaud2022condensed} .
 Our initial results suggest that the proposed framework has the potential to accelerate the solution of AC OPF problems by an order of magnitude compared to existing tools, such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances, up to moderate tolerances ($10^{-6}$).
@@
 
-\pagebreak
+\newpage
 

### 2023-06-30 abstract/main.tex
merged

@@
-\input{preamble}
+% \input{preamble}
 \title{Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}
 
-\author{\IEEEauthorblockN{Sungho Shin\IEEEauthorrefmark{1},
-François Pacaud\IEEEauthorrefmark{2},
-Mihai Anitescu\IEEEauthorrefmark{1},
-\IEEEauthorblockA{\IEEEauthorrefmark{1}Mathematics and Computer Science Division,
-  Argonne National Laboratory \\
+\author{
+\IEEEauthorblockN{Sungho Shin and Mihai Anitescu}
+\IEEEauthorblockA{
+  Mathematics and Computer Science Division\\
+  Argonne National Laboratory\\
+  Lemont, IL, USA\\
+sshin@anl.gov, anitescu@mcs.anl.gov}
+\and
+\IEEEauthorblockN{François Pacaud}
+\IEEEauthorblockA{Centre Automatique et Systèmes\\
+  Mines Paris - PSL \\
+  Paris, France\\
+francois.pacaud@minesparis.psl.eu}
 }
-\IEEEauthorblockA{\IEEEauthorrefmark{2}Centre Automatique et Systèmes, Mines Paris - PSL \\
-}
-}
-}
+
 \date{\small
@@
 
-While GPUs have demonstrated impressive capabilities in various computing domains, their effectiveness in large-scale, constrained nonlinear optimization regimes, such as AC OPFs, has been limited due to the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms \cite{anitescu2021targeting}.
-Although GPU computation can accelerate various other parts of the optimization process in a straightforward manner, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations.
-Thus, to fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU, incorporating AD, linear algebra, and optimization, while minimizing data transfers to and from host memory.
+While graphics processing units (GPUs) have showcased impressive capabilities in various computing domains, their utilization in large-scale constrained nonlinear optimization regimes, such as alternating current (AC) optimal power flow (OPF) problems, has been somewhat limited.
+This limitation stems from the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms (Anitescu et.
+al.
+2021).
+Although GPU computation can accelerate various other parts of the optimization process, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations.
+To fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU that incorporates AD, linear algebra, and optimization while minimizing data transfers to and from host memory.
 
-We present a comprehensive computational framework and the associated software implementations---an optimization solver MadNLP.jl and an AD tool SIMDiff.jl---for solving AC OPF problems on GPUs \cite{madnlp, simdiff}.
-Our approach utilizes the following techniques: (i) condensed-space interior-point methods (IPMs), (ii) an inequality relaxation strategy, (iii) sparse matrix factorization with a fixed pivot sequence, and (iv) a single-instruction, multiple-data (SIMD) abstraction of nonlinear programs.
-Notably, our method relaxes power flow equality constraints by allowing small violations, which enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space (referred to as the condensation strategy).
-Although this strategy is not new \cite{nocedal2006numerical}, it has traditionally been considered less efficient than the standard full-space method due to increased fill-in in the sparse factorization.
-However, when implemented on GPUs, it offers the key advantage of guaranteeing positive definiteness in the condensed KKT system through the application of standard regularization techniques.
-This allows for the utilization of linear solvers with a fixed numerical pivot sequence (an efficient implementation available as part of CUDA library), facilitating efficient KKT system solutions on GPUs.
-Although this method is susceptible to ill-conditioning, our results demonstrate that the solver is robust enough to solve problems with a relative accuracy of up to $10^{-6}$.
+This paper presents a comprehensive computational framework and the associated software implementations for solving AC OPF problems on GPUs.
+Our approach utilizes the following techniques: (i) condensed-space interior-point methods (IPMs) with an inequality relaxation strategy, (ii) sparse matrix factorization with a fixed pivot sequence, and (iii) a single-instruction, multiple-data (SIMD) abstraction of nonlinear programs.
+Specifically, our method relaxes power flow equality constraints by allowing small violations, which enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space through the condensation procedure.
+Although this strategy is not new (see Nocedal and Wright, 2006), it has traditionally been considered less efficient than the standard full-space method due to increased fill-in in the sparse factorization.
+However, when implemented on GPUs, it offers the key advantage of guaranteeing positive definiteness in the condensed KKT system through the application of standard regularization techniques, which in turn, allows for the utilization of linear solvers with a fixed numerical pivot sequence (so-called refactorization).
+An efficient implementation of the sparse refactorization is available as part of the CUDA library, facilitating the implementation of efficient KKT system solutions on GPUs.
+Although this method is susceptible to ill-conditioning, our results demonstrate that the solver is robust enough to solve problems with a relative accuracy of $10^{-6}$.
 
-Furthermore, by leveraging the SIMD abstraction of nonlinear programs, which preserves the parallelizable structure in the model, the model functions and derivative evaluations can be parallelized, facilitating evaluations on the GPU.
-We show that the AC power flow model is particularly well-suited for this abstraction, as it involves repetitive expressions for each component of the model (e.g., buses, lines, generators), and the number of computational patterns does not increase with the network's size.
-As observed by the impressive performance of Gravity \cite{hijazi2018gravity}, exploiting this structure significantly accelerates model function and derivative evaluations.
+Furthermore, by leveraging the SIMD abstraction of nonlinear programs, which preserves the parallelizable structure in the model, the model functions and derivative evaluations can be parallelized, thereby facilitating evaluations on the GPU.
+We demonstrate that the AC power flow model is particularly well-suited for this abstraction as it involves repetitive expressions for each type of component in the model (e.g., buses, lines, generators), and the number of computational patterns does not increase with the network's size.
+This structure has been effectively utilized by Gravity (Hijazi et.
+al.
+2018), demonstrating significant acceleration in model evaluations.
 
-We will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach.
-The proposed method is implemented in our packages, MadNLP.jl and SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library.
-We compare our method using the standard benchmark library \cite{pglib} against the standard CPU approach \cite{wachter2006implementation} and the recently developed reduced-space interior-point method on GPUs \cite{pacaud2022condensed} .
-Our initial results suggest that the proposed framework has the potential to accelerate the solution of AC OPF problems by an order of magnitude compared to existing tools, such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances, up to moderate tolerances ($10^{-6}$).
-We also discuss the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.
+The paper will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach.
+Our method is implemented in our packages, a nonlinear optimization solver MadNLP.jl and an automatic differentiation tool SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library.
+We compare our method against the standard CPU approach (W\"{a}chter and Beigler, 2006) and the recently developed reduced-space interior-point method on GPUs (Pacaud et.
+al., 2022) using the standard data available in pglib-opf.
+Preliminary results indicate that our proposed framework has the potential to accelerate the solution of AC OPF problems up to moderate tolerances ($10^{-6}$) by an order of magnitude compared to existing tools such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances.
+We will conclude with a discussion on the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.
 
-\newpage
-
-\bibliographystyle{plain}
-\bibliography{main}
+% \bibliographystyle{plain}
+% \bibliography{main}
 

### 2023-07-20 paper/main.tex
code directory added

--- 2023-06-30 abstract/main.tex
@@
 \maketitle
+\begin{abstract}
+This paper presents a computational framework for solving AC optimal
+power flow (OPF) problems using GPUs.  Despite the impressive
+capabilities of GPUs in various computing domains, their application
+in AC OPF has been limited by various challenges, including sparse
+matrix factorization and automatic differentiation.  To address these
+issues, we propose two strategies: (i) a condensed-space
+interior-point method (IPM) with inequality relaxation, and (ii) a
+single-instruction, multiple-data (SIMD) abstraction of nonlinear
+programs.  By relaxing equality constraints and condensing the
+Karush-Kuhn-Tucker system to the primal space, we enable efficient
+sparse matrix factorization on GPUs without numerical pivoting.
+Further, we introduce the SIMD abstraction for parallelized derivative
+evaluations on GPUs.  The paper provides numerical benchmark results,
+comparing our approaches with state-of-the-art tools such as JuMP,
+AMPL, and Ipopt, demonstrating an order of magnitude 
+speedup.  The paper concludes by discussing the future extensibility.
+\end{abstract}
 
-While graphics processing units (GPUs) have showcased impressive capabilities in various computing domains, their utilization in large-scale constrained nonlinear optimization regimes, such as alternating current (AC) optimal power flow (OPF) problems, has been somewhat limited.
-This limitation stems from the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms (Anitescu et.
-al.
-2021).
-Although GPU computation can accelerate various other parts of the optimization process, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations.
-To fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU that incorporates AD, linear algebra, and optimization while minimizing data transfers to and from host memory.
+\section{Introduction}
+While graphics processing units (GPUs) have showcased impressive
+capabilities in various computing domains, their utilization in
+large-scale constrained nonlinear optimization regimes, such as
+alternating current (AC) optimal power flow (OPF) problems, has been
+somewhat limited. This limitation stems from the challenges associated
+with parallel factorization of indefinite sparse matrices commonly
+encountered within constrained optimization algorithms
+\cite{anitescu2021targeting}. Although GPU computation can accelerate
+various other parts of the optimization process, including automatic
+differentiation (AD) and sparse matrix-vector multiplications, the
+slow data transfer between host and device memory hinders the ad-hoc
+implementation of GPU accelerations. To fully leverage the
+capabilities of modern GPU hardware, it is essential to implement a
+comprehensive computational framework on the GPU that incorporates AD,
+linear algebra, and optimization while minimizing data transfers to
+and from host memory.
 
-This paper presents a comprehensive computational framework and the associated software implementations for solving AC OPF problems on GPUs.
-Our approach utilizes the following techniques: (i) condensed-space interior-point methods (IPMs) with an inequality relaxation strategy, (ii) sparse matrix factorization with a fixed pivot sequence, and (iii) a single-instruction, multiple-data (SIMD) abstraction of nonlinear programs.
-Specifically, our method relaxes power flow equality constraints by allowing small violations, which enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space through the condensation procedure.
-Although this strategy is not new (see Nocedal and Wright, 2006), it has traditionally been considered less efficient than the standard full-space method due to increased fill-in in the sparse factorization.
-However, when implemented on GPUs, it offers the key advantage of guaranteeing positive definiteness in the condensed KKT system through the application of standard regularization techniques, which in turn, allows for the utilization of linear solvers with a fixed numerical pivot sequence (so-called refactorization).
-An efficient implementation of the sparse refactorization is available as part of the CUDA library, facilitating the implementation of efficient KKT system solutions on GPUs.
-Although this method is susceptible to ill-conditioning, our results demonstrate that the solver is robust enough to solve problems with a relative accuracy of $10^{-6}$.
+This paper presents a comprehensive computational framework and the
+associated software implementations for solving AC OPF problems on
+GPUs. Our approach utilizes the following techniques: (i)
+condensed-space interior-point methods (IPMs) with an inequality
+relaxation strategy, (ii) sparse matrix factorization with a fixed
+pivot sequence, and (iii) a single-instruction, multiple-data (SIMD)
+abstraction of nonlinear programs. Specifically, our method relaxes
+power flow equality constraints by allowing small violations, which
+enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the
+primal space through the condensation procedure. Although this
+strategy is not new (see Nocedal and Wright, 2006), it has
+traditionally been considered less efficient than the standard
+full-space method due to increased fill-in in the sparse
+factorization. However, when implemented on GPUs, it offers the key
+advantage of guaranteeing positive definiteness in the condensed KKT
+system through the application of standard regularization techniques,
+which in turn, allows for the utilization of linear solvers with a
+fixed numerical pivot sequence (so-called refactorization). An
+efficient implementation of the sparse refactorization is available as
+part of the CUDA library, facilitating the implementation of efficient
+KKT system solutions on GPUs. Although this method is susceptible to
+ill-conditioning, our results demonstrate that the solver is robust
+enough to solve problems with a relative accuracy of $10^{-6}$.
 
-Furthermore, by leveraging the SIMD abstraction of nonlinear programs, which preserves the parallelizable structure in the model, the model functions and derivative evaluations can be parallelized, thereby facilitating evaluations on the GPU.
-We demonstrate that the AC power flow model is particularly well-suited for this abstraction as it involves repetitive expressions for each type of component in the model (e.g., buses, lines, generators), and the number of computational patterns does not increase with the network's size.
-This structure has been effectively utilized by Gravity (Hijazi et.
-al.
-2018), demonstrating significant acceleration in model evaluations.
+Furthermore, by leveraging the SIMD abstraction of nonlinear programs,
+which preserves the parallelizable structure in the model, the model
+functions and derivative evaluations can be parallelized, thereby
+facilitating evaluations on the GPU. We demonstrate that the AC power
+flow model is particularly well-suited for this abstraction as it
+involves repetitive expressions for each type of component in the
+model (e.g., buses, lines, generators), and the number of
+computational patterns does not increase with the network's size. This
+structure has been effectively utilized by Gravity (Hijazi
+et. al. 2018), demonstrating significant acceleration in model
+evaluations.
 
-The paper will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach.
-Our method is implemented in our packages, a nonlinear optimization solver MadNLP.jl and an automatic differentiation tool SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library.
-We compare our method against the standard CPU approach (W\"{a}chter and Beigler, 2006) and the recently developed reduced-space interior-point method on GPUs (Pacaud et.
-al., 2022) using the standard data available in pglib-opf.
-Preliminary results indicate that our proposed framework has the potential to accelerate the solution of AC OPF problems up to moderate tolerances ($10^{-6}$) by an order of magnitude compared to existing tools such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances.
-We will conclude with a discussion on the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.
+We present comprehensive numerical benchmark results to demonstrate
+the efficiency of our proposed approach. Our method is implemented in
+our packages, a nonlinear optimization solver MadNLP.jl and an
+automatic differentiation tool SIMDiff.jl, with the solution of the
+KKT system being carried out using the external CUSOLVER library. We
+compare our method against the standard CPU approach (W\"{a}chter and
+Beigler, 2006) and the recently developed reduced-space interior-point
+method on GPUs (Pacaud et. al., 2022) using the standard data
+available in pglib-opf. Preliminary results indicate that our proposed
+framework has the potential to accelerate the solution of AC OPF
+problems up to moderate tolerances ($10^{-6}$) by an order of
+magnitude compared to existing tools such as Ipopt (interfaced with
+MATPOWER or PowerModels.jl), especially for large-scale instances. We
+will conclude with a discussion on the future extensibility of our
+method for more complex optimization tasks, such as multi-period,
+security-constrained, and joint transmission-distribution
+optimization.
 
-% \bibliographystyle{plain}
-% \bibliography{main}
+\paragraph*{Contributions}
+The contribution of this paper is three-fold.
+First, we present a condensed-space interior point method with inequality relaxation strategy is a general framework for solving sparse, large-scale nonlinar optimization problems on GPUs.
+Second, we present a SIMD abstraction of nonlinear program, which facilitates the implementation of parallel automatic differentiation on GPUs.
+Third, we present a comprehensive numerical benchmark result for the proposed two strategy using the standard test cases, thereby demonstrating the effectiveness of the two proposed methods.
+To the best of our knowledge, our approach is the fastest algorithm to solve AC OPF problems to $10^{-5}$ accuracy.
+
+\paragraph*{Organization} The paper is organized as follows.
+In the remainder of the current section, we introduce the mathematical notation.
+In Section \ref{sec:prelim}, we provide general preliminary knowledge on numerical optimization and GPU computing.
+In Section \ref{sec:simd}, we present the SIMD abstraction strategy for large-scale nonlinear programs and their advantages in terms of implementing parallel automatic differentiation.
+Section \ref{sec:ipm} presents the optimization algorithm under study, the condensed-space interior point method wiht inequality relaxation strategy.
+Section \ref{sec:num} presents the numerical results, comparing our approach with other state-of-the-art solution methods on GPUs.
+Finally, conclusions and future outlooks are given in Section \ref{sec:conc}.
+
+\paragraph*{Notation}
+We denote the set of real numbers and the set of integers by $\mathbb{R}$ and $\mathbb{I}$. We let $[M]:=\{1,2,\cdots,M\}$.
+
+\section{Preliminaries}\label{sec:prelim}
+This section introduces two key background information: GPU computing and numerical optimization for large-scale nonlinear optimzation problems.
+
+\subsection{GPU Computing}\label{sec:gpu}
+
+GPUs are particularly effective in performing the following operations:
+\begin{subequations}\label{eqn:pattern}
+  \begin{align}
+    y&\leftarrow \left[g(x; q_j)\right]_{j\in [J]}\\
+    y&\leftarrow y + \sum_{k\in [K]} h(x;s_k)\\
+    o&\leftarrow \sum_{i\in [I]} f(x; p_i),
+  \end{align}
+\end{subequations}
+
+\subsection{Numerical Optimization}\label{sec:numopt}
+The state-of-the-art solution strategies for AC OPF problems rely on numerical optimziation software, either general-purpose \cite{} or specialized \cite{}.
+Three main pillars of modern nonlinear optimization software include not only optimization solver, but also automatic differentiation tool (often provided by or interfaced with algebraic modeling language) and linear solvers.
+In fact, most of the expensive computational efforts are delegated to the external linear solver and automatic differentiation library.
+Optimization solver orchestrates the operation of these tools to drive the solution iterate to the stationary point of the optimization problem.
+
+\paragraph{Optimization Solvers}
+Optimization solvers
+
+\paragraph{Automatic Differentation}
+For the efficient solution of the optimization problems, it is crucial to be have the ability to
+
+\paragraph{Linear Solvers}
+
+\begin{figure}
+  \centering 
+  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
+    \node[draw,align=left] (I) at (0,2.75) {
+      $\bullet$ Instruction $g(\cdot,\cdot)$\\
+      $\bullet$ Pointer to array $x$
+    };
+    \node[draw] at (-4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (A\x) at (0,-\x/2) {$s_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (0,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (C\x) at (0,-\x/2) {$y_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \foreach\x in {1,2,3,4,5,7}{
+      \draw[->] (A\x.east) -- (B\x.west);
+      \draw[->] (B\x.east) -- (C\x.west);
+      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
+    }
+    \node at (-4,-2.5) {Input Data};
+    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
+    \node at (4,-2.5) {Output Data};
+  \end{tikzpicture}
+  \label{fig:simd}
+  \caption{A schematic description of SIMD parallelism}
+\end{figure}
+
+\section{SIMD Abstraction of Nonlinear Programs}\label{sec:simd}
+\begin{subequations}\label{eqn:prob}
+  \begin{align}
+    \min_{x^L\leq x \leq x^U}& \sum_{\ell\in[L]}\sum_{i\in [I_k]} f^{(\ell)}(x; p^{(\ell)}_i)\\
+    \st\; &\forall m\in[M]:\\
+    &\begin{aligned}[t]
+      g^{(m)}_\flat\leq &\left[g^{(m)}(x; q_j)\right]_{j\in [J]} \\
+      &+\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) \leq g^{(m)}_\sharp,
+    \end{aligned}
+  \end{align}
+\end{subequations}
+\section{Condensed-Space Interior Point Methods}\label{sec:ipm}
+
+\section{Numerical Results}\label{sec:num}
+
+\section{Conclusions and Future Outlook}\label{sec:conc}
+
+\bibliographystyle{IEEEtran}
+\bibliography{main}
 

### 2023-07-25 paper/main.tex
populated all sections

--- 2023-06-30 abstract/main.tex
@@
 % \input{preamble}
-\title{Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}
+\title{Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods 
+}
 
@@
 \maketitle
+\begin{abstract}
+  This paper introduces a novel computational framework for efficiently
+  solving AC optimal power flow (OPF) problems using GPUs. Although GPUs
+  have demonstrated remarkable performance in various computing domains,
+  their application in AC OPF has been limited due to challenges such as
+  serial computations for automatic differentiation (AD) and sparse matrix
+  factorization. To address these obstacles, we propose two strategies.
+  First, we utilize a single-instruction, multiple-data (SIMD)
+  abstraction of nonlinear programs (NLP). This approach allows us to
+  specify model equations while preserving their parallelizable
+  structure. By implementing AD on GPUs
+  efficiently, we can fully leverage the computational power of these
+  devices.  Second, we employ a Condensed-Space Interior-Point Method
+  (IPM) with inequality relaxation. This technique involves relaxing
+  equality constraints to inequalities and condensing the
+  Karush-Kuhn-Tucker system to the primal space. As a result, we enable
+  efficient sparse matrix factorization on GPUs without requiring a
+  numerical pivoting procedure.  By combining these strategies, we
+  achieve efficient AC OPF problem solutions on GPUs, keeping the
+  problem data residing in the device memory and performing the majority
+  of computations on the GPUs. Comprehensive numerical benchmark results
+  showcase the substantial computational advantage of our approaches for
+  solving AC OPF problems up to moderate precision on NVIDIA
+  GPUs. Remarkably, our method achieves an order of magnitude speedup
+  compared to state-of-the-art tools on CPUs, such as JuMP.jl, AMPL, and
+  Ipopt. The paper concludes by discussing the future extensibility of
+  our method.
+\end{abstract}
 
-While graphics processing units (GPUs) have showcased impressive capabilities in various computing domains, their utilization in large-scale constrained nonlinear optimization regimes, such as alternating current (AC) optimal power flow (OPF) problems, has been somewhat limited.
-This limitation stems from the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms (Anitescu et.
-al.
-2021).
-Although GPU computation can accelerate various other parts of the optimization process, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations.
-To fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU that incorporates AD, linear algebra, and optimization while minimizing data transfers to and from host memory.
+\section{Introduction}
 
-This paper presents a comprehensive computational framework and the associated software implementations for solving AC OPF problems on GPUs.
-Our approach utilizes the following techniques: (i) condensed-space interior-point methods (IPMs) with an inequality relaxation strategy, (ii) sparse matrix factorization with a fixed pivot sequence, and (iii) a single-instruction, multiple-data (SIMD) abstraction of nonlinear programs.
-Specifically, our method relaxes power flow equality constraints by allowing small violations, which enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space through the condensation procedure.
-Although this strategy is not new (see Nocedal and Wright, 2006), it has traditionally been considered less efficient than the standard full-space method due to increased fill-in in the sparse factorization.
-However, when implemented on GPUs, it offers the key advantage of guaranteeing positive definiteness in the condensed KKT system through the application of standard regularization techniques, which in turn, allows for the utilization of linear solvers with a fixed numerical pivot sequence (so-called refactorization).
-An efficient implementation of the sparse refactorization is available as part of the CUDA library, facilitating the implementation of efficient KKT system solutions on GPUs.
-Although this method is susceptible to ill-conditioning, our results demonstrate that the solver is robust enough to solve problems with a relative accuracy of $10^{-6}$.
+While graphics processing units (GPUs) have demonstrated remarkable
+capabilities in various computing domains, their adoption in
+large-scale constrained nonlinear optimization regimes, such as
+alternating current (AC) optimal power flow (OPF) problems, has faced
+some limitations. One of the primary challenges arises from the
+automatic differentiation (AD) of sparse model equations and the parallel
+factorization of indefinite sparse matrices commonly encountered
+within constrained optimization algorithms
+\cite{anitescu2021targeting}. While GPU computation can trivially
+accelerate several parts of the optimization process, especially
+various internal computations within the optimization solver, the
+sluggish data transfer between host and device memory hampers the
+ad-hoc implementation of GPU accelerations (Fig. \ref{fig:simd}). To
+fully leverage the potential that modern GPU hardware has to offer, it
+becomes imperative to have a comprehensive computational framework for
+optimization on GPUs so that AD, linear algebra
+operations, and optimization can be performed entirely on the GPU.
+Specifically, for best performance, both the problem data and the
+solver's intermediate computational data must be exclusively resident
+within the device memory, with the majority of operations executed on
+the GPU.
 
-Furthermore, by leveraging the SIMD abstraction of nonlinear programs, which preserves the parallelizable structure in the model, the model functions and derivative evaluations can be parallelized, thereby facilitating evaluations on the GPU.
-We demonstrate that the AC power flow model is particularly well-suited for this abstraction as it involves repetitive expressions for each type of component in the model (e.g., buses, lines, generators), and the number of computational patterns does not increase with the network's size.
-This structure has been effectively utilized by Gravity (Hijazi et.
-al.
-2018), demonstrating significant acceleration in model evaluations.
+\begin{figure}[t]
+  \centering
+  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
 
-The paper will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach.
-Our method is implemented in our packages, a nonlinear optimization solver MadNLP.jl and an automatic differentiation tool SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library.
-We compare our method against the standard CPU approach (W\"{a}chter and Beigler, 2006) and the recently developed reduced-space interior-point method on GPUs (Pacaud et.
-al., 2022) using the standard data available in pglib-opf.
-Preliminary results indicate that our proposed framework has the potential to accelerate the solution of AC OPF problems up to moderate tolerances ($10^{-6}$) by an order of magnitude compared to existing tools such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances.
-We will conclude with a discussion on the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.
+    \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
+    \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
+    \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};
 
-% \bibliographystyle{plain}
-% \bibliography{main}
+    \node[align=center] at (1.25, 3.75) {Data};
+    \node[align=center] at (6.75, 3.75) {Data};
+
+    % Host Memory
+    \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};
+
+    % Device Memory
+    \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};
+
+    % Arrow with Dashed Line
+    \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
+    \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
+    \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;
+
+    \def\rows{10}
+    \def\cols{10}
+    \def\elementwidth{2}
+    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
+
+      % Draw Grid
+      \foreach \i in {0,...,\rows} {
+        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
+      }
+      \foreach \i in {0,...,\cols} {
+        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
+      }
+    }
+  \end{tikzpicture}
+  \caption{A schematic description of host (CPU) and device (GPU) memory.}\label{fig:memory}
+\end{figure}
+
+This paper presents our approach to implement a comprehensive
+computational framework for solving large-scale 
+OPF problems on NVIDIA GPUs, along with the associated software
+implementations: SIMDiff.jl, an AD tool, and
+MadNLP.jl, a nonlinear optimizaiton solver. Our approach incorporates
+two novel strategies: (i) a single-instruction, multiple-data (SIMD)
+abstraction of nonlinear programs (NLPs), enabling streamlined
+parallel AD on GPUs, and (ii) a condensed-space
+interior-point method (IPM) with an inequality relaxation strategy,
+which facilitates the use of highly efficient {\it refactorization}
+routines for sparse matrix factorization with fixed pivot sequences.
+
+While derivative evaluation can be generally cheaper than linear
+algebra operations, our numerical results show that AD often
+constitutes more than half of the total solver time when using
+off-the-shelf AD implementations like JuMP.jl or AMPL. Instead of the
+general-purpose tools, our method leverages a specialized AD
+implementation based on the SIMD abstraction of NLPs. This abstraction
+alllows to preserve the parallelizable structure within the model
+equations, facilitating efficient derivative evaluations on the
+GPU. The AC power flow model is particularly well-suited for this
+abstraction as it involves repetitive expressions for each component
+type (e.g., buses, lines, generators), and the number of computational
+patterns does not increase with the network's size. Numerical results
+reported in this paper demonstrate that our proposed strategies can
+achieve over 20 times speedup. In comparison to general AD implementations
+on CPUs (such as AMPL and JuMP.jl), our GPU-based method is
+approximately 50 times faster in the largest instances.
+
+Linear algebra operations, especially sparse indefinite matrix
+factorization, are typically the bottleneck in nonlinear optimization
+solvers. Parallelizing this operation is challenging due to the need
+for numerical pivoting, which ensures the numerical
+stability. However, when the matrix can be factorized without
+numerical pivoting, a significant part of the operation can be
+parallelized, and the numerical factorization can be efficiently
+performed on GPUs. We develop a condensed-space IPM strategy that
+allows the use of sparse matrix factorization routines without
+numerical pivoting. This strategy relaxes equality constraints by
+permitting small violations, which enables expressing the
+Karush-Kuhn-Tucker (KKT) system entirely in the primal space
+through the condensation procedure. Although this strategy is not new
+\cite{nocedal2006numerical}, it has traditionally been considered less
+efficient than the standard full-space method due to increased nonzero
+entries in the KKT system. However, when implemented on GPUs, it
+offers the key advantage of ensuring positive definiteness in the
+condensed KKT system through the application of standard
+regularization techniques. This, in turn, allows for the utilization
+of linear solvers with a fixed numerical pivot sequence (known as
+refactorization). An efficient implementation of sparse
+refactorization is available as part of the CUDA library, enabling the
+implementation of efficient KKT system solutions on GPUs. Although
+this method is susceptible to numerical stability issues due to
+increased condition number in the KKT system, our results demonstrate
+that the solver is robust enough to solve problems with a relative
+accuracy of $\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$.
+
+We present numerical benchmark results to showcase the efficiency of
+our proposed approach, utilizing two of our packages: MadNLP.jl and
+SIMDiff.jl. The solution of the KKT system is performed using the
+external cuSOLVER library. To assess the performance of our method, we
+compare it against standard CPU approaches using the widely used data
+available in pglib-opf \cite{wachter2006implementation}.  Our
+benchmark results demonstrate that our proposed computational
+framework has significant potential for accelerating the solution of
+AC OPF problems, particularly for large-scale instances, especially
+when a moderate tolerance (e.g., $10^{-4}$) is
+considered. Specifically, when running on NVIDIA GPUs, our method
+achieves a notable 4x speedup compared to our solver running on CPU,
+in the case of the largest instance. Moreover for the same instance,
+our approach surpasses the performance of existing tools (such as
+Ipopt interfaced with JuMP.jl) by an order of magnitude.  This finding
+underscores the importance of harnessing the computational power of
+GPUs as a key enabler for tackling the challenges posed by large-scale
+OPF problems.
+
+\paragraph*{Organization}
+The paper is organized as follows. In the remainder of the current
+section, we introduce the mathematical notation. In Section
+\ref{sec:prelim}, we provide general preliminary knowledge on
+numerical optimization and GPU computing. In Section \ref{sec:simd},
+we present the SIMD abstraction strategy for large-scale nonlinear
+programs and their advantages in terms of implementing parallel
+AD. Section \ref{sec:ipm} presents the
+optimization algorithm under study, the condensed-space interior point
+method with inequality relaxation strategy. Section \ref{sec:num}
+presents the numerical results, comparing our approach with other
+state-of-the-art solution methods on GPUs. Finally, conclusions and
+future outlooks are given in Section \ref{sec:conc}.
+
+\paragraph*{Notation}
+We denote the set of real numbers and the set of integers by
+$\mathbb{R}$ and $\mathbb{I}$. We let $[M]:=\{1,2,\cdots,M\}$. We let
+$[v_i]_{i\in[M]}:=[v_1;v_2;\cdots,v_M]$.  A vector of ones with an
+appropriate size is denoted by $\boldsymbol{1}$. We use the following
+convention for any symbol $x$: $X:=\diag(x)$.
+
+\section{Preliminaries}\label{sec:prelim}
+This section covers three essential background topics: numerical
+optimization, AD, and GPU computing.
+
+\subsection{Numerical Optimization}\label{sec:numopt}
+We consider NLPs of the form:
+\begin{subequations}\label{eqn:cpt}
+  \begin{align}
+    \min_{x^\flat \leq x \leq x^\sharp}\;
+    & f(x)\\
+    \st\;
+    & g(x) =0. \label{eqn:cpt-con}
+  \end{align} 
+\end{subequations}
+Numerous solution algorithms have been developed in the NLP
+literature.  These methods can be broadly classified into active-set
+methods and interior-point methods \cite{nocedal2006numerical}.
+Active-set methods aim to find the set of active consraints associated
+with the optimal solution in a combinatorial manner, while
+interior-point methods replace inequality constraints with smooth
+barrier functions, enabling the use of Newton-type second-order
+algorithms. Interior-point methods are known to be more scalable for
+problems with a large number of constraints and suitable for
+parallelization, thanks to the fixed sparsity pattern of the KKT
+matrix. We adopt the interior-point approach and develop our
+optimizationmethods on GPUs.
+
+In terms of practical computations, three key components play
+vital roles: derivative evalutions (often provided by the AD
+capabilities of the algebraic modeling languages), linear algebra
+operations, and internal computations within the solver. Notably, most
+of the computational efforts are delegated to the external linear
+solver and AD library. The optimization solver
+orchestrates the operation of these tools to drive the solution
+iterate towards the stationary point of the optimization problem, but
+computations performed within the solver are generally much cheapr and
+simpler compared to the operations in AD and linear solver packages.
+
+Since the successful implementation of the open-source interior point
+solver in Ipopt, many subsequent implementations of nonlinear
+optimizaiton solvers
+\cite{chiang2014structured,rodriguez2023scalable,shin2021graph} have
+been based on Ipopt \cite{wachter2006implementation}. We also use
+Ipopt as our main reference for the IPM implementation. Below, we
+outline the overall computational procedure employed wihtin nonlinear
+optimization frameworks:
+
+1) Given the current primal-dual iterate $(x^{(\ell)},y^{(\ell)},
+z^{\flat(\ell)},z^{\sharp(\ell)})$, the AD package computes:
+$\nabla_x f(x^{(\ell)})$, $\nabla_x g(x^{(\ell)})$, $\nabla^2_{xx}
+\mathcal{L}(x^{(\ell)},y^{(\ell)},z^{\flat(\ell)},z^{\sharp(\ell)})$, where
+$\mathcal{L}(x^{\flat},y^{\flat},z^{\flat},z^{\sharp}):=f(x) - y^\top
+g(x) - z^\flat (x-x^\flat) - z^\sharp (x^\sharp-x)$.
+
+2) The following sparse indefinite system (known as the KKT system) is
+solved using sparse indefinite factorization (typically, via sparse
+LBL$^\top$ factorization) and triangular solve routines.
+\begin{align}\label{eqn:kkt-indefinite}
+  &\begin{bmatrix}
+    W^{(\ell)}  + \Sigma^{(\ell)} + \delta^{(\ell)}_w I& A^{(\ell)\top}\\
+    A^{(\ell)} & -\delta_c I\\
+  \end{bmatrix}
+  \begin{bmatrix}
+    \Delta x\\
+    \Delta y\\
+  \end{bmatrix}=
+  \begin{bmatrix}
+    f - \mu X^{-1} \boldsymbol{1}\\
+    \Delta y\\
+  \end{bmatrix}
+\end{align}
+Here, $W^{(\ell)}:=\nabla^{2}_{xx}\mathcal{L}(x^{(\ell)},y^{(\ell)})$,
+$A^{(\ell)}:= \nabla_xg(x^{(\ell)})$, and $\delta_w, \delta_c>0$ are
+the regularization parameters determined based on the inertia
+correction procedure.
+
+3) The optimization solver employs the filter line search procedure to
+determine the step size. While alternative approaches, such as
+trust-region-type methods and merit function-based criteria, can be
+used for determining the step size, the filter line search method is
+most commonly used due to its implementation in Ipopt.
+
+\subsection{AD}
+Efficiently evaluating the derivatives of $f(\cdot)$, $g(\cdot)$, and
+$\mathcal{L}(\cdot,\cdot)$ is crucial for the efficient solution of
+optimization problems.
+
+Numerical differentiation of computer programs can be achieved through
+three different methods: (i) finite difference method, (ii) symbolic
+differentiation, and (iii) AD. The finite difference method suffers
+from numerical rounding errors, and its computational complexity grows
+unfavorably with respect to the number of function arguments, making
+it less preferable unless no other alternatives are
+available. Symbolic differentiation uses computer algebra systems to
+obtain symbolic expressions of first or higher-order
+derivatives. While this method can differentiate functions up to high
+numerical precision, it suffers from "expression swelling" effect and
+struggles to compute the derivatives of long nested expressions in a
+computationally efficient way.
+
+In contrast, AD differentiates computer
+programs by leveraging the computation graph. By inspecting the
+computation graph and applying chain rules, AD
+can efficiently and accurately evaluate derivatives. This approach has
+become the dominant paradigm for derivative computation within the
+scientific computing domain, including nonlinear optimization and
+machine learning. For large-scale optimization, such as AC OPFs, AD
+tools are often implemented as part of domain-specific modeling
+languages, such as JuMP, CasADi, and AMPL (optimization) and Torch and
+Flux (machine learning).
+
+There are two alternative ways of propagating derivatives through the
+recursive application of chain rules: (i) forward-mode and (ii)
+reverse-mode, which operate in opposite directions (from leaves to
+root and from root to leaves, respectively). Reverse-mode automatic
+differentiation, also known as an adjoint method, has proven to be
+particularly effective for dealing with the function expressions in
+large-scale optimization problems.
+
+The Julia Language, our language of choice, provides convenient ways
+of implementing highly efficient automatic
+differentiation. \textit{Any Julia function}, including commonly used
+operations such as addition, multiplication, trigonometric functions,
+exponential functions, and others, can be straightforwardly extended
+through the use of multiple dispatch paradigm, allowing functions to
+be dynamically dispatched based on the run-time type. Several
+implementations of AD are available in Julia, including
+ReverseDiff.jl, ForwardDiff.jl, Zygote.jl, and JuMP.jl. While these
+tools are general and useful for various applications, they are not
+optimized for evaluating derivatives of AC OPF problems, which contains
+an obvious parallelizable structure that can be exploited.
+
+\subsection{GPU Computing}\label{sec:gpu}
+With the increasing prevalence of GPU in various scientific computing
+tasks, there has been growing interest in leveraging these parallel
+systems to efficiently solve large-scale OPF problems. Some recent
+concurrent approaches include HyKKT, a hybrid (direct-iterative) KKT
+system solver, \cite{regev2023hykkt} and HiOP, a HPC solver for
+nonlinear optimization problems, \cite{hiop_techrep}. However,
+adapting the IPM to GPUs presents challenges due to the fundamental
+differences between GPU and CPU programming paradigms. While CPUs
+execute a sequence of instructions on a single input (single
+instruction, single data, or SISD, in Flynn's taxonomy), GPUs run the
+same instruction simultaneously on hundreds of threads using the SIMD
+paradigm (see Fig. \ref{fig:simd}). The SIMD parallelism works well
+for algorithms that can be decomposed into simple instructions running
+entirely in parallel, but not all algorithms seamlessly fit this
+paradigm. For example, branching in the control flow can hinder
+lockstep execution across multiple threads, and in turn, prevents
+efficient implementations on GPUs. On the other hand, when the
+algorithm's structure allows for efficient parallelization, the SIMD
+parallelism in GPUs becomes highly effective, and oftentimes enables
+orders of magnitudes speedup.
+
+We highlight that the following, arguably quite common, computational
+patterns are particularly effective when implemented on GPUs:
+\begin{align}
+  y&\leftarrow \left[g(x; q_j)\right]_{j\in [J]}\tag{Pattern 1}\label{eqn:pattern-1}\\
+  y&\leftarrow y + \sum_{k\in [K]} h(x;s_k)\tag{Pattern 2}\label{eqn:pattern-2}\\
+  o&\leftarrow \sum_{i\in [I]} f(x; p_i)\tag{Pattern 3}\label{eqn:pattern-3}.
+\end{align}
+Here, it is assumed that $f:\mathbb{R}^{n_x}\times
+\mathbb{R}^{n_{p}}\rightarrow \mathbb{R}$, $g:\mathbb{R}^{n_x}\times
+\mathbb{R}^{n_{s}}\rightarrow \mathbb{R}$, and
+$h:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{p}}\rightarrow
+\mathbb{R}^{n_y}$ are simple instructions that require only a handful
+of operations. \ref{eqn:pattern-1} is typically most effective on
+GPUs, where each available thread can operate independently without
+needing to simultaneously manipulate the same data reference in device
+memory. \ref{eqn:pattern-2} and \ref{eqn:pattern-3} are more
+challenging to implement, as they require simultaneous access to the
+same device memory location, but they can be parallelized effectively
+by using buffers.
+
+Many of the operations required in AD of sparse
+physical models, as well as the application of optimization
+algorithms, are based on the computational patterns mentioned
+above. For example, an OPF model can be implemented
+with 14 different computational patterns, all of which fall within the
+aforementioned categories. Additionally, the computation within
+optimization solvers, such as forming the left-hand-side for the KKT
+systems, computing the $\|\cdot\|_\infty$ norm of the constraint
+violation, and assembling the condensed KKT system, can be carried out
+by using these computational patterns. The only exception is the
+factorization of the sparse KKT matrix, which requires more sophisticated
+implementations.
+
+Implementing kernel functions for the above computational patterns in
+the Julia Language is straightforward, as it provides excellent
+high-level interfaces for array and kernel programming for GPU
+arrays. The code can even be device-agnostic, thanks to the portable
+programming capability brought by KernelAbstractions.jl. All
+of the AD and optimization capabilities in our
+tools, MadNLP.jl and SIMDiff.jl, are implemented in Julia Language,
+utilizing its kernel and array programming capabilities.
+
+\begin{figure}[t]
+  \centering 
+  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
+    \node[draw,align=left] (I) at (0,2.75) {
+      $\bullet$ Instruction $g(\cdot,\cdot)$\\
+      $\bullet$ Pointer to array $x$
+    };
+    \node[draw] at (-4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (A\x) at (0,-\x/2) {$s_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (0,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (C\x) at (0,-\x/2) {$y_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \foreach\x in {1,2,3,4,5,7}{
+      \draw[->] (A\x.east) -- (B\x.west);
+      \draw[->] (B\x.east) -- (C\x.west);
+      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
+    }
+    \node at (-4,-2.5) {Input Data};
+    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
+    \node at (4,-2.5) {Output Data};
+  \end{tikzpicture}
+  \label{fig:simd}
+  \caption{A schematic description of SIMD parallelism in GPU computing}
+\end{figure}
+
+\section{SIMD Abstraction of NLPs}\label{sec:simd}
+
+This section describes our implementation of SIMD abstraction and
+sparse AD of the model equations. The abstraction and AD are
+implemented as part of our algebriac modeling language SIMDiff.jl.
+
+The SIMD abstraction under consideration is as follows:
+\begin{subequations}\label{eqn:prob}
+  \begin{align}
+    \min_{x^L\leq x \leq x^U}& \sum_{\ell\in[L]}\sum_{i\in [I_k]} f^{(\ell)}(x; p^{(\ell)}_i)\\
+    \st\; &\forall m\in[M]:\\
+                             &\begin{aligned}[t]
+                               g^{(m)}_\flat\leq &\left[g^{(m)}(x; q_j)\right]_{j\in [J]} \\
+                               &+\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) \leq g^{(m)}_\sharp,
+                             \end{aligned}
+  \end{align}
+\end{subequations}
+where $f^{(\ell)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
+$h^{(n)}(\cdot,\cdot)$ are twice differentiable functions with respect
+to the first argument, ${{p^{(k)}i}{i\in [N_k]}}{k\in[K]}$,
+${{q^{(k)}{i}}{i\in [M\ell]}}{\ell\in[L]}$, and
+${{s^{(k)}{j}}{j\in[J\ell]}}_{\ell\in[L]}$ are problem data, which can
+either be discrete or continuous. We assume that our functions
+$f^{(\ell)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
+$h^{(n)}(\cdot,\cdot)$ can be expressed in the form of computational
+graphs of moderate length; that is, the resulting Lagrangian Hessian
+and Jacobian matrices are sparse. One can observe that the problem in
+\eqref{eqn:prob} is expressed by a combination of the computational
+patterns in Section \ref{sec:gpu}.
+
+Our implementation of the algebraic modeling interface enforces the
+user to specify the model equations in an {\tt Iterable} data type in
+Julia. Julia's {\tt Iterable} data type consists of the instruction
+and the data over which the instruction is executed. This allows
+maintaining the NLP problem information in the form of SIMD
+abstraction in \eqref{eqn:prob}.
+
+Now, we discuss the key benefits of the SIMD abstraction.
+
+\paragraph{Parallel AD Implementation}
+The main reason for introducing a distinctive modeling abstraction in
+\eqref{eqn:prob} is because without the abstraction, it is difficult
+for the AD package to detect the parallelizable structure within the
+model. Many of the physics-based models, such as AC OPF, have
+repetitive structure. One of the manifestations of it is
+that the mathematical statement of the model is concise, even if the
+model itself contains millions or even billions of variables and
+constraints. Specifically, it suffices to use 15 computational
+patterns to fully specify the model. These patterns arise from (1)
+generation cost, (2) reference bus voltage angle constraint, (3-6)
+active and reactive power flow (to and from), (7) voltage angle
+difference constraint, (8-9) apparent power flow limits (to and from),
+(10-11) power balance equations, (12-13) generators' contributions to
+the power balance equations, and (14-15) in/out flows contributions to
+the power balance equations. However, such repetitive structure is not
+well exploited during AD in the standard
+nonlinear optimization modelign tools. By preserving the repetitive
+structures in the model, one can facilitate the implementation of
+parallelized derivative computations.
+
+SIMDiff.jl implements the standard reverse-mode automatic
+differentiation. Using the multiple dispatch feature of Julia,
+SIMDiff.jl can generate highly efficient derivative computation code,
+specifically compiled for each computational pattern in the model. Furthermore,
+the derivative evaluation code can straightforwardly extended so that
+it can run over the data on acceraltor devices. In particular,
+the derivative computation code is obtained as a type-stable function,
+which can be run on devices through array and kernel programming.
+
+\paragraph{Memory Efficiency}
+The memory footprint of storing model
+information can be significantly reduced with the SIMD abstraction. As
+opposed to storing the symbolic expressions and sparsity patterns for
+each constraint and objective term, which can potentially be more than
+millions, we only store the symbolic expression and sparsity patterns
+for each computational pattern. This allows drastically reducing the
+memory requirements. Thanks to this, our implementation has a
+significantly shorter model creation time and smaller memory footprint
+compared to the existing sparse AD tools.
+
+\paragraph{Efficient Symbolic Analysis}
+The symbolic analysis is essential in sparse automatic
+differentiation. This step is needed for determining the sparsity
+pattern of the evaluated derivatives. In sparse AD, the initial
+symbolic analysis of nonlinear expressions can be expensive, as the
+number of objective terms and constraints can be more than
+millions. However, often times those analyses are applied for the same
+computational patterns, and the time and memory spent for symbolic
+analysis can be significantly reduced if the SIMD-compatible
+computational patterns are present in the model. SIMDiff.jl exploits
+the SIMD abstraction of the model equations to save the computational
+cost spent for symbolic analysis, and this allows for efficient
+evaluation of derivatives without the use of the expensive symbolic
+anlaysis nor coloring procedures.
+
+\section{Condensed-Space Interior Point Methods with Inequality Relaxation Strategy}\label{sec:ipm}
+We present the condensed-space interior point method within the
+context of the general NLP formulation in \eqref{eqn:cpt}. Our method
+has two key differences from standard interior point method
+implementations: (i) the use of inequality relaxation and (ii) the
+condensed treatment of the KKT system.
+
+\subsection{Inequality Relaxation}
+
+At the beginning of the algorithm, we apply inequality relaxation to replace the equality constraints in \eqref{eqn:cpt} with inequalities by introducing slack variables $s\in\mathbb{R}^{m}$:
+\begin{align}\label{eqn:relax}
+  g(x)- s = 0,\quad s^{\flat}\leq s\leq  s^\sharp,
+\end{align}
+where $s^\flat,s^\sharp\in\mathbb{R}^{m}$ are chosen to be close to zero.
+This relaxed problem can be stated as follows:
+\begin{align}\label{eqn:relaxed} 
+  \begin{aligned}[t]
+    \min_{\big[\substack{x^\flat\\s^\flat}\big]\leq \big[\substack{x\\s}\big] \leq \big[\substack{x^\sharp\\s^\sharp}\big]}\;
+    &  f(x)\\
+    \st\;
+    & g(x) - s = 0.
+  \end{aligned}
+\end{align}
+
+In our implementation, we set $s^{\flat},s^\sharp$ as
+$-\epsilon_{\text{tol}}$ and $+\epsilon_{\text{tol}}$, respectively,
+where $\epsilon_{\text{tol}}>0$ is a user-specified relative tolerance
+of the interior point method (IPM). This type of relaxation is
+commonly used in practical IPM implementations; for example, in Ipopt,
+the solver relaxes the bounds and inequality constraints by
+$O(\epsilon_{\text{tol}})$ to prevent an empty interior of the
+feasible set (see \cite[Section 3.5]{wachter2006implementation}).
+For condensed-space IPM, we cannot maintain the same level of precision
+due to the increased condition number of the KKT system.
+We have found that setting $\epsilon_{\text{tol}}$ to be
+$\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$ ensures numerical
+stability while achieving satisfactory convergence. Thus, our solver
+sets the tolerance to $10^{-4}$ by default when using condensed IPM.
+
+\subsection{Barrier Subproblem}
+
+The barrier subproblem is considered as follows:
+\begin{subequations}\label{eqn:barrier}
+  \begin{align}
+    \min_{x}\;
+    &
+      \begin{aligned}[t]
+        &f(x) - \mu\boldsymbol{1}^\top \log (x- x^\flat) -\mu\boldsymbol{1}^\top \log (x^\sharp -x)\\
+        &- \mu\boldsymbol{1}^\top \log (s- s^\flat) -\mu\boldsymbol{1}^\top \log (s^\sharp -s)
+      \end{aligned}
+          \label{eqn:barrier-obj}\\
+    \st\;
+    &g(x) = 0. \label{eqn:barrier-con}
+  \end{align}
+\end{subequations}
+Here, $\mu>0$ is the barrier parameter. The smooth log-barrier
+function is employed to avoid handling inequalities in a combinatorial
+fashion (as in active set methods). A superlinear local convergence to
+the first-order stationary point can be achieved by repeatedly
+applying Newton's step to the KKT conditions of \eqref{eqn:barrier}
+with $\mu=\mu^{(\ell)}$ and a decreasing sequence of
+${\mu^{(\ell)}}_{\ell=0,1,\cdots}$. A global convergence can be
+achieved by employing a certain globalization strategy (either based
+on a merit function or filter method) \cite{nocedal2006numerical}.
+
+\subsection{Newton's Step Computation}
+The Newton step direction can be computed by considering the first-order optimality conditions for the barrier subproblem in \eqref{eqn:barrier}:
+\begin{align}\label{eqn:first} 
+  \begin{aligned}[t]
+    \nabla_{x} f(x^{(\ell)}) - \nabla_{x}g(x^{(\ell)})^\top y^{(\ell)}  - z_x^\flat  + z_x^\sharp = 0\;&\\
+    \begin{aligned}
+      - z_s^\flat  + z_s^\sharp  &= 0\\
+      Z^\flat_x (x^{(\ell)}-x^\flat) - \mu\boldsymbol{1} &= 0\\
+      Z^\flat_s (s^{(\ell)}-s^\flat) - \mu\boldsymbol{1}&= 0,
+    \end{aligned}
+    \quad
+    \begin{aligned}
+      g(x^{(\ell)}) &= 0\\
+      Z^\sharp_x (x^\sharp-x^{(\ell)}) - \mu\boldsymbol{1}&= 0\\
+      Z^\sharp_s (s^\sharp-s^{(\ell)}) - \mu\boldsymbol{1}&= 0,
+    \end{aligned}&
+  \end{aligned}
+\end{align}
+where $y\in\mathbb{R}^{m}$, $z_x^\flat,z_x^\sharp\in\mathbb{R}^{n}$,
+and $z_s^\flat,z_s^\sharp\in\mathbb{R}^{m}$ are Lagrange multipliers
+associated with the equality and bound constraints in
+\eqref{eqn:relaxed}. The Newton step for solving the nonlinear
+equation in \eqref{eqn:first} can be computed by solving the following
+KKT system in \eqref{eqn:very-long-eqn}.
+\begin{figure*}[h!]
+  % \hline
+  \begin{align}\label{eqn:very-long-eqn}
+    \begin{bmatrix}
+      W^{(\ell)}  + \delta^{(\ell)}_w I & & A^{(\ell)\top}& -I & I &  \\
+      & \delta^{(\ell)}_w I & -I&&&-I & I\\
+      A^{(\ell)}& -I & -\delta_c I\\
+      Z_x^{(\ell)\flat}&&&X^{(\ell)}-X^\flat\\
+      -Z_x^{(\ell)\sharp}&&&&X^\sharp-X^{(\ell)}\\
+      &Z_s^{(\ell)\flat}&&&&S^{(\ell)}-S^\flat\\
+      &-Z_s^{(\ell)\sharp}&&&&&S^\sharp-S^{(\ell)}\\
+    \end{bmatrix}
+    \begin{bmatrix}
+      \Delta x \\
+      \Delta s \\
+      \Delta y \\
+      \Delta z_x^\flat \\
+      \Delta z_x^\sharp \\
+      \Delta z_s^\flat \\
+      \Delta z_s^\sharp \\
+    \end{bmatrix} =
+    \begin{bmatrix}
+      p_{x }\\
+      p_{s }\\
+      p_{y }\\
+      p_{z_x^\flat }\\
+      p_{z_x^\sharp }\\
+      p_{z_s^\flat }\\
+      p_{z_s^\sharp }\\
+    \end{bmatrix}
+  \end{align}
+  % \hline
+\end{figure*}
+Here, $p_x,\cdots p_{z_s^\sharp}$ are defined by the left-hand-sides of the equations in
+\eqref{eqn:first}. In what follows, we shall drop the superscript
+$(\cdot)^{(\ell)}$ for concise notation.
+
+Now, we observe that a significant portion of the system in
+\eqref{eqn:very-long-eqn} can be eliminated by exploiting the block
+structure, leading to an equivalent system stated in a smaller
+space. In particular, the lower-right $4\times 4$ block is always
+invertible since the IPM procedure ensures that the iterates stay in
+the strict interior of the feasible set. This allows for eliminating
+the lower-right 4x4 block, resulting in:
+\begin{align}\label{eqn:very-long-reduced}
+  &\begin{bmatrix}
+    W^{(\ell)}  + \Sigma_x + \delta^{(\ell)}_w I & & A^{(\ell)\top} \\
+    & \Sigma_s + \delta^{(\ell)}_w I & -I\\
+    A^{(\ell)}& -I & -\delta_c I\\
+  \end{bmatrix}
+  \begin{bmatrix}
+    \Delta x \\
+    \Delta s \\
+    \Delta y \\
+  \end{bmatrix}=
+  \begin{bmatrix}
+    q_x \\
+    q_s\\
+    q_y\\
+  \end{bmatrix},
+\end{align}
+where the dependence on the evaluation point is suppressed for concise notation, and:
+\begin{align*}
+  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
+  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
+  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
+  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
+  q_y&:=p_y.
+\end{align*}
+The bound dual steps can be recovered as follows:
+\begin{align}\label{eqn:recover-1}
+  \begin{aligned}[t]
+    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
+    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
+    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
+    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
+  \end{aligned}
+\end{align}
+Note that the reduced system in \eqref{eqn:very-long-reduced}
+corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
+in the original version of the algorithm, we did not introduce the
+slack variables, so it does not have the additional structure imposed
+by the slack variables.
+
+The key advantage of the inequality relaxation strategy is that it
+imposes additional structure in the reduced KKT system, allowing us to
+further reduce the dimension of the problem. In particular, the
+lower-right 2x2 block in \eqref{eqn:very-long-reduced} can be
+eliminated. This procedure is called {\it condensation}. Through this,
+we obtain the following system, purely written in the original primal
+space:
+\begin{align}\label{eqn:very-long-condensed}
+  (W &+ \delta_wI + \Sigma_x + A D A^{\top} ) \Delta x  =\\\nonumber
+     &  q_x + A^\top \left[q_s + \left(\delta_c \Sigma_s + (1+\delta_c\delta_w) I\right)^{-1} (q_y -\delta_c q_y )\right],
+\end{align}
+where:
+\begin{align*}
+  D := \left(\delta_c \Sigma_s + (1+\delta^{}_c\delta^{}_w) I\right)^{-1} \left(\Sigma_s + \delta^{}_w I\right).
+\end{align*}
+The dual and slack step directions can be recovered as follows:
+\begin{align}
+  \Delta s &:= \left(\delta_c \Sigma_s + (1+\delta^{}_c\delta^{}_w) I\right)^{-1} \left(\delta_c q_s - (q_y + A^{(\ell)}\Delta x)\right)\nonumber\\
+  \Delta y &:= (\Sigma_s + \delta_w I) \Delta s -q_s.\label{eqn:recover-2}
+\end{align}
+
+Therefore, the only sparse matrix that needs to be factorized is the
+matrix in \eqref{eqn:very-long-condensed}, which is in
+$\mathbb{R}^{n\times n}$. In general NLPs, the condensation strategy
+can arbitrarily increase the density of the KKT system, but for AC OPF
+problems, the condensed KKT system is still sparse.
+This is because the maximum number of nonzeros per row in constraint
+Jacobian is bounded by the graph degree, which should be reasonably
+small in practice.
+
+The reason that the condensation strategy is particularly relevant for
+GPUs is that the matrix in \eqref{eqn:very-long-condensed} is positive
+definite upon the application of standard inertia correction
+method. Typically, to guarantee that the computed step direction is a
+descent direction, we need a condition that
+$\text{inertia}(M_\text{reduced}) = (n+5m,0,m)$. One can observe that
+by applying Sylvester's law of inertia, we have:
+\begin{align*}
+  &\text{inertia}(M_\text{reduced}) = (n+5m,0,m)\\
+  &\iff \text{inertia}(M_\text{cond}) = (n,0,0).
+\end{align*}
+Thus, any choice of $\delta_w,\delta_c>0$ that makes the condensed KKT
+system positive definite yields the KKT system with the desired
+inertia information. This implies that the condensed KKT matrix
+$M_{\text{cond}}$ can be factorized with fixed pivoting (e.g.,
+Cholesky factorization or LU factorization), which is significantly
+more amenable to parallel implementation than indefinite LBL$^\top$
+factorization, which is commonly used in IPM on CPUs.
+
+Typically, the system in \eqref{eqn:very-long-condensed} is
+severely ill-conditioned, and a single triangular solve may not
+provide a sufficiently accurate step direction. Accordingly, iterative
+refinement methods are employed to refine the solution by performing
+multiple triangular solves. Notably, iterative refinement is applied
+to the full KKT system \eqref{eqn:very-long-eqn} by using the dual step
+recovering procedure \eqref{eqn:recover-1} and \eqref{eqn:recover-2}.
+
+\subsection{Line Search and IPM Iterations}
+
+The step size can be determined using the line search
+procedure. Although there are numerous alternative approaches, we
+follow the filter line search method implemented in the Ipopt solver
+\cite{wachter2006implementation}. The step can be implemented as follows:
+\begin{align}
+  (x,s,y) &\leftarrow (x,s,y)+ \alpha (\Delta x, \Delta s, \Delta y),\label{eqn:iter}\\\nonumber
+  (z_x^\flat, z_x^\sharp, z_s^\flat, z_s^\sharp) &\leftarrow (z_x^\flat, z_x^\sharp, z_s^\flat, z_s^\sharp) + \alpha_z (\Delta z_x^\flat, \Delta z_x^\sharp, \Delta z_s^\flat, \Delta z_s^\sharp).
+\end{align}
+
+The iteration in \eqref{eqn:iter} is repeated until the convergence
+criterion is satisfied. The convergence criterion is defined as
+$\text{residual}(x^{(\ell)}, s^{(\ell)}, y^{(\ell)},
+z^{(\ell)\flat}_x, z^{(\ell)\sharp}_x, z^{(\ell)\flat}_s,
+z^{(\ell)\sharp}s) <\epsilon_{\text{tol}}$, where
+$\text{residual}(\cdot)$ is a scaled version of the residual to the
+first-order conditions in \eqref{eqn:first}.
+
+Furthermore, to enhance the convergence behavior, various additional
+strategies are implemented, such as the second-order correction,
+restoration phase, and automatic scaling. For the most part, we follow
+the reference for the implementation of these strategies
+\cite{wachter2006implementation}.
+
+Finally, we summarize our condensed-space interior point method in
+Algorithm \ref{alg:con-ipm}.
+
+\begin{algorithm}[t]
+  \caption{Condensed-Space Interior Point Method}
+  \label{alg:con-ipm}
+  \begin{algorithmic}[1]
+    \REQUIRE Primal-dual solution guesses $x,y, z^\flat, z^\sharp$, bounds $x^\flat,x^\sharp, s^\flat, s^\sharp$, callbacks $f(\cdot)$, $g(\cdot)$, $\nabla_x f(\cdot)$, $\nabla_x g(\cdot)$, $\nabla^2_{xx} \mathcal{L}(\cdot)$, and tolerance $\epsilon_{\text{tol}}$
+    \STATE Relax the equality constraints by \eqref{eqn:relax} and initialize the slack $s$ and the associated dual variables $z^\flat_s, z^\sharp_s$.
+    \WHILE{$\text{residual}(x,y,z_x^\flat,z_x^\sharp,z_s^\flat,z_s^\sharp)> \epsilon_{\text{tol}}$}
+    \STATE Solve the condensed KKT system \eqref{eqn:very-long-condensed} to compute the primal step $\Delta x$ and recover the dual steps $\Delta y, \Delta z_x^\flat, \Delta z_x^\sharp, \Delta z_s^\flat, \Delta z_s^\sharp$ by \eqref{eqn:recover-1} and \eqref{eqn:recover-2}.
+    \STATE Determine the need for regularization via the inertia-free regularization procedure.
+    \STATE Choose a step size $\alpha>0$ satisfying sufficient progress conditions and acceptable by filter via line search.
+    \STATE Compute the feasible dual step size $\alpha_z>0$.
+    \STATE Update the solution by \eqref{eqn:iter}.
+    \STATE Update filter and barrier parameter $\mu$.
+    \ENDWHILE
+    \RETURN The first-order stationary points $x^\star,y^\star, z^{\flat\star}, z^{\sharp\star}$
+  \end{algorithmic}
+\end{algorithm}
+
+\subsection{Notes on the Implementation}
+
+We have implemented the condensed-space IPM by adapting our code base
+in MadNLP.jl. In MadNLP, the IPM iteration is implemented with high
+levels of abstractions, while the specific handling of the data
+structures within the KKT systems is carried out by data-type specific
+kernel functions. This design allows us to run the same high-level
+mathematical abstractions for different data structures, such as {\tt
+  SparseKKTSystem}, {\tt DenseKKTSystem}, {\tt DenseCondensedKKTSystem},
+etc. For the implementation of the condensed-space IPM presented in
+this paper, we have added a new type of KKT system called {\tt
+  SparseCondensedKKTSystem} and implemented additional kernels needed
+for handling the data structures specific to this KKT system
+type. This approach ensures that we are performing mathematically
+equivalent operations as in the mature, extensively tested existing
+code base. This also allows us to easily switch between different KKT
+system types, which is crucial for experimenting with various solvers
+and data structures, as well as for efficiently leveraging GPU
+acceleration when available. Furthermore, by maintaining this level of
+abstraction, the condensed-space IPM can be seamlessly integrated into
+the existing framework, making it easier to maintain and extend in the
+future.
+
+\section{Numerical Results}\label{sec:num}
+
+This section presents the numerical benchmark results, comparing our
+method against state-of-the-art methods on CPUs for solving standard
+AC OPF problems.
+
+\subsection{Methods}
+
+We compared four different configurations:
+\begin{align}
+  \label{config-1}\tag{Config 1} \bullet\;&\text{MadNLP.jl + SIMDiff.jl + cuSOLVER (GPU)}\\
+  \label{config-2}\tag{Config 2} \bullet\;&\text{MadNLP.jl + SIMDiff.jl + Ma27 (CPU)}\\
+  \label{config-3}\tag{Config 3} \bullet\;&\text{Ipopt + AMPL + Ma27 (CPU)}\\
+  \label{config-4}\tag{Config 4} \bullet\;&\text{Ipopt + JuMP.jl + Ma27 (CPU)}.
+\end{align}
+
+\ref{config-1} is our main GPU configuration, and \ref{config-2}
+represents our implementation running on CPU. \ref{config-3} and
+\ref{config-4} are used as benchmarks. \ref{config-1} and
+\ref{config-2} share a common code base, thanks to the multiple
+dispatch feature of Julia Language, but they differ in how they handle
+the KKT systems. In \ref{config-1}, MadNLP.jl applies the
+condensed-space interior point method along with the inequality
+relaxation strategy, while in \ref{config-2}, MadNLP.jl applies
+factorization to the non-condensed, indefinite KKT system (as in
+\eqref{eqn:kkt-indefinite}). In \ref{config-1}, we use the cuSOLVER
+library to solve the condensed KKT system. The initial symbolic
+factorization is performed using the KLU package
+\cite{davis2004algorithm}. Subsequent numerical factorization and
+triangular solves are performed by cuSOLVER with the fixed pivot
+sequence obtained from the initial factorization with KLU. Software
+and hardware details of each configuration are illustrated in Table
+\ref{tbl:settings}. The OPF problem is formulated using
+the model from the rosetta-opf project \cite{rosetta-opf}, and the
+test cases are obtained from the pglib-opf repository
+\cite{babaeinejadsarookolaee2019power}. The external packages are
+called from Julia, through thin wrapper packages, such as Ipopt.jl and
+AmplNLWriters.jl. A tolerance of $10^{-4}$ is set for MadNLP.jl and
+Ipopt solvers, with other solver options adjusted to ensure a fair
+comparison across different solvers. The results can be reproduced
+with the script available at
+\url{https://github.com/sshin23/pscc-2024-opf-gpu}.
+
+\subsection{Results}
+
+The numerical benchmark results, including total solution time and its
+breakdown into linear algebra and derivative evaluation time (with the
+remainder considered as solver internal time), are shown in Table
+\ref{tbl:results}. The quality of the solution (objective value and
+constraint violation measured by $\|\cdot\|_\infty$) is shown in Table
+\ref{tbl:quality}. Figure \ref{fig:speedup} visually represents the
+speedup brought by GPUs, comparing \ref{config-1} and \ref{config-2}
+in terms of total solution time, derivative evaluation time, and
+solver internal time. Key findings are as follows.
+
+\begin{table*}[t]
+  \scriptsize
+  \centering
+  \caption{Details of Numerical Experiment Settings}
+  \begin{tabular}{|l|c|c|c|c|} 
+    \hline
+    & {\textbf{MadNLP.jl + SIMDiff.jl + cuSOLVER}} 
+    & {\textbf{MadNLP.jl + SIMDiff.jl + Ma27}} 
+    & {\textbf{Ipopt + AMPL + Ma27}}
+    & {\textbf{Ipopt + JuMP.jl + Ma27}}\\
+    &\textbf{(gpu)} &\textbf{(cpu)} &\textbf{(cpu)}& \textbf{(cpu)}\\
+    \hline
+    \textbf{Optimization Solver} & \multicolumn{2}{c|}{MadNLP.jl (dev)$^*$} & \multicolumn{2}{c|}{Ipopt (v3.13.3)} \\
+    \hline
+    \textbf{Derivative Evaluations} & \multicolumn{2}{c|}{SIMDiff.jl (dev)$^*$} &  AMPL Solver Library & JuMP.jl (v1.12.0)\\
+    \hline
+    \textbf{Linear Solver} &  cuSOLVER (v11.4.5) &\multicolumn{3}{c|}{Ma27 (v2015.06.23)}\\
+    \hline
+    \textbf{Hardware} & NVIDIA Quadro GV100 & \multicolumn{3}{c|}{Intel Xeon Gold 6140}\\ 
+    \hline
+  \end{tabular}\\
+  $^*$Specific commit hashes are available at \url{https://github.com/sshin23/pscc-2024-opf-gpu}
+  \label{tbl:settings} 
+\end{table*}
+\begin{table*}[t]
+  \scriptsize 
+  \centering
+  \caption{Numerical Results}
+  \input{table.tex}\\
+  $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.
+  \label{tbl:results}
+\end{table*}
+\begin{table*}[t]
+  \scriptsize
+  \centering
+  \caption{Solution Quality}
+  \label{tbl:quality}
+  \input{table-2.tex}\\
+  $^\dag$Directly computed by using the callback functions. $^\ddag$Reported by Ipopt output.
+\end{table*}
+\begin{figure}[t]
+  \includegraphics[width=.45\textwidth]{speedup-sol.pdf}
+  \caption{Speedup achieved by using GPUs.}
+  \label{fig:speedup}
+\end{figure}
+
+First, when comparing the solvers' performance in terms of the
+interior point method (IPM) iteration counts, MadNLP.jl is as
+efficient as the state-of-the-art solver Ipopt. The IPM iteration
+count is nearly the same as that of Ipopt for achieving the same level
+of accuracy in the final solution (see Table \ref{tbl:quality}). This
+suggests that running mathematically equivalent operations on GPUs can
+yield a similar degree of effectiveness in terms of IPM convergence.
+
+Next, we discuss the effectiveness of parallel AD on GPUs. We observe that even on CPUs, SIMDiff.jl
+is substantially faster than the AD routines
+implemented in AMPL or JuMP.jl. This efficiency arises from SIMDiff.jl
+obtaining the derivative functions in the form of type-stable,
+compiled code. Derivative evaluation using the compiled LLVM
+code, specific to the model equation, generated by the Julia compiler
+is significantly faster than the
+derivative evaluations implemented in AMPL and JuMP.jl. When comparing
+SIMDiff.jl running on CPUs and GPUs, we observe a further speedup of
+up to almost 20x for large instances. This demonstrates that
+parallelization of AD brings significant computational gain, enabled
+by the SIMD abstraction of NLPs, particularly because most operations
+in the derivative evaluations are either \ref{eqn:pattern-1} or
+\ref{eqn:pattern-2} operations, which are highly effective on GPUs.
+
+While the speedup achieved by linear solvers is only moderate, this
+has a high impact on the overall speedup, as linear solver time
+constitutes a significant portion of the total solution time. For
+large instances, approximately 4x speedup can be achieved, though this
+can vary depending on the instances. For example, for case24464\_goc,
+the GPU linear solver was slower than the CPUs. The investigation of
+under which circumstances cuSOLVERRF is more effective warrants
+further research.
+
+Solver internal time could also be significantly accelerated through
+GPU utilization. We can observe that the speedup in solver internal
+operations is consistently greater than the speedup in linear
+solvers. However, due to the frequent use of \ref{eqn:pattern-3}
+operations, the speedup in solver internal operations is lesser than
+that of derivative evaluations.
+
+Overall, our GPU implementation exhibits significant speedup across
+all components: derivative evaluation, linear algebra, and solver
+internal computation, resulting in substantial gains in total solution
+time. The results indicate that GPUs become more effective for
+large-scale instances, particularly when the number of variables is
+greater than 20,000. Notably, for the largest instance,
+case30000\_goc, our GPU implementation is 4.5 times faster than our
+CPU implementation and 9.5 times faster than the solver configuration
+with state-of-the-art tools (Ipopt, JuMP.jl, and Ma27). This demonstrates
+that GPUs can bring significant computational gains for large-scale AC
+OPF problems, enabling the solution of previously inconceivable
+problems due to the limitations of existing CPU-based solution tools.
+
+\section{Conclusions and Future Outlook}\label{sec:conc}
+We have presented a highly efficient nonlinear optimization framework
+for solving large-scale AC OPF problems. By leveraging the SIMD
+abstraction of NLPs and a condensed-space interior point method, we
+have effectively eliminated the need for serial computations, enabling
+the implementation of a solution framework that can run entirely on
+GPUs. Our method has demonstrated promising results, achieving a 4x
+speedup when compared to CPU implementations for large-scale optimal
+power flow problems. Notably, our approach outperforms one of the
+state-of-the-art CPU-based implementations by approximately 10
+times. Therefore, our proposed computational framework, along with our
+packages MadNLP.jl and SIMDiff.jl, represents a significant
+advancement in scalable solution large-scale OPF problems. These
+findings underscore the potential of accelerated computing
+in large-scale optimization area, like AC OPF. However,
+the condensation procedure leads to an increase in the condition
+number of the KKT system, resulting in decreased final solution
+accuracy. Addressing the challenges posed by ill-conditioning remains
+an important aspect for future work. In the following paragraphs, we
+discuss some remaining open questions and future outlooks.
+
+\paragraph*{Obtaining Higher Numerical Precisions} While we have
+focused on the interior point method, other constrained optimization
+paradigms, such as penalty methods, augmented Lagrangian methods, and
+sequential quadratic programming, exist, and similar strategies based
+on condensed linear systems can be developed. It would be valuable to
+investigate which one can best handle the ill-conditioning issue of
+the condensed KKT system and achieve the highest degree of accuracy.
+
+\paragraph*{Security-Constrained, Multi-Period, Distribution OPFs}
+Though the proposed method has demonstrated significant computational
+advantages for transmission AC OPF problems, efficient CPUs can still
+handle these problems reasonably well. We anticipate more substantial
+performance gains for larger-scale optimization problems, such as
+security-constrained and multi-period OPFs or joint optimization
+problems involving transmission, distribution, and gas network
+systems. We are interested in exploring Schur complement-based
+decomposition approaches, combined with the condensation-based
+strategy, similarly to \cite{pacaud2023parallel}, to enable even greater scalability.
+
+\paragraph*{Implementation of Sparse Cholesky Solver} While cuSOLVERRF
+has been effective for solving the condensed KKT systems using LU
+factorization, Cholesky factorization holds promise for better
+performance due to lower computational complexity and the ability to
+reveal the inertia of the KKT system. We aim to explore other options
+for solving KKT systems, including sparse Cholesky implementations
+\cite{chen2008algorithm,pineda2022theseus,regev2023hykkt}.
+
+\paragraph*{Portability} Our implementation is currently only tested
+on NVIDIA GPUs, but our GPU implementation is largely based on
+KernelAbstractions.jl in Julia, which should be compatible with
+various GPU architectures, including AMD, Intel, and Apple GPUs. By
+incorporating cross-architecture linear solvers, we envision achieving
+a portable accelerated NLP framework for different accelerator
+architectures.
+
+\paragraph*{Optimization with Data-Driven Surrogates} With the growing
+importance of emerging energy storage and conversion technologies like
+batteries and fuel cells, incorporating their physics into long-term
+economic performance is crucial. Data-driven strategies are often
+considered due to the impracticality of first-principle modeling. Our
+GPU-accelerated method opens doors to the integration of data-driven
+surrogate models into state-of-the-art optimization frameworks. This
+would allow us to effectively incorporate the governing physics of
+these systems while benefiting from the speed and power of GPU
+acceleration.
+
+\bibliographystyle{IEEEtran}
+\bibliography{main}
+\vspace{0.1cm}
+\begin{flushright}
+	\scriptsize \framebox{\parbox{2.5in}{Government License: The
+			submitted manuscript has been created by UChicago Argonne,
+			LLC, Operator of Argonne National Laboratory (``Argonne").
+			Argonne, a U.S. Department of Energy Office of Science
+			laboratory, is operated under Contract
+			No. DE-AC02-06CH11357.  The U.S. Government retains for
+			itself, and others acting on its behalf, a paid-up
+			nonexclusive, irrevocable worldwide license in said
+			article to reproduce, prepare derivative works, distribute
+			copies to the public, and perform publicly and display
+			publicly, by or on behalf of the Government.
+The Department of Energy will provide public access to these results of federally sponsored research in accordance with the DOE Public Access Plan.
+http://energy.gov/downloads/doe-public-access-plan.
+}}
+	\normalsize
+\end{flushright}
 

### 2023-07-25 paper/speedup-sol.tex
populated all sections

% Recommended preamble:
% \usetikzlibrary{arrows.meta}
% \usetikzlibrary{backgrounds}
% \usepgfplotslibrary{patchplots}
% \usepgfplotslibrary{fillbetween}
% \pgfplotsset{%
%     layers/standard/.define layer set={%
%         background,axis background,axis grid,axis ticks,axis lines,axis tick labels,pre main,main,axis descriptions,axis foreground%
%     }{
%         grid style={/pgfplots/on layer=axis grid},%
%         tick style={/pgfplots/on layer=axis ticks},%
%         axis line style={/pgfplots/on layer=axis lines},%
%         label style={/pgfplots/on layer=axis descriptions},%
%         legend style={/pgfplots/on layer=axis descriptions},%
%         title style={/pgfplots/on layer=axis descriptions},%
%         colorbar style={/pgfplots/on layer=axis descriptions},%
%         ticklabel style={/pgfplots/on layer=axis tick labels},%
%         axis background@ style={/pgfplots/on layer=axis background},%
%         3d box foreground style={/pgfplots/on layer=axis foreground},%
%     },
% }

\begin{tikzpicture}[/tikz/background rectangle/.style={fill={rgb,1:red,1.0;green,1.0;blue,1.0}, fill opacity={1.0}, draw opacity={1.0}}, show background rectangle]
\begin{axis}[point meta max={nan}, point meta min={nan}, legend cell align={left}, legend columns={1}, title={}, title style={at={{(0.5,1)}}, anchor={south}, font={{\fontsize{14 pt}{18.2 pt}\selectfont}}, color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, rotate={0.0}, align={center}}, legend style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, line width={1}, solid, fill={rgb,1:red,1.0;green,1.0;blue,1.0}, fill opacity={1.0}, text opacity={1.0}, font={{\fontsize{8 pt}{10.4 pt}\selectfont}}, text={rgb,1:red,0.0;green,0.0;blue,0.0}, cells={anchor={center}}, at={(0.02, 0.98)}, anchor={north west}}, axis background/.style={fill={rgb,1:red,1.0;green,1.0;blue,1.0}, opacity={1.0}}, anchor={north west}, xshift={1.0mm}, yshift={-1.0mm}, width={150.4mm}, height={99.6mm}, scaled x ticks={false}, xlabel={number of variables}, x tick style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, opacity={1.0}}, x tick label style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, opacity={1.0}, rotate={0}}, xlabel style={at={(ticklabel cs:0.5)}, anchor=near ticklabel, at={{(ticklabel cs:0.5)}}, anchor={near ticklabel}, font={{\fontsize{11 pt}{14.3 pt}\selectfont}}, color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, rotate={0.0}}, xmode={log}, log basis x={10}, xmajorgrids={true}, xmin={1042}, xmax={208624}, xticklabels={{$10^{4}$,$10^{5}$}}, xtick={{10000.0,100000.0}}, xtick align={inside}, xticklabel style={font={{\fontsize{8 pt}{10.4 pt}\selectfont}}, color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, rotate={0.0}}, x grid style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={0.1}, line width={0.5}, solid}, xticklabel pos={left}, x axis line style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, line width={1}, solid}, scaled y ticks={false}, ylabel={speedup}, y tick style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, opacity={1.0}}, y tick label style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, opacity={1.0}, rotate={0}}, ylabel style={at={(ticklabel cs:0.5)}, anchor=near ticklabel, at={{(ticklabel cs:0.5)}}, anchor={near ticklabel}, font={{\fontsize{11 pt}{14.3 pt}\selectfont}}, color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, rotate={0.0}}, ymode={log}, log basis y={10}, ymajorgrids={true}, ymin={0.042449470320200695}, ymax={37.677607781546705}, yticklabels={{$10^{-1}$,$10^{0}$,$10^{1}$}}, ytick={{0.1,1.0,10.0}}, ytick align={inside}, yticklabel style={font={{\fontsize{8 pt}{10.4 pt}\selectfont}}, color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, rotate={0.0}}, y grid style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={0.1}, line width={0.5}, solid}, yticklabel pos={left}, y axis line style={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, line width={1}, solid}, colorbar={false}]
    \addplot[color={rgb,1:red,0.0;green,0.6056;blue,0.9787}, name path={90991ba5-e570-4efd-b7ef-9814b55ce668}, only marks, draw opacity={1.0}, line width={0}, solid, mark={*}, mark size={3.0 pt}, mark repeat={1}, mark options={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, fill={rgb,1:red,0.0;green,0.6056;blue,0.9787}, fill opacity={0.0}, line width={0.75}, rotate={0}, solid}]
        table[row sep={\\}]
        {
            \\
            1042.0  0.12132541013470992  \\
            1468.0  0.10314692434543314  \\
            4254.0  0.3591290345722146  \\
            5432.0  0.4318689855256274  \\
            11192.0  0.493180776719653  \\
            17128.0  0.800124080241001  \\
            19008.0  0.9921851426883461  \\
            23238.0  1.0782015249239902  \\
            24540.0  1.2249585614191905  \\
            25086.0  1.0314377489080664  \\
            35270.0  1.8122793796033891  \\
            36696.0  1.9603940524376853  \\
            37872.0  1.8070726670030763  \\
            38814.0  1.908074706072124  \\
            41398.0  1.727056190842467  \\
            42532.0  1.8603003250968289  \\
            76804.0  3.761460351555901  \\
            78748.0  1.5909204206596708  \\
            83572.0  2.421056606260939  \\
            85568.0  2.4138090392087532  \\
            96750.0  1.3842870098320392  \\
            117370.0  2.213283662070483  \\
            179562.0  2.511302487858865  \\
            203374.0  0.5769039469634973  \\
            208624.0  4.274713987006297  \\
        }
        ;
    \addlegendentry {total solution}
    \addplot[color={rgb,1:red,0.8889;green,0.4356;blue,0.2781}, name path={45fed1e9-b750-4c50-97d6-e4d2ebf7d495}, only marks, draw opacity={1.0}, line width={0}, solid, mark={square*}, mark size={3.0 pt}, mark repeat={1}, mark options={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, fill={rgb,1:red,0.8889;green,0.4356;blue,0.2781}, fill opacity={0.0}, line width={0.75}, rotate={0}, solid}]
        table[row sep={\\}]
        {
            \\
            1042.0  0.1489498744199213  \\
            1468.0  0.1727159137624846  \\
            4254.0  0.4340465934605082  \\
            5432.0  0.5434423830639694  \\
            11192.0  0.8004594515472352  \\
            17128.0  2.082689485354689  \\
            19008.0  2.4770197579261883  \\
            23238.0  3.05899341669706  \\
            24540.0  1.8717332637281556  \\
            25086.0  2.906247958020815  \\
            35270.0  4.496479681105164  \\
            36696.0  3.995367978071199  \\
            37872.0  5.152797200922378  \\
            38814.0  4.421274460394101  \\
            41398.0  4.854859953917821  \\
            42532.0  4.4411634259064  \\
            76804.0  11.566047092387956  \\
            78748.0  5.45272118651195  \\
            83572.0  6.085193656249472  \\
            85568.0  6.910004676366815  \\
            96750.0  6.369970804347526  \\
            117370.0  6.95452612891449  \\
            179562.0  10.976999346486657  \\
            203374.0  18.838803890773352  \\
            208624.0  12.675774859535938  \\
        }
        ;
    \addlegendentry {AD}
    \addplot[color={rgb,1:red,0.2422;green,0.6433;blue,0.3044}, name path={597ac343-eab2-47fd-bfb0-3d1869cc7b86}, only marks, draw opacity={1.0}, line width={0}, solid, mark={diamond*}, mark size={3.0 pt}, mark repeat={1}, mark options={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, fill={rgb,1:red,0.2422;green,0.6433;blue,0.3044}, fill opacity={0.0}, line width={0.75}, rotate={0}, solid}]
        table[row sep={\\}]
        {
            \\
            1042.0  0.12134780833101436  \\
            1468.0  0.08489894064040139  \\
            4254.0  0.3400556528465654  \\
            5432.0  0.4015885378861249  \\
            11192.0  0.4178468802690407  \\
            17128.0  0.6671440282545749  \\
            19008.0  0.8418469098787906  \\
            23238.0  0.8970213864029066  \\
            24540.0  1.1302597892149948  \\
            25086.0  0.8736098235492576  \\
            35270.0  1.6332401723649006  \\
            36696.0  1.8665104859925663  \\
            37872.0  1.5712590526539294  \\
            38814.0  1.7310996666393024  \\
            41398.0  1.5290569916050272  \\
            42532.0  1.7474748794876105  \\
            76804.0  3.488972057355353  \\
            78748.0  1.395946330507901  \\
            83572.0  2.3246624958297284  \\
            85568.0  2.2224099048005086  \\
            96750.0  1.2509789777402023  \\
            117370.0  2.0490766644040312  \\
            179562.0  2.389075257357081  \\
            203374.0  0.48297264689483094  \\
            208624.0  4.199296385639948  \\
        }
        ;
    \addlegendentry {linear solver}
    \addplot[color={rgb,1:red,0.7644;green,0.4441;blue,0.8243}, name path={0daca70f-5467-4208-80b3-fa1e8e4d7149}, only marks, draw opacity={1.0}, line width={0}, solid, mark={star}, mark size={3.0 pt}, mark repeat={1}, mark options={color={rgb,1:red,0.0;green,0.0;blue,0.0}, draw opacity={1.0}, fill={rgb,1:red,0.7644;green,0.4441;blue,0.8243}, fill opacity={0.0}, line width={0.75}, rotate={0}, solid}]
        table[row sep={\\}]
        {
            \\
            1042.0  0.11053845268205038  \\
            1468.0  0.18528018635787544  \\
            4254.0  0.3971318279284334  \\
            5432.0  0.502547137067705  \\
            11192.0  0.7925028620399515  \\
            17128.0  1.4642569252535946  \\
            19008.0  1.5724747212032781  \\
            23238.0  1.8834530645984655  \\
            24540.0  1.588367729468711  \\
            25086.0  1.6835889681072704  \\
            35270.0  2.4543461608339245  \\
            36696.0  2.2839061860763974  \\
            37872.0  2.5758891325324624  \\
            38814.0  2.548435587364985  \\
            41398.0  2.5197402455776663  \\
            42532.0  2.2091833306911117  \\
            76804.0  4.095029335003687  \\
            78748.0  2.4198733504538077  \\
            83572.0  2.638122640974531  \\
            85568.0  2.8374797565002723  \\
            96750.0  2.5465858998015847  \\
            117370.0  2.5350811531961326  \\
            179562.0  3.059535361656725  \\
            203374.0  3.8895762766124165  \\
            208624.0  3.6340740717834334  \\
        }
        ;
    \addlegendentry {solver internal}
    \addplot[color={rgb,1:red,0.502;green,0.502;blue,0.502}, name path={8f65f3c3-f37c-42b0-ad0a-2ff41a891b22}, draw opacity={1.0}, line width={1}, solid, forget plot]
        table[row sep={\\}]
        {
            \\
            5.204406012731031  1.0  \\
            4.1769648153550915e7  1.0  \\
        }
        ;
\end{axis}
\end{tikzpicture}

### 2023-07-27 paper/main.tex
paper ready for Mihai's review

--- 2023-06-30 abstract/main.tex
@@
-% \input{preamble}
-\title{Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}
-
-\author{
-\IEEEauthorblockN{Sungho Shin and Mihai Anitescu}
-\IEEEauthorblockA{
-  Mathematics and Computer Science Division\\
-  Argonne National Laboratory\\
-  Lemont, IL, USA\\
-sshin@anl.gov, anitescu@mcs.anl.gov}
-\and
-\IEEEauthorblockN{François Pacaud}
-\IEEEauthorblockA{Centre Automatique et Systèmes\\
-  Mines Paris - PSL \\
-  Paris, France\\
-francois.pacaud@minesparis.psl.eu}
+\input{preamble}
+\title{Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods 
 }
@@
 \maketitle
+\begin{abstract}
+  This paper introduces a novel computational framework for solving
+alternating current (AC) optimal power flow (OPF) problems using
+graphics processing units (GPUs). While GPUs have demonstrated
+remarkable performance in various computing domains, their application
+in AC OPF has been limited due to challenges associated with porting
+sparse automatic differentiation (AD) and sparse linear solver
+routines to GPUs. We aim to address these issues with two key
+strategies.  First, we utilize a single-instruction, multiple-data
+(SIMD) abstraction of nonlinear programs (NLP). This approach enables
+the specification of model equations while preserving their
+parallelizable structure, and in turn, facilitates the implementation
+of AD routines that can exploit such structure.  Second, we employ a
+condensed-space interior-point method (IPM) with an inequality
+relaxation strategy. This technique involves relaxing equality
+constraints to inequalities and condensing the Karush-Kuhn-Tucker
+system into a much smaller positive definite matrix. This strategy
+offers the key advantage of being able to factorize the KKT matrix
+without numerical pivoting, which has hampered the parallelization of
+the algorithm.  By combining these two strategies, we can perform the
+majority of operations on GPUs while keeping the data residing in the
+device memory only. Comprehensive numerical benchmark results showcase
+the substantial computational advantage of our approach. Remarkably,
+for solving large-scale AC OPF problems to a moderate accuracy, our
+implementations—MadNLP.jl and SIMDiff.jl—running on NVIDIA GPUs
+achieve an order of magnitude speedup compared to state-of-the-art
+tools running on contemporary CPUs.
+\end{abstract}
 
-While graphics processing units (GPUs) have showcased impressive capabilities in various computing domains, their utilization in large-scale constrained nonlinear optimization regimes, such as alternating current (AC) optimal power flow (OPF) problems, has been somewhat limited.
-This limitation stems from the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms (Anitescu et.
-al.
-2021).
-Although GPU computation can accelerate various other parts of the optimization process, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations.
-To fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU that incorporates AD, linear algebra, and optimization while minimizing data transfers to and from host memory.
+\begin{IEEEkeywords}
+  nonlinear programming, automatic differentiation, GPU computing, optimal power flow
+\end{IEEEkeywords}
+\section{Introduction}
 
-This paper presents a comprehensive computational framework and the associated software implementations for solving AC OPF problems on GPUs.
-Our approach utilizes the following techniques: (i) condensed-space interior-point methods (IPMs) with an inequality relaxation strategy, (ii) sparse matrix factorization with a fixed pivot sequence, and (iii) a single-instruction, multiple-data (SIMD) abstraction of nonlinear programs.
-Specifically, our method relaxes power flow equality constraints by allowing small violations, which enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space through the condensation procedure.
-Although this strategy is not new (see Nocedal and Wright, 2006), it has traditionally been considered less efficient than the standard full-space method due to increased fill-in in the sparse factorization.
-However, when implemented on GPUs, it offers the key advantage of guaranteeing positive definiteness in the condensed KKT system through the application of standard regularization techniques, which in turn, allows for the utilization of linear solvers with a fixed numerical pivot sequence (so-called refactorization).
-An efficient implementation of the sparse refactorization is available as part of the CUDA library, facilitating the implementation of efficient KKT system solutions on GPUs.
-Although this method is susceptible to ill-conditioning, our results demonstrate that the solver is robust enough to solve problems with a relative accuracy of $10^{-6}$.
+While graphics processing units (GPUs) have demonstrated remarkable
+capabilities in various computing domains, their adoption in
+large-scale constrained nonlinear programming (NLP) regimes, such as
+alternating current (AC) optimal power flow (OPF) problems, has faced
+some limitations. One of the primary challenges arises from the
+automatic differentiation (AD) of sparse model equations and the
+parallel factorization of indefinite sparse matrices, which are
+commonly encountered within constrained numerical optimization tasks
+\cite{anitescu2021targeting}. While GPU computation can trivially
+accelerate several parts of the optimization process, especially
+various internal computations within the optimization solver, the
+sluggish data transfer between host and device memory hampers the
+ad-hoc implementation of GPU accelerations (Fig. \ref{fig:simd}). To
+fully leverage the potential that modern GPU hardware has to offer, it
+becomes imperative to have a comprehensive computational framework for
+optimization on GPUs. That is, we need an AD/algebraic
+modeling framework, sparse linear solvers, and NLP solvers that can
+operate entirely on the GPU. Specifically, for the best performance,
+both the problem data and the solver's intermediate computational data
+must be exclusively resident within the device memory, with the
+majority of operations executed on the GPU.
 
-Furthermore, by leveraging the SIMD abstraction of nonlinear programs, which preserves the parallelizable structure in the model, the model functions and derivative evaluations can be parallelized, thereby facilitating evaluations on the GPU.
-We demonstrate that the AC power flow model is particularly well-suited for this abstraction as it involves repetitive expressions for each type of component in the model (e.g., buses, lines, generators), and the number of computational patterns does not increase with the network's size.
-This structure has been effectively utilized by Gravity (Hijazi et.
-al.
-2018), demonstrating significant acceleration in model evaluations.
+\begin{figure}[t]
+  \centering
+  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
 
-The paper will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach.
-Our method is implemented in our packages, a nonlinear optimization solver MadNLP.jl and an automatic differentiation tool SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library.
-We compare our method against the standard CPU approach (W\"{a}chter and Beigler, 2006) and the recently developed reduced-space interior-point method on GPUs (Pacaud et.
-al., 2022) using the standard data available in pglib-opf.
-Preliminary results indicate that our proposed framework has the potential to accelerate the solution of AC OPF problems up to moderate tolerances ($10^{-6}$) by an order of magnitude compared to existing tools such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances.
-We will conclude with a discussion on the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.
+    \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
+    \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
+    \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};
 
-% \bibliographystyle{plain}
-% \bibliography{main}
+    \node[align=center] at (1.25, 3.75) {Data};
+    \node[align=center] at (6.75, 3.75) {Data};
+
+    % Host Memory
+    \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};
+
+    % Device Memory
+    \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};
+
+    % Arrow with Dashed Line
+    \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
+    \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
+    \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;
+
+    \def\rows{10}
+    \def\cols{10}
+    \def\elementwidth{2}
+    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
+
+      % Draw Grid
+      \foreach \i in {0,...,\rows} {
+        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
+      }
+      \foreach \i in {0,...,\cols} {
+        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
+      }
+    }
+  \end{tikzpicture}
+  \caption{A schematic description of host (CPU) and device (GPU) memory.}\label{fig:memory}
+\end{figure}
+
+This paper presents our approach to implement a comprehensive
+computational framework for solving large-scale AC OPF problems on
+NVIDIA GPUs, along with the associated software implementations:
+SIMDiff.jl \cite{simdiff}, an algebraic modeling/AD tool, and MadNLP.jl \cite{madnlp}, an
+NLP solver. Our approach incorporates two novel strategies: (i) a
+single-instruction, multiple-data (SIMD) abstraction of nonlinear
+programs (NLPs), enabling streamlined parallel AD on GPUs, and (ii) a
+condensed-space interior-point method (IPM) with an inequality
+relaxation strategy, which facilitates the use of highly efficient
+{\it refactorization} routines for sparse matrix factorization with
+fixed pivot sequences.
+
+While derivative evaluation can be generally cheaper than linear
+algebra operations, our numerical results show that AD often
+constitutes more than half of the total solver time when using
+off-the-shelf AD implementations like JuMP.jl or AMPL. Instead of the
+general-purpose tools, our method leverages a specialized AD
+implementation based on the SIMD abstraction of NLPs. This abstraction
+alllows to preserve the parallelizable structure within the model
+equations, facilitating efficient derivative evaluations on the
+GPU. The AC power flow model is particularly well-suited for this
+abstraction as it involves repetitive expressions for each component
+type (e.g., buses, lines, generators), and the number of computational
+patterns does not increase with the network's size. Numerical results
+reported in this paper demonstrate that our proposed strategies can
+achieve over 20 times speedup by running on GPU. In comparison to
+general AD implementations on CPUs (such as AMPL and JuMP.jl), our
+GPU-based method can be approximately 50 times faster.
+
+Linear algebra operations, especially sparse indefinite matrix
+factorization, are typically the bottleneck in NLP solution methods.
+Parallelizing this operation has been considered to be challenging,
+primarily due to the need for numerical pivoting, which is sequential
+in nature. However, when the matrix can be factorized without
+numerical pivoting, a significant part of the operation can be
+parallelized, and the numerical factorization can be efficiently
+performed on GPUs. We develop a condensed-space IPM strategy that
+allows the use of sparse matrix factorization routines without
+numerical pivoting. This strategy relaxes equality constraints by
+permitting small violations, which enables expressing the
+Karush-Kuhn-Tucker (KKT) system entirely in the primal space through
+the condensation procedure. Although this strategy is not new
+\cite{nocedal2006numerical}, it has traditionally been considered less
+efficient than the standard full-space method due to increased nonzero
+entries in the KKT system. However, when implemented on GPUs, it
+offers the key advantage of ensuring positive definiteness in the
+condensed KKT system. This, in turn, allows for the utilization
+of linear solvers with a fixed numerical pivot sequence (known as
+refactorization), and an efficient implementation is available in the CUDA library.
+Although this method is susceptible to numerical stability issues due to
+increased condition number in the KKT system, our results demonstrate
+that the solver is robust enough to solve problems with a relative
+accuracy of $\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$.
+
+We present numerical benchmark results to showcase the efficiency of
+our method, utilizing our implementations: MadNLP.jl and
+SIMDiff.jl. The solution of the KKT system is performed using the
+external cuSOLVER library. To assess the performance of our method, we
+compare it against standard CPU approaches using the data available in
+pglib-opf \cite{wachter2006implementation}.  Our benchmark results
+demonstrate that our proposed computational framework has significant
+potential for accelerating the solution of AC OPF problems, especially
+when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
+Specifically, when running on NVIDIA GPUs, our method achieves a
+notable 4x speedup compared to our solver running on CPU, in the case
+of the largest instance. Moreover for the same instance, our approach
+surpasses the performance of existing tools (such as Ipopt interfaced
+with JuMP.jl) by an order of magnitude.  This finding underscores the
+importance of harnessing the computational power of GPUs for tackling
+the computatoinal challenges in power systems.
+
+\paragraph*{Related Work}
+Several recent works have explored the use of GPUs for large-scale
+nonlinear optimization problems. Cao et al. \cite{cao2016augmented}
+proposed an augmented Lagrangian interior-point approach that employs
+augmented Lagrangian outer iteration and the treatment of linear
+systems using a preconditioned conjugate gradient method.  Another
+recent development is the hybrid (direct-iterative) KKT system solver
+specifically designed for GPUs \cite{regev2023hykkt}.  Before the
+introduction of the condensed space IPM with an inequality relaxation
+strategy, the authors have investigated the use of reduction
+strategies (state variable elimination) to treat KKT systems in a
+dense form on the GPU
+\cite{pacaud2023parallel,pacaud2022feasible,pacaud2023accelerating,cole2023exploiting}.
+An NLP solver for high-performance computers (HPC) with GPU
+accelerators called HiOP has been under development
+\cite{hiop_techrep}, which has a similar scope as our solver
+MadNLP.jl.
+
+\paragraph*{Organization}
+The paper is organized as follows. In the remainder of the current
+section, we introduce the mathematical notation. In Section
+\ref{sec:prelim}, we provide general preliminary knowledge on
+numerical optimization and GPU computing. In Section \ref{sec:simd},
+we present the SIMD abstraction of NLPs and their advantages in terms
+of implementing parallel AD. Section \ref{sec:ipm} presents the
+optimization algorithm under study, the condensed-space IPM with an
+inequality relaxation strategy. Section \ref{sec:num} presents the
+numerical results, comparing our approach with other state-of-the-art
+solution methods on the CPUs. Finally, conclusions and future outlooks
+are given in Section \ref{sec:conc}.
+
+\paragraph*{Notation}
+We denote the set of real numbers and the set of integers by
+$\mathbb{R}$ and $\mathbb{I}$. We let $[M]:=\{1,2,\cdots,M\}$. We let
+$[v_i]_{i\in[M]}:=[v_1;v_2;\cdots,v_M]$.  A vector of ones with an
+appropriate size is denoted by $\boldsymbol{1}$. An identiy matrix
+with an appropriate size is denoted by $I$. For matrices $A$ and $B$,
+$A\succ(\succeq) B$ indicates that $A-B$ is positive (semi)-definite
+while $A>(\geq) B$ denotes a componentwise inequality. We use the
+following convention for any symbol $x$: $X:=\diag(x)$.
+
+\section{Preliminaries}\label{sec:prelim}
+This section covers three essential background topics: numerical
+optimization, AD, and GPU computing.
+
+\subsection{Numerical Optimization}\label{sec:numopt}
+We consider NLPs of the following form:
+\begin{subequations}\label{eqn:cpt}
+  \begin{align}
+    \min_{x^\flat \leq x \leq x^\sharp}\;
+    & f(x)\\
+    \st\;
+    & g(x) =0. \label{eqn:cpt-con}
+  \end{align} 
+\end{subequations}
+Numerous solution algorithms have been developed in the NLP
+literature. In terms of strategies to deal with inequality
+constraints, the NLP solution algorithms can be broadly classified
+into active-set methods and IPMs \cite{nocedal2006numerical}.
+Active-setmethods aim to find the set of active consraints associated
+with the optimal solution in a combinatorial manner, while IPMs
+replace inequality constraints with smooth barrier functions, enabling
+the use of Newton-type second-order algorithms. IPMs are known to be
+more scalable for problems with a large number of constraints and
+suitable for parallelization, thanks to the fixed sparsity pattern of
+the KKT matrix. Given these advantages, we have chosen IPMs as the
+backbone algorithm for developing our optimization methods on GPUs.
+
+In terms of practical computations, three key components play
+vital roles: derivative evalutions (often provided by the AD
+capabilities of the algebraic modeling languages), linear algebra
+operations, and various internal computations within the
+solver. Notably, most of the computational efforts are delegated to
+the external linear solver and AD library, while the optimization solver
+orchestrates the operation of these tools to drive the solution
+iterate towards the stationary point of the optimization problem.
+
+Since the successful implementation of the open-source IPM solver in
+Ipopt, many subsequent implementations of NLP solvers
+\cite{chiang2014structured,rodriguez2023scalable,shin2021graph} have
+based their implementation on Ipopt
+\cite{wachter2006implementation}. We also use Ipopt as our main
+reference for the IPM implementation. Below, we outline the overall
+computational procedure employed wihtin the NLP solution frameworks:
+
+1) Given the current primal-dual iterate $(x^{(\ell)},y^{(\ell)},
+z^{\flat(\ell)},z^{\sharp(\ell)})$, the AD package evaluates:
+\begin{align*}
+  \nabla_x f(x^{(\ell)}),\quad
+  \nabla_x g(x^{(\ell)}),\quad
+  \nabla^2_{xx} \mathcal{L}(x^{(\ell)},y^{(\ell)},z^{\flat(\ell)},z^{\sharp(\ell)}),
+\end{align*}
+where
+\begin{align*}
+  \mathcal{L}(x,y,z^{\flat},z^{\sharp}):=f(x) - y^\top
+  g(x) - z^\flat (x-x^\flat) - z^\sharp (x^\sharp-x).
+\end{align*}
+
+2) The following sparse indefinite system (known as the KKT system) is
+solved using sparse indefinite factorization (typically, via sparse
+LBL$^\top$ factorization) and triangular solve routines.
+\begin{align}\label{eqn:kkt-indefinite}
+  &\begin{bmatrix}
+    W^{(\ell)}  + \Sigma^{(\ell)} + \delta^{(\ell)}_w I& A^{(\ell)\top}\\
+    A^{(\ell)} & -\delta_c^{(\ell)} I\\
+  \end{bmatrix}
+  \begin{bmatrix}
+    \Delta x\\
+    \Delta y\\
+  \end{bmatrix}=
+  \begin{bmatrix}
+    r_x^{(\ell)}\\
+    r_y^{(\ell)}\\
+  \end{bmatrix},
+\end{align}
+where
+\begin{align*}
+  W^{(\ell)}
+  &:=\nabla^{2}_{xx}\mathcal{L}(x^{(\ell)},y^{(\ell)},z^{\flat(\ell)},z^{\sharp(\ell)}),
+  &&A^{(\ell)}:= \nabla_xg(x^{(\ell)})\\
+  \Sigma^{(\ell)}&:= (X^{(\ell)})^{-1}Z^{(\ell)}\\
+  r_x^{(\ell)}
+  &:=\nabla_x f(x^{(\ell)}) - \mu (X^{(\ell)})^{-1} \boldsymbol{1}, 
+  &&r_y^{(\ell)}:=g(x^{(\ell)}),
+\end{align*}
+and $\delta^{(\ell)}_w, \delta^{(\ell)}_c>0$ are the regularization parameters
+determined based on the inertia correction procedure.
+
+3) The optimization solver employs the filter line search procedure to
+determine the step size. The primal-dual iterate is updated by
+applying the determined step size and direction. This process is
+repeated until convergence criteria (typically based on the residual
+to the first-order optimality conditions) are satisfied.
+
+\subsection{AD}
+Numerical differentiation of computer programs can be achieved through
+three different methods: (i) finite difference method, (ii) symbolic
+differentiation, and (iii) AD. The finite difference method suffers
+from numerical rounding errors, and its computational complexity grows
+unfavorably with respect to the number of function arguments, making
+it less preferable unless no other alternatives are
+available. Symbolic differentiation uses computer algebra systems to
+obtain symbolic expressions of first or higher-order
+derivatives. While this method can differentiate functions up to high
+numerical precision, it suffers from "expression swelling" effect and
+struggles to compute the derivatives of long nested expressions in a
+computationally efficient way.
+
+In contrast, AD differentiates computer
+programs by leveraging the computation graph. By inspecting the
+computation graph and applying chain rules, AD
+can efficiently and accurately evaluate derivatives. This approach has
+become the dominant paradigm for derivative computation within the
+scientific computing domain, including NLP and
+machine learning. For large-scale optimization, such as AC OPFs, AD
+tools are often implemented as part of domain-specific modeling
+languages, such as JuMP, CasADi, and AMPL (optimization) and Torch and
+Flux (machine learning).
+
+There are two alternative ways of propagating derivatives through the
+recursive application of chain rules: (i) forward-mode and (ii)
+reverse-mode, which operate in opposite directions (from leaves to
+root and from root to leaves, respectively). Reverse-mode automatic
+differentiation, also known as an adjoint method, has proven to be
+particularly effective for dealing with the function expressions in
+large-scale optimization problems.
+
+The Julia Language, our language of choice, offers convenient and
+efficient ways to implement automatic differentiation. Through the use
+of the multiple dispatch paradigm, \textit{any Julia function},
+including commonly used operations like addition, multiplication,
+trigonometric and exponential functions, among others, can be easily
+overloaded. Multiple dispatch allows functions to be dynamically
+dispatched based on the run-time type, a crucial feature for
+implementing differentiable programming. Several AD implementations
+have been developed in Julia Language, such as ReverseDiff.jl,
+ForwardDiff.jl, Zygote.jl, and JuMP.jl. While these tools are general
+and useful for various applications, they are not optimized for
+evaluating derivatives of AC OPF problems, as they are not designed to
+exploit the parallelizable structures in the model.
+
+\subsection{GPU Computing}\label{sec:gpu}
+With the increasing prevalence of GPU in various scientific computing
+domains, there has been growing interest in leveraging these emerging
+architectures to efficiently solve large-scale NLPs, like AC OPF problems.
+However, adapting the NLP solution algorithms,
+such as IPM, to GPUs presents challenges due to the fundamental
+differences between GPU and CPU programming paradigms. While CPUs
+execute a sequence of instructions on a single input (single
+instruction, single data, or SISD, in Flynn's taxonomy), GPUs run the
+same instruction simultaneously on hundreds of threads using the SIMD
+paradigm (see Fig. \ref{fig:simd}). The SIMD parallelism works well
+for algorithms that can be decomposed into simple instructions running
+entirely in parallel, but not all algorithms fit this
+paradigm. For example, branching in the control flow can hinder
+lockstep execution across multiple threads, and in turn, prevent
+efficient implementations on GPUs. On the other hand, when the
+algorithm's structure allows for efficient parallelization, the SIMD
+parallelism in GPUs can offer orders of magnitudes speedup.
+
+We highlight that the following, arguably quite common, computational
+patterns are particularly effective when implemented on GPUs:
+\begin{align}
+  y&\leftarrow \left[\phi(x; q_j)\right]_{j\in [J]}\tag{Pattern 1}\label{eqn:pattern-1}\\
+  o&\leftarrow  \mathop{\rm{Op}}_{i\in [I]} \psi(x;p_i) \tag{Pattern 2}\label{eqn:pattern-2}\\
+  x&\leftarrow  \chi_{s_1}\circ\cdots\circ \chi_{s_K}(x) \tag{Pattern 3}\label{eqn:pattern-3}
+\end{align}
+Here, $\psi:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{p}}\rightarrow
+\mathbb{R}$ , $\phi:\mathbb{R}^{n_x}\times
+\mathbb{R}^{n_{q}}\rightarrow \mathbb{R}$, and
+$\chi:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{s}}\rightarrow
+\mathbb{R}^{n_x}$ are simple instructions that require only a small number
+of operations; $\mathop{\rm{Op}}$ is a monoid operator on
+$\mathbb{R}\cup\{+\infty,-\infty\}$, such as addition, multiplication,
+maximum, and minimum; and the notation $\chi_{s_k}(x):=\chi(x,s_k)$ is used, and
+assume that $\circ$ is commutative on $\{\chi_{s_k}(\cdot)\}_{\forall
+s_k}$. \ref{eqn:pattern-1} is typically most effective on GPUs, where
+each thread employed can operate independently without needing to
+simultaneously manipulate the same device memory location. While
+\ref{eqn:pattern-2} and \ref{eqn:pattern-3} is less effective as they
+require such a manipulation, but they still can be significantly
+faster than operations in CPUs, as substantial part of the operation
+can still be parallelized by the use of buffers. In simple cases, the
+implementation of these operations can be performed with the standard
+{\tt map} and {\tt mapreduce} programming models. However, in more
+complex cases, especially for \ref{eqn:pattern-3}, the
+implementation may require pre-inspection of memory write-access
+patterns and the use of custom kernels.
+
+Many of the operations required in AD of sparse physical models, as
+well as the application of optimization algorithms, are based on the
+computational patterns mentioned above. For example, an AC OPF model can
+be implemented with 15 different computational patterns, most of which
+fall within the aforementioned categories. Furthermore, the
+computation within optimization solvers, such as forming the
+left-hand-side for the KKT systems, computing the $\|\cdot\|_\infty$
+norm of the constraint violation, and assembling the condensed KKT
+system, can be carried out by using these computational patterns as well. The
+only exception is the factorization of the sparse KKT matrix, which
+requires more sophisticated implementations.
+
+Implementing kernel functions for the above computational patterns in
+the Julia Language is straightforward, as it provides excellent
+high-level interfaces for array and kernel programming for GPU
+arrays. The code can even be device-agnostic, thanks to the portable
+programming capabilities brought by KernelAbstractions.jl. All
+of the AD and optimization capabilities in our
+tools, MadNLP.jl and SIMDiff.jl, are implemented in Julia Language,
+by leveraging its kernel and array programming capabilities.
+
+\begin{figure}[t]
+  \centering 
+  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
+    \node[draw,align=left] (I) at (0,2.75) {
+      $\bullet$ Instruction $g(\cdot,\cdot)$\\
+      $\bullet$ Pointer to array $x$
+    };
+    \node[draw] at (-4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (A\x) at (0,-\x/2) {$s_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (0,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (C\x) at (0,-\x/2) {$y_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \foreach\x in {1,2,3,4,5,7}{
+      \draw[->] (A\x.east) -- (B\x.west);
+      \draw[->] (B\x.east) -- (C\x.west);
+      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
+    }
+    \node at (-4,-2.5) {Input Data};
+    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
+    \node at (4,-2.5) {Output Data};
+  \end{tikzpicture}
+  \label{fig:simd}
+  \caption{A schematic description of SIMD parallelism}
+\end{figure}
+
+\section{SIMD Abstraction of NLPs}\label{sec:simd}
+
+This section describes our implementation of SIMD abstraction and
+sparse AD of the model equations. The abstraction and AD are
+implemented as part of our algebraic modeling language SIMDiff.jl.
+
+\subsection{Abstraction}
+The SIMD abstraction under consideration is as follows:
+\begin{align}\label{eqn:prob}
+  \min_{x^\flat\leq x \leq x^\sharp}
+  & \sum_{l\in[L]}\sum_{i\in [I_l]} f^{(l)}(x; p^{(l)}_i)\\\nonumber
+  \st\; &\forall m\in[M]:\\\nonumber
+  &\begin{aligned}[t]
+    \left[g^{(m)}(x; q_j)\right]_{j\in [J_m]} +\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) =0,
+  \end{aligned}
+  \end{align}
+where $f^{(\ell)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
+$h^{(n)}(\cdot,\cdot)$ are twice differentiable functions with respect
+to the first argument, $\{\{p^{(k)}_i\}_{i\in [N_k]}\}_{k\in[K]}$,
+$\{\{q^{(k)}_{i}\}_{i\in [M_l]}\}_{m\in[M]}$, and
+$\{\{\{s^{(n)}_{k}\}_{k\in[K_n]}\}_{n\in[N_m]}\}_{m\in[M]}$ are
+problem data, which can either be discrete or continuous.  We assume
+that our functions $f^{(l)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
+$h^{(n)}(\cdot,\cdot)$ can be expressed in the form of computational
+graphs of moderate length. One can observe that the problem in
+\eqref{eqn:prob} is expressed by the computational patterns in Section
+\ref{sec:gpu}. In particular, the objective function falls within \ref{eqn:pattern-1},
+the first term in the constraint falls within \ref{eqn:pattern-2}, and the second term in the constraint falls within \ref{eqn:pattern-3}.
+Accordingly, the evaluation and differentiation of the
+model equations in \eqref{eqn:prob} is amenable to SIMD parallelism.
+
+To implement the SIMD abstraction in the modeling environment, the
+algebraic modeling interface in SIMDiff.jl enforces the users to
+specify the model equations in an {\tt Iterable} data type in
+Julia Language. This composite data type consists of an instruction (a julia
+function) and data (a host or device array) over which the instruction
+is executed. This naturally facilitates maintaining the NLP model
+information in the form of SIMD abstraction in \eqref{eqn:prob}, and
+facilitates the model evaluation and differentiation on GPU
+accelerators.
+
+\subsection{Parallel AD}
+Many of the physics-based models, such as AC OPF, have highly repetitive
+structure. One of the manifestations of it is that the mathematical
+statement of the model is concise, even if the practical model may contain
+millions of variables and constraints. This is possible due to the use of
+repetition over a certain index and data sets. For example,
+it suffices to use 15 computational patterns to fully specify the
+AC OPF model. These patterns arise from (1) generation cost, (2) reference
+bus voltage angle constraint, (3-6) active and reactive power flow (to
+and from), (7) voltage angle difference constraint, (8-9) apparent
+power flow limits (to and from), (10-11) power balance equations,
+(12-13) generators' contributions to the power balance equations, and
+(14-15) in/out flows contributions to the power balance
+equations. However, such repetitive structure is not well exploited
+during AD in the standard NLP modeling paradigms.  In
+fact, without the SIMD abstraction, it is difficult for the AD package
+to detect the parallelizable structure within the model, as it will
+require the full inspection of the computational graph. By preserving
+the repetitive structures in the model, one can facilitate the
+implementation of parallelized derivative computations.
+
+Using the multiple dispatch feature of Julia, SIMDiff.jl generates
+highly efficient derivative computation code, specifically compiled
+for each computational pattern in the model. These derivative
+evaluation code can be run over the data in various GPU array formats,
+implemented via array and kernel programming in Julia Language. In
+turn, SIMDiff.jl has the capability to efficiently evaluate first and
+second order derivatives using GPU accelerators.
+
+\subsection{Sparsity Analysis}
+The sparsity analysis is needed for determining the sparsity pattern
+of the evaluated derivatives. In the case of large-sclae sparse
+problems, the initial sparsity analysis of nonlinear expressions can
+be expensive, as the sparsity should be analyzed for potentially
+millions of objective terms and constraints. However, oftentimes
+these analyses are applied for the same computational patterns, and
+the time and memory spent for sparsity analysis can be significantly
+reduced if the repetitive structures are exploited.
+
+SIMDiff.jl exploits the SIMD abstraction of the model equations to
+save the computational cost spent for sparsity analysis.  This is
+accomplished by applying sparsity analysis for the instruction for
+each computational pattern and expand the obtained sparsity pattern
+over the data over which the instruction is executed. Specifically,
+the sparisty analysis code exploits Julia's multiple dispatch feature
+to obtain parameterized sparsity pattern for each instruction, and the
+obtained parameterized sparsity pattern is materialized once the data
+array is given. This process drastically save the computational cost
+for the sparsity analysis.
+
+\section{Condensed-Space IPMs with an Inequality Relaxation Strategy}\label{sec:ipm}
+We present the condensed-space IPM within the
+context of the general NLP formulation in \eqref{eqn:cpt}. Our method
+has two key differences from standard IPM
+implementations: (i) the use of inequality relaxation and (ii) the
+condensed treatment of the KKT system.
+
+\subsection{Inequality Relaxation}
+
+At the beginning of the algorithm, we apply inequality relaxation to replace the equality constraints in \eqref{eqn:cpt} with inequalities by introducing slack variables $s\in\mathbb{R}^{m}$:
+\begin{align}\label{eqn:relax}
+  g(x)- s = 0,\quad s^{\flat}\leq s\leq  s^\sharp,
+\end{align}
+where $s^\flat,s^\sharp\in\mathbb{R}^{m}$ are chosen to be close to zero.
+This relaxed problem can be stated as follows:
+\begin{align}\label{eqn:relaxed} 
+  \begin{aligned}[t]
+    \min_{\big[\substack{x^\flat\\s^\flat}\big]\leq \big[\substack{x\\s}\big] \leq \big[\substack{x^\sharp\\s^\sharp}\big]}\;
+    &  f(x)\\
+    \st\;
+    & g(x) - s = 0.
+  \end{aligned}
+\end{align}
+
+In our implementation, we set $s^{\flat},s^\sharp$ as
+$-\epsilon_{\text{tol}}\boldsymbol{1}$ and $+\epsilon_{\text{tol}}\boldsymbol{1}$, respectively,
+where $\epsilon_{\text{tol}}>0$ is a user-specified relative tolerance
+of the IPM. This type of relaxation is
+commonly used in practical IPM implementations; for example, in Ipopt,
+the solver relaxes the bounds and inequality constraints by
+$O(\epsilon_{\text{tol}})$ to prevent an empty interior of the
+feasible set (see \cite[Section 3.5]{wachter2006implementation}).
+For condensed-space IPM, we cannot maintain the same level of precision
+due to the increased condition number of the KKT system.
+We have found that setting $\epsilon_{\text{tol}}$ to be
+$\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$ ensures numerical
+stability while achieving satisfactory convergence behavior. Thus, our
+solver sets the tolerance to $10^{-4}$ by default when using condensed
+IPM.
+
+\subsection{Barrier Subproblem}
+
+The IPM replaces the equality and inequality-constrained problem in \eqref{eqn:relax} with an equality-constrained  barrier subproblem:
+\begin{subequations}\label{eqn:barrier}
+  \begin{align}
+    \min_{x}\;
+    &
+      \begin{aligned}[t]
+        &f(x) - \mu\boldsymbol{1}^\top \log (x- x^\flat) -\mu\boldsymbol{1}^\top \log (x^\sharp -x)\\
+        &- \mu\boldsymbol{1}^\top \log (s- s^\flat) -\mu\boldsymbol{1}^\top \log (s^\sharp -s)
+      \end{aligned}
+          \label{eqn:barrier-obj}\\
+    \st\;
+    &g(x) = 0. \label{eqn:barrier-con}
+  \end{align}
+\end{subequations}
+Here, $\mu>0$ is the barrier parameter. The smooth log-barrier
+function is employed to avoid handling inequalities in a combinatorial
+fashion (as in active set methods). A superlinear local convergence to
+the first-order stationary point can be achieved by repeatedly
+applying Newton's step to the KKT conditions of \eqref{eqn:barrier}
+with $\mu\searrow 0 $.
+
+\subsection{Newton's Step Computation}
+The Newton step direction is computed by solving a so-called KKT system; to explain this, we consider the first-order optimality conditions (KKT conditions) for the barrier subproblem in \eqref{eqn:barrier}:
+\begin{align}\label{eqn:first} 
+  \begin{aligned}[t]
+    \nabla_{x} f(x) - \nabla_{x}g(x)^\top y  - z_x^\flat  + z_x^\sharp = 0\;&\\
+    \begin{aligned}
+      - z_s^\flat  + z_s^\sharp  &= 0\\
+      Z^\flat_x (x-x^\flat) - \mu\boldsymbol{1} &= 0\\
+      Z^\flat_s (s-s^\flat) - \mu\boldsymbol{1}&= 0,
+    \end{aligned}
+    \quad
+    \begin{aligned}
+      g(x) &= 0\\
+      Z^\sharp_x (x^\sharp-x) - \mu\boldsymbol{1}&= 0\\
+      Z^\sharp_s (s^\sharp-s) - \mu\boldsymbol{1}&= 0,
+    \end{aligned}&
+  \end{aligned}
+\end{align}
+where $y\in\mathbb{R}^{m}$, $z_x^\flat,z_x^\sharp\in\mathbb{R}^{n}$,
+and $z_s^\flat,z_s^\sharp\in\mathbb{R}^{m}$ are Lagrange multipliers
+associated with the equality and bound constraints in
+\eqref{eqn:relaxed}. The Newton step for solving the nonlinear
+equations in \eqref{eqn:first} can be computed by solving the KKT system in \eqref{eqn:very-long-eqn}.
+\begin{figure*}[h!]
+  % \begin{strip}
+  % \hline
+  \begin{align}\label{eqn:very-long-eqn}
+    \underbrace{
+    \begin{bmatrix}
+      W^{(\ell)}  + \delta^{(\ell)}_w I & & A^{(\ell)\top}& -I & I &  \\
+      & \delta^{(\ell)}_w I & -I&&&-I & I\\
+      A^{(\ell)}& -I & -\delta^{(\ell)}_c I\\
+      Z_x^{(\ell)\flat}&&&X^{(\ell)}-X^\flat\\
+      -Z_x^{(\ell)\sharp}&&&&X^\sharp-X^{(\ell)}\\
+      &Z_s^{(\ell)\flat}&&&&S^{(\ell)}-S^\flat\\
+      &-Z_s^{(\ell)\sharp}&&&&&S^\sharp-S^{(\ell)}\\
+    \end{bmatrix}
+    }_{M_\text{full}}
+    \begin{bmatrix}
+      \Delta x \\
+      \Delta s \\
+      \Delta y \\
+      \Delta z_x^\flat \\
+      \Delta z_x^\sharp \\
+      \Delta z_s^\flat \\
+      \Delta z_s^\sharp \\
+    \end{bmatrix} =
+    \begin{bmatrix}
+      p^{(\ell)}_{x }\\
+      p^{(\ell)}_{s }\\
+      p^{(\ell)}_{y }\\
+      p^{(\ell)}_{z_x^\flat }\\
+      p^{(\ell)}_{z_x^\sharp }\\
+      p^{(\ell)}_{z_s^\flat }\\
+      p^{(\ell)}_{z_s^\sharp }\\
+    \end{bmatrix}
+  \end{align}
+% \hline
+% \end{strip}
+\end{figure*}
+Here, we recall the definitions of $W^{(\ell)}$ and $A^{(\ell)}$ from Section \ref{sec:numopt}, and 
+$p^{(\ell)}_x,\cdots p^{(\ell)}_{z_s^\sharp}$ are defined by the left-hand-sides of the equations in
+\eqref{eqn:first}. In sequel, we shall drop the superscript
+$(\cdot)^{(\ell)}$ for concise notation.
+
+Now, we observe that a significant portion of the system in
+\eqref{eqn:very-long-eqn} can be eliminated by exploiting the block
+structure, leading to an equivalent system stated in a smaller
+space. In particular, the lower-right $4\times 4$ block is always
+invertible since the IPM procedure ensures that the iterates stay in
+the strict interior of the feasible set. This allows for eliminating
+the lower-right 4x4 block, resulting in:
+\begin{align}\label{eqn:very-long-reduced}
+  &
+    \underbrace{
+    \begin{bmatrix}
+      W  + \Sigma_x + \delta_w I & & A^{\top} \\
+      & \Sigma_s + \delta_w I & -I\\
+      A& -I & -\delta_c I\\
+    \end{bmatrix}
+  }_{M_{\text{red}}}
+  \begin{bmatrix}
+    \Delta x \\
+    \Delta s \\
+    \Delta y \\
+  \end{bmatrix}=
+  \begin{bmatrix}
+    q_x \\
+    q_s\\
+    q_y\\
+  \end{bmatrix},
+\end{align}
+where the dependence on the evaluation point is suppressed for concise notation, and:
+\begin{align*}
+  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
+  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
+  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
+  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
+  q_y&:=p_y.
+\end{align*}
+The bound dual steps can be recovered as follows:
+\begin{align}\label{eqn:recover-1}
+  \begin{aligned}[t]
+    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
+    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
+    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
+    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
+  \end{aligned}
+\end{align}
+Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
+are always diagonal, so their computation is not expensive.
+Also, note that the reduced system in \eqref{eqn:very-long-reduced}
+corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
+in the original version of the algorithm, we did not introduce the
+slack variables, so it does not have the additional structure imposed
+by the slack variables.
+
+The key advantage of the inequality relaxation strategy is that it
+imposes additional structure to the reduced KKT system, allowing us to
+further reduce the dimension of the problem. In particular, the
+lower-right 2x2 block in \eqref{eqn:very-long-reduced} can be
+eliminated, which is a procedure called {\it condensation}; here, the invertibility of the lower-right block can be verified from the fact that $\delta_w,\delta_c\geq 0$ and $\Sigma_s\succ 0$.
+Through this,
+we obtain the following system, written in the  primal
+space only:
+\begin{align}\label{eqn:very-long-condensed}
+  (\underbrace{W + \delta_wI + \Sigma_x + A^{\top} D A}_{M_\text{cond}} ) \Delta x = q_x + A^\top (C q_y +  Dq_s ),
+\end{align}
+where:
+\begin{align*}
+  C := \left(\delta_c \Sigma_s + (1+\delta^{}_c\delta^{}_w) I\right)^{-1}, \;
+  D := \left(\Sigma_s + \delta^{}_w I\right)C,
+\end{align*}
+and the dual and slack step directions can be recovered by:
+\begin{align}
+  \Delta s &:= C \left(\delta_c q_s - (q_y + A\Delta x)\right)\nonumber\\
+  \Delta y &:= (\Sigma_s + \delta_w I) \Delta s -q_s.\label{eqn:recover-2}
+\end{align}
+Again, the matrices involved in the inversions above are always
+diagonal, so their computation is not expensive.
+
+Therefore, the only sparse matrix that needs to be factorized is the
+matrix in \eqref{eqn:very-long-condensed}, which is in
+$\mathbb{R}^{n\times n}$. In general NLPs, the condensation strategy
+can arbitrarily increase the density of the KKT system, but for AC OPF
+problems, the condensed KKT system is still sparse.
+This is because the maximum number of nonzeros per row in constraint
+Jacobian is bounded by the graph degree, which should be reasonably
+small in practice.
+
+The reason that the condensation strategy is particularly relevant for
+GPUs is that the matrix in \eqref{eqn:very-long-condensed} is positive
+definite upon the application of standard inertia correction
+method. Typically, to guarantee that the computed step direction is a
+descent direction, we need a condition that
+$\text{inertia}(M_\text{red}) = (n+5m,0,m)$. Here, inertia refers to
+the tuple of positive, zero, and negative eigenvalues. Accoridngly, we
+employ inertia correction methods to modify the augmented KKT system
+so that the desired conditions on the inertia is satisfied.
+
+By Sylvester's law, we have:
+\begin{align*}
+  &\text{inertia}(M_\text{red}) = (n+5m,0,m)\\
+  &\iff \text{inertia}(M_\text{cond}) = (n,0,0).
+\end{align*}
+Thus, any choice of $\delta_w,\delta_c>0$ that makes the condensed KKT
+system positive definite yields the the desired inertia condition.
+An important observation here is that the condensed KKT matrix
+with desired inertia condition is always positive definite.
+Thus, $M_{\text{cond}}$ can be factorized with fixed pivoting (e.g.,
+Cholesky factorization or LU refactorization), which is significantly
+more amenable to parallel implementation than indefinite LBL$^\top$
+factorization, which is standard in the state-of-the-art CPU
+algorithms but requires the use of pivoting.
+
+Although we call \eqref{eqn:very-long-condensed} a condensed KKT system,
+$M_\text{cond}$ is not necessarily a dense matrix. In fact, in the case of
+AC OPF problems, $M_\text{cond}$ is still highly sparse, as $W$
+and $A$ are graph-induced banded systems. Thus, exploiting sparsity is
+still necessary to enable scalable computations.
+
+Due to the fact that the solutions of NLPs often have multiple active
+constraints, and the presence of active constraints can arbitrarily
+increase the conditioning of the KKT System, the system in
+\eqref{eqn:very-long-condensed} can be severely
+ill-conditioned. As such, a single triangular solve may not provide a
+sufficiently accurate step direction. Accordingly, iterative
+refinement methods are employed to refine the solution by performing
+multiple triangular solves. Notably, iterative refinement is applied
+to the full KKT system, rather than the condensed system in
+\eqref{eqn:very-long-condensed}.
+
+\subsection{Line Search and IPM Iterations}
+
+The step size can be determined using the line search procedure.
+Although there are numerous alternative approaches, we follow the
+filter line search method implemented in the Ipopt solver. The line
+search procedure employed here determines the step size by performing
+backtracking line search until a trial point satisfying sufficient
+progress conditions are satisfied and acceptable by the
+filter. Furthermore, to enhance the convergence behavior, various
+additional strategies are implemented, such as the second-order
+correction, restoration phase, and automatic scaling. For the details
+of the implementation of the filter line search and various additional
+strategies, the readers are referred to
+\cite{wachter2006implementation}.
+
+The step size and diretion obtained above can be implemented as follows:
+\begin{align}
+  (x,s,y) &\leftarrow (x,s,y)+ \alpha (\Delta x, \Delta s, \Delta y),\label{eqn:iter}\\\nonumber
+  (z_x^\flat, z_x^\sharp, z_s^\flat, z_s^\sharp) &\leftarrow (z_x^\flat, z_x^\sharp, z_s^\flat, z_s^\sharp) + \alpha_z (\Delta z_x^\flat, \Delta z_x^\sharp, \Delta z_s^\flat, \Delta z_s^\sharp).
+\end{align}
+The iteration in \eqref{eqn:iter} is repeated until the convergence
+criterion is satisfied. The convergence criterion is defined as
+\begin{align}\label{eqn:criteria}
+\text{residual}(x^{(\ell)}, s^{(\ell)}, y^{(\ell)},
+z^{(\ell)\flat}_x, z^{(\ell)\sharp}_x, z^{(\ell)\flat}_s,
+z^{(\ell)\sharp}_s) <\epsilon_{\text{tol}},
+\end{align}
+where
+$\text{residual}(\cdot)$ is a scaled version of the residual to the
+first-order conditions in \eqref{eqn:first}. Finally, we summarize our
+condensed-space IPM in Algorithm \ref{alg:con-ipm}.
+
+\begin{algorithm}[t]
+  \caption{Condensed-Space IPM}
+  \label{alg:con-ipm}
+  \begin{algorithmic}[1]
+    \REQUIRE Primal-dual solution guesses $x,y, z^\flat, z^\sharp$, bounds $x^\flat,x^\sharp, s^\flat, s^\sharp$, callbacks $f(\cdot)$, $g(\cdot)$, $\nabla_x f(\cdot)$, $\nabla_x g(\cdot)$, $\nabla^2_{xx} \mathcal{L}(\cdot)$, and tolerance $\epsilon_{\text{tol}}$
+    \STATE Relax the equality constraints by \eqref{eqn:relax} and initialize the slack $s$ and the associated dual variables $z^\flat_s, z^\sharp_s$.
+    \WHILE{convergence criteria \eqref{eqn:criteria} not satisfied}
+    \STATE Solve the condensed KKT system \eqref{eqn:very-long-condensed} with $\delta_w=\delta_c=0$ to compute the primal step $\Delta x$ and recover the dual steps $\Delta y, \Delta z_x^\flat, \Delta z_x^\sharp, \Delta z_s^\flat, \Delta z_s^\sharp$ by \eqref{eqn:recover-1} and \eqref{eqn:recover-2}.
+    \STATE Determine the need for regularization and if necessary, recompute the step directions with proper choices of $\delta_w,\delta_c>0$.
+    \STATE Choose step sizes $\alpha,\alpha_z>0$ via line search.
+    \STATE Update the solution by \eqref{eqn:iter}.
+    \STATE Update filter and barrier parameter $\mu$.
+    \ENDWHILE
+    \RETURN The first-order stationary points $x^\star,y^\star, z^{\flat\star}, z^{\sharp\star}$
+  \end{algorithmic}
+\end{algorithm}
+
+\subsection{Notes on the Implementation}
+We have implemented the condensed-space IPM by adapting our code base
+in MadNLP.jl, which implements the IPM as in Ipopt in Julia Language.
+A key feature of MadNLP is that the IPM is implemented with
+high levels of abstractions, while the specific handling of the data
+structures within the KKT systems is carried out by data-type specific
+kernel functions. This design allows us to apply the mathematical
+operations equivalent to Ipopt to different data structures, such as
+{\tt SparseKKTSystem}, {\tt DenseKKTSystem}, {\tt
+DenseCondensedKKTSystem}, etc, whose data are stored either on host or
+device memory. For the implementation of the condensed-space IPM
+presented in this paper, we have added a new type of KKT system called
+{\tt SparseCondensedKKTSystem} and implemented additional kernels
+needed for handling the data structures specific to this KKT system
+type. This approach ensures that we are performing mathematically
+equivalent operations as in the mature, extensively tested existing
+code base. This also allows us to easily switch between different KKT
+system types, which is crucial for experimenting with various solvers
+and data structures, as well as for efficiently leveraging GPU
+acceleration when available. Furthermore, by maintaining this level of
+abstraction, the condensed-space IPM can be seamlessly integrated into
+the existing framework, making it easier to maintain and extend in the
+future.
+
+\section{Numerical Results}\label{sec:num}
+
+This section presents the numerical benchmark results, comparing our
+method against state-of-the-art methods on CPUs for solving standard
+AC OPF problems.
+
+\subsection{Methods}
+
+We compared four different configurations of NLP soultion frameworks:
+\begin{align}
+  \label{config-1}\tag{Config 1} \bullet\;&\text{MadNLP.jl + SIMDiff.jl + cuSOLVER (GPU)}\\
+  \label{config-2}\tag{Config 2} \bullet\;&\text{MadNLP.jl + SIMDiff.jl + Ma27 (CPU)}\\
+  \label{config-3}\tag{Config 3} \bullet\;&\text{Ipopt + AMPL + Ma27 (CPU)}\\
+  \label{config-4}\tag{Config 4} \bullet\;&\text{Ipopt + JuMP.jl + Ma27 (CPU)}.
+\end{align}
+
+\ref{config-1} is our main GPU configuration; \ref{config-2}
+represents our implementation running on CPU; and \ref{config-3} and
+\ref{config-4} are used as benchmarks. \ref{config-1} and
+\ref{config-2} share significant amount of code, especially the
+high-level abstractions, but they differ in how they handle the KKT
+systems. In \ref{config-1}, MadNLP.jl applies the condensed-space IPM
+along with the inequality relaxation strategy, while in
+\ref{config-2}, MadNLP.jl applies IPM based on the indefinite,
+non-condensed KKT system, as in \eqref{eqn:kkt-indefinite}. In
+\ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
+system. The initial symbolic factorization is performed using the KLU
+package \cite{davis2010algorithm}, and the subsequent numerical
+factorization and triangular solves are performed by cuSOLVER with the
+fixed pivot sequence from KLU.  Software and hardware details of each
+configuration are illustrated in Table \ref{tbl:settings}. The AC OPF
+problem is formulated using the model from the rosetta-opf project
+\cite{rosetta-opf}, and the test cases are obtained from the pglib-opf
+repository \cite{babaeinejadsarookolaee2019power}. The external
+packages are called from Julia Language, through thin wrapper
+packages, such as Ipopt.jl and AmplNLWriters.jl. A tolerance of
+$10^{-4}$ is set for MadNLP.jl and Ipopt solvers, with other solver
+options adjusted to ensure a fair comparison across different
+solvers. The results can be reproduced with the script available at
+\url{https://github.com/sshin23/opf-on-gpu}.
+
+\subsection{Results}
+
+The numerical benchmark results, including total solution time and its
+breakdown into linear algebra and derivative evaluation time (with the
+remainder considered as solver internal time), are shown in Table
+\ref{tbl:results}. The quality of the solution (objective value and
+constraint violation measured by $\|\cdot\|_\infty$) is shown in Table
+\ref{tbl:quality}. Figure \ref{fig:speedup} visually represents the
+speedup brought by GPUs, by comparing the timing results of
+\ref{config-1} and \ref{config-2}.
+
+\begin{table*}[t]
+  \scriptsize
+  \centering
+  \caption{Details of Numerical Experiment Settings}
+  \begin{tabular}{|l|c|c|c|c|} 
+    \hline
+    & {\textbf{MadNLP.jl + SIMDiff.jl + cuSOLVER}} 
+    & {\textbf{MadNLP.jl + SIMDiff.jl + Ma27}} 
+    & {\textbf{Ipopt + AMPL + Ma27}}
+    & {\textbf{Ipopt + JuMP.jl + Ma27}}\\
+    &\textbf{(gpu)} &\textbf{(cpu)} &\textbf{(cpu)}& \textbf{(cpu)}\\
+    \hline
+    \textbf{Optimization Solver} & \multicolumn{2}{c|}{MadNLP.jl (dev)$^*$} & \multicolumn{2}{c|}{Ipopt (v3.13.3)} \\
+    \hline
+    \textbf{Derivative Evaluations} & \multicolumn{2}{c|}{SIMDiff.jl (dev)$^*$} &  AMPL Solver Library & JuMP.jl (v1.12.0)\\
+    \hline
+    \textbf{Linear Solver} &  cuSOLVER (v11.4.5) &\multicolumn{3}{c|}{Ma27 (v2015.06.23)}\\
+    \hline
+    \textbf{Hardware} & NVIDIA Quadro GV100 & \multicolumn{3}{c|}{Intel Xeon Gold 6140}\\ 
+    \hline
+  \end{tabular}\\
+  $^*$Specific commit hashes are available at \url{https://github.com/sshin23/opf-on-gpu}
+  \label{tbl:settings} 
+\end{table*}
+\begin{table*}[t]
+  \scriptsize 
+  \centering
+  \caption{Numerical Results}
+  \input{table.tex}\\
+  $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.
+  \label{tbl:results}
+\end{table*}
+\begin{table*}[t]
+  \scriptsize
+  \centering
+  \caption{Solution Quality}
+  \label{tbl:quality}
+  \input{table-2.tex}\\
+\end{table*}
+\begin{figure}[t]
+  \includegraphics[width=.45\textwidth]{speedup-sol.pdf}
+  \caption{Speedup achieved by using GPUs.}
+  \label{fig:speedup}
+\end{figure}
+
+First, when comparing the solvers' performance in terms of the
+interior point method (IPM) iteration counts, MadNLP.jl is as
+efficient as the state-of-the-art solver Ipopt. The IPM iteration
+count is nearly the same as that of Ipopt for achieving the same level
+of accuracy in the final solution (see Table \ref{tbl:quality}). This
+suggests that running mathematically equivalent operations on GPUs (by
+using shared code base in high-level abstractions) can yield a similar
+degree of effectiveness in terms of IPM convergence.
+
+Next, we discuss the effectiveness of parallel AD on GPUs. We observe
+that even on CPUs, SIMDiff.jl is substantially faster than the AD
+routines implemented in AMPL or JuMP.jl. This efficiency arises from
+SIMDiff.jl obtaining the derivative functions specifcially compiled
+for the type of model, which is optimized for the distinctive
+computational pattern in the model. When comparing SIMDiff.jl running
+on CPUs and GPUs, we observe a further speedup of up to almost 20x for
+large instances (e.g., case24465\_goc); remarkably, this is 50 times
+faster than JuMP.jl. This demonstrates that parallelization of AD
+brings significant computational gain.
+This is because most operations in the derivative
+evaluations are \ref{eqn:pattern-1}
+operations, which are highly effective on GPUs.
+
+We next discuss the linear solver time.  While the speedup achieved by
+linear solvers is only moderate, this has a high impact on the overall
+speedup, as linear solver time constitutes a significant portion of
+the total solution time. For large instances, approximately 4x speedup
+can be achieved, though this can vary depending on the instances. For
+example, for case24464\_goc, the GPU linear solver was slower than the
+CPUs. The investigation of under which circumstances cuSOLVERRF is
+more effective warrants further research.
+
+Solver internal time could also be significantly accelerated through
+GPU utilization. We can observe that the speedup in solver internal
+operations is consistently greater than the speedup in linear
+solvers. However, due to the frequent use of \ref{eqn:pattern-2}
+operations, the speedup in solver internal operations is lesser than
+that of derivative evaluations.
+
+Overall, our GPU implementation exhibits significant speedup across
+all components: derivative evaluation, linear algebra, and solver
+internal computation, resulting in substantial gains in total solution
+time. The results indicate that GPUs become more effective for
+large-scale instances, particularly when the number of variables is
+greater than 20,000. Notably, for the largest instance,
+case30000\_goc, our GPU implementation is 4 times faster than our CPU
+implementation and approximately 10 times faster than the solver
+configuration with state-of-the-art tools (Ipopt, JuMP.jl, and
+Ma27). This demonstrates that GPUs can bring significant computational
+gains for large-scale AC OPF problems, enabling the solution of
+previously inconceivable problems due to the limitations of existing
+CPU-based solution tools.
+
+\section{Conclusions and Future Outlook}\label{sec:conc}
+We have presented an NLP solution framework for solving large-scale AC
+OPF problems. By leveraging the SIMD abstraction of NLPs and a
+condensed-space IPM, we have effectively eliminated the need for
+serial computations, enabling the implementation of a solution
+framework that can run entirely on GPUs. Our method has demonstrated
+promising results, achieving a 4x speedup when compared to CPU
+implementations for large-scale AC OPF problems. Notably, our approach
+outperforms one of the state-of-the-art CPU-based implementations by
+approximately 10 times. These results, along with our packages
+MadNLP.jl and SIMDiff.jl, showcase a significant advancement in our
+capabilities in dealing with large-scale optimization problems in
+power systems and underscore the potential of accelerated computing in
+large-scale optimization area. However, the condensation procedure
+leads to an increase in the condition number of the KKT system,
+resulting in decreased final solution accuracy. Addressing the
+challenges posed by ill-conditioning remains to be an important aspect
+for future work. In the following paragraphs, we discuss some
+remaining open questions and future outlooks.
+
+\paragraph*{Obtaining Higher Numerical Precisions}
+While we have focused on the IPM, other constrained optimization
+paradigms, such as penalty methods and augmented Lagrangian methods
+exist, and similar strategies based on condensed linear systems can be
+developed. It would be valuable to investigate which algorithm would
+be the right paradigm for constrained large-scale optimization on GPU
+that can best handle the ill-conditioning issue of the condensed KKT
+system, and in turn, achieve the highest degree of accuracy.
+
+\paragraph*{Security-Constrained, Multi-Period, Distribution OPFs}
+Though the proposed method has demonstrated significant computational
+advantages for transmission AC OPF problems, our results can also
+interpreted that efficient CPUs can still handle these problems
+reasonably well. We anticipate that there will be more substantial
+performance gains for larger-scale optimization problems, such as
+security-constrained and multi-period OPFs or joint optimization
+problems involving transmission, distribution, and gas network
+systems. We are interested in exploring Schur complement-based
+decomposition approaches, combined with the condensation-based
+strategy, similarly to \cite{pacaud2023parallel}, to demonstrate even
+greater scalability.
+
+\paragraph*{Alternative Linear Solvers}
+While cuSOLVERRF has been effective for solving the condensed KKT
+systems using LU factorization, Cholesky factorization holds promise
+for better performance due to lower computational complexity and the
+ability to reveal the inertia of the KKT system. In the future, we are
+interested in exploring other linear solvers options, such as CHOLMOD
+\cite{chen2008algorithm}, Baspacho \cite{pineda2022theseus}, and HyKKT
+\cite{regev2023hykkt}.
+
+\paragraph*{Portability}
+Our implementation is currently only tested on NVIDIA GPUs, but our
+GPU implementation is largely based on array programming and
+KernelAbstractions.jl in Julia, which are in principle  compatible with
+various GPU architectures, including AMD, Intel, and Apple GPUs. By
+incorporating cross-architecture linear solvers, we envision
+supporting broader class of GPU accelerators in the future.
+
+\bibliographystyle{IEEEtran}
+\bibliography{main}
+\vspace{0.1cm}
+\begin{flushright}
+	\scriptsize \framebox{\parbox{2.5in}{Government License: The
+			submitted manuscript has been created by UChicago Argonne,
+			LLC, Operator of Argonne National Laboratory (``Argonne").
+			Argonne, a U.S. Department of Energy Office of Science
+			laboratory, is operated under Contract
+			No. DE-AC02-06CH11357.  The U.S. Government retains for
+			itself, and others acting on its behalf, a paid-up
+			nonexclusive, irrevocable worldwide license in said
+			article to reproduce, prepare derivative works, distribute
+			copies to the public, and perform publicly and display
+			publicly, by or on behalf of the Government.
+The Department of Energy will provide public access to these results of federally sponsored research in accordance with the DOE Public Access Plan.
+http://energy.gov/downloads/doe-public-access-plan.
+}}
+	\normalsize
+\end{flushright}
 

### 2023-07-27 paper/main.tex
finished sec 1

@@
 offers the key advantage of being able to factorize the KKT matrix
-without numerical pivoting, which has hampered the parallelization of
+without numerical pivoting, which in the past has hampered the parallelization of
 the algorithm.  By combining these two strategies, we can perform the
@@
 general AD implementations on CPUs (such as AMPL and JuMP.jl), our
-GPU-based method can be approximately 50 times faster.
+GPU-based  differentiation method can be approximately 50 times faster.
 
@@
 of linear solvers with a fixed numerical pivot sequence (known as
-refactorization), and an efficient implementation is available in the CUDA library.
+refactorization), an efficient implementation of which is available in the CUDA library.
 Although this method is susceptible to numerical stability issues due to
@@
 with JuMP.jl) by an order of magnitude.  This finding underscores the
-importance of harnessing the computational power of GPUs for tackling
-the computatoinal challenges in power systems.
+importance of harnessing the power of GPUs for tackling
+the computational challenges in power systems.
 

### 2023-07-27 paper/main.tex
finished section 3

@@
 into active-set methods and IPMs \cite{nocedal2006numerical}.
-Active-setmethods aim to find the set of active consraints associated
+Active-set methods aim to find the set of active constraints associated
 with the optimal solution in a combinatorial manner, while IPMs
@@
 $\mathbb{R}\cup\{+\infty,-\infty\}$, such as addition, multiplication,
-maximum, and minimum; and the notation $\chi_{s_k}(x):=\chi(x,s_k)$ is used, and
-assume that $\circ$ is commutative on $\{\chi_{s_k}(\cdot)\}_{\forall
+maximum, and minimum. In \ref{eqn:pattern-3}, we denote $\chi_{s_k}(x):=\chi(x,s_k)$ and
+assume that $\circ$ is commutative for $\{\chi_{s_k}(\cdot)\}_{\forall
 s_k}$. \ref{eqn:pattern-1} is typically most effective on GPUs, where
@@
 simultaneously manipulate the same device memory location. While
-\ref{eqn:pattern-2} and \ref{eqn:pattern-3} is less effective as they
-require such a manipulation, but they still can be significantly
+\ref{eqn:pattern-2} and \ref{eqn:pattern-3} are less effective as they
+require such manipulations, they still can be significantly
 faster than operations in CPUs, as substantial part of the operation
@@
 of the AD and optimization capabilities in our
-tools, MadNLP.jl and SIMDiff.jl, are implemented in Julia Language,
+tools, MadNLP.jl and SIMDiff.jl, are implemented in the Julia Language,
 by leveraging its kernel and array programming capabilities.
@@
 $h^{(n)}(\cdot,\cdot)$ are twice differentiable functions with respect
-to the first argument, $\{\{p^{(k)}_i\}_{i\in [N_k]}\}_{k\in[K]}$,
+to the first argument, whereas $\{\{p^{(k)}_i\}_{i\in [N_k]}\}_{k\in[K]}$,
 $\{\{q^{(k)}_{i}\}_{i\in [M_l]}\}_{m\in[M]}$, and
@@
 that our functions $f^{(l)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
-$h^{(n)}(\cdot,\cdot)$ can be expressed in the form of computational
+$h^{(n)}(\cdot,\cdot)$ can be expressed with computational
 graphs of moderate length. One can observe that the problem in
@@
 \subsection{Parallel AD}
-Many of the physics-based models, such as AC OPF, have highly repetitive
+Many of the physics-based models, such as AC OPF, have a highly repetitive
 structure. One of the manifestations of it is that the mathematical
@@
 over the data over which the instruction is executed. Specifically,
-the sparisty analysis code exploits Julia's multiple dispatch feature
-to obtain parameterized sparsity pattern for each instruction, and the
+the sparsity analysis code exploits Julia's multiple dispatch feature
+to obtain a parameterized sparsity pattern for each instruction, and the
 obtained parameterized sparsity pattern is materialized once the data
-array is given. This process drastically save the computational cost
+array is given. This process drastically saves in computational cost
 for the sparsity analysis.

### 2023-07-28 paper/main.tex
first pass

@@
 While derivative evaluation can be generally cheaper than linear
-algebra operations, our numerical results show that AD often
-constitutes more than half of the total solver time when using
-off-the-shelf AD implementations like JuMP.jl or AMPL. Instead of the
-general-purpose tools, our method leverages a specialized AD
+algebra operations, our numerical results on AC OPF problems
+show that AD often constitutes more than half of the total solver time when using
+off-the-shelf AD implementations like JuMP.jl~\cite{dunning2017jump} or AMPL~\cite{fourer1990modeling}. Instead,
+our method leverages a specialized AD
 implementation based on the SIMD abstraction of NLPs. This abstraction
-alllows to preserve the parallelizable structure within the model
+allows to preserve the parallelizable structure within the model
 equations, facilitating efficient derivative evaluations on the
@@
 SIMDiff.jl. The solution of the KKT system is performed using the
-external cuSOLVER library. To assess the performance of our method, we
+external {\tt cuSOLVER} library. To assess the performance of our method, we
 compare it against standard CPU approaches using the data available in
@@
 accelerators called HiOP has been under development
-\cite{hiop_techrep}, which has a similar scope as our solver
+\cite{hiop_techrep}, with a similar scope as our solver
 MadNLP.jl.
@@
 $[v_i]_{i\in[M]}:=[v_1;v_2;\cdots,v_M]$.  A vector of ones with an
-appropriate size is denoted by $\boldsymbol{1}$. An identiy matrix
+appropriate size is denoted by $\boldsymbol{1}$. An identity matrix
 with an appropriate size is denoted by $I$. For matrices $A$ and $B$,
 $A\succ(\succeq) B$ indicates that $A-B$ is positive (semi)-definite
-while $A>(\geq) B$ denotes a componentwise inequality. We use the
+while $A>(\geq) B$ denotes a component-wise inequality. We use the
 following convention for any symbol $x$: $X:=\diag(x)$.
@@
 We consider NLPs of the following form:
-\begin{subequations}\label{eqn:cpt}
-  \begin{align}
-    \min_{x^\flat \leq x \leq x^\sharp}\;
-    & f(x)\\
-    \st\;
-    & g(x) =0. \label{eqn:cpt-con}
-  \end{align} 
-\end{subequations}
+\begin{equation}\label{eqn:cpt}
+    \min_{x^\flat \leq x \leq x^\sharp}\;  f(x) \quad \st\;
+     g(x) =0 \, .
+\end{equation}
 Numerous solution algorithms have been developed in the NLP
@@
 constraints, the NLP solution algorithms can be broadly classified
-into active-set methods and IPMs \cite{nocedal2006numerical}.
+into active-set methods and interior-point methods (IPMs) \cite{nocedal2006numerical}.
 Active-set methods aim to find the set of active constraints associated
@@
 In terms of practical computations, three key components play
-vital roles: derivative evalutions (often provided by the AD
+vital roles: derivative evaluations (often provided by the AD
 capabilities of the algebraic modeling languages), linear algebra
@@
 reference for the IPM implementation. Below, we outline the overall
-computational procedure employed wihtin the NLP solution frameworks:
+computational procedure employed within the NLP solution frameworks:
 
@@
 
-\subsection{AD}
+\subsection{Automatic Differentiation}
 Numerical differentiation of computer programs can be achieved through
@@
 In contrast, AD differentiates computer
-programs by leveraging the computation graph. By inspecting the
-computation graph and applying chain rules, AD
-can efficiently and accurately evaluate derivatives. This approach has
-become the dominant paradigm for derivative computation within the
+programs by inspecting the computation graph and applying chain rules,
+to evaluate derivatives efficiently and accurately.
+This approach has become the dominant paradigm for derivative computation within the
 scientific computing domain, including NLP and
@@
 tools are often implemented as part of domain-specific modeling
-languages, such as JuMP, CasADi, and AMPL (optimization) and Torch and
+languages, such as JuMP, CasADi, and AMPL (optimization) and Tensorflow, Torch and
 Flux (machine learning).
@@
 recursive application of chain rules: (i) forward-mode and (ii)
-reverse-mode, which operate in opposite directions (from leaves to
-root and from root to leaves, respectively). Reverse-mode automatic
-differentiation, also known as an adjoint method, has proven to be
-particularly effective for dealing with the function expressions in
+reverse-mode, which operate in opposite directions (respectively, from leaves to
+root and from root to leaves). Reverse-mode automatic
+differentiation, also known as the adjoint method, has proven to be
+particularly effective for dealing with function expressions in
 large-scale optimization problems.
@@
 
-We highlight that the following, arguably quite common, computational
+We highlight that the following, arguably common, computational
 patterns are particularly effective when implemented on GPUs:
@@
 computational patterns mentioned above. For example, an AC OPF model can
-be implemented with 15 different computational patterns, most of which
+be implemented with 15 different computational patterns (see Section~\ref{sec:simd:ad}), most of which
 fall within the aforementioned categories. Furthermore, the
@@
 specify the model equations in an {\tt Iterable} data type in
-Julia Language. This composite data type consists of an instruction (a julia
+Julia Language. This composite data type consists of an instruction (a Julia
 function) and data (a host or device array) over which the instruction
@@
 \subsection{Parallel AD}
-Many of the physics-based models, such as AC OPF, have a highly repetitive
+\label{sec:simd:ad}
+Many physics-based models, such as AC OPF, have a highly repetitive
 structure. One of the manifestations of it is that the mathematical
@@
 AC OPF model. These patterns arise from (1) generation cost, (2) reference
-bus voltage angle constraint, (3-6) active and reactive power flow (to
-and from), (7) voltage angle difference constraint, (8-9) apparent
-power flow limits (to and from), (10-11) power balance equations,
+bus voltage angle constraint, (3-6) active and reactive power flow (from and to),
+(7) voltage angle difference constraint, (8-9) apparent
+power flow limits (from and to), (10-11) power balance equations,
 (12-13) generators' contributions to the power balance equations, and
@@
 The sparsity analysis is needed for determining the sparsity pattern
-of the evaluated derivatives. In the case of large-sclae sparse
+of the evaluated derivatives. In the case of large-scale sparse
 problems, the initial sparsity analysis of nonlinear expressions can
@@
 \end{align}
-where $s^\flat,s^\sharp\in\mathbb{R}^{m}$ are chosen to be close to zero.
+where $s^\flat,s^\sharp\in\mathbb{R}^{m}$ are lower and upper bounds chosen to be close to zero.
 This relaxed problem can be stated as follows:
-\begin{align}\label{eqn:relaxed} 
-  \begin{aligned}[t]
+\begin{equation}\label{eqn:relaxed}
     \min_{\big[\substack{x^\flat\\s^\flat}\big]\leq \big[\substack{x\\s}\big] \leq \big[\substack{x^\sharp\\s^\sharp}\big]}\;
-    &  f(x)\\
-    \st\;
-    & g(x) - s = 0.
-  \end{aligned}
-\end{align}
-
+      f(x)\quad \st\;  g(x) - s = 0 \, .
+\end{equation}
 In our implementation, we set $s^{\flat},s^\sharp$ as
@@
   \begin{align}
-    \min_{x}\;
+    \min_{x, s}\;
     &
@@
     \st\;
-    &g(x) = 0. \label{eqn:barrier-con}
+    &g(x) - s = 0. \label{eqn:barrier-con}
   \end{align}
@@
     \begin{aligned}
-      g(x) &= 0\\
+      g(x) - s  &= 0\\
       Z^\sharp_x (x^\sharp-x) - \mu\boldsymbol{1}&= 0\\
@@
 \eqref{eqn:relaxed}. The Newton step for solving the nonlinear
-equations in \eqref{eqn:first} can be computed by solving the KKT system in \eqref{eqn:very-long-eqn}.
+equations in \eqref{eqn:first} can be computed by solving the unreduced KKT system in \eqref{eqn:very-long-eqn}.
 \begin{figure*}[h!]
@@
 Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
-are always diagonal, so their computation is not expensive.
+are always diagonal, so their computation is cheap.
 Also, note that the reduced system in \eqref{eqn:very-long-reduced}
@@
 Again, the matrices involved in the inversions above are always
-diagonal, so their computation is not expensive.
+diagonal, so their computation is cheap.
 
 Therefore, the only sparse matrix that needs to be factorized is the
-matrix in \eqref{eqn:very-long-condensed}, which is in
-$\mathbb{R}^{n\times n}$. In general NLPs, the condensation strategy
+matrix in the left-hand-side of \eqref{eqn:very-long-condensed}, with
+dimension $n \times n$.
+In general NLPs, the condensation strategy
 can arbitrarily increase the density of the KKT system, but for AC OPF
-problems, the condensed KKT system is still sparse.
+problems, the condensed KKT system remains sparse.
 This is because the maximum number of nonzeros per row in constraint
@@
 $\text{inertia}(M_\text{red}) = (n+5m,0,m)$. Here, inertia refers to
-the tuple of positive, zero, and negative eigenvalues. Accoridngly, we
+the tuple of positive, zero, and negative eigenvalues. Accordingly, we
 employ inertia correction methods to modify the augmented KKT system
@@
 Thus, any choice of $\delta_w,\delta_c>0$ that makes the condensed KKT
-system positive definite yields the the desired inertia condition.
+system positive definite yields the desired inertia condition.
 An important observation here is that the condensed KKT matrix
@@
 more amenable to parallel implementation than indefinite LBL$^\top$
-factorization, which is standard in the state-of-the-art CPU
+factorization, which is standard in the state-of-the-art IPM
 algorithms but requires the use of pivoting.
@@
 
-The step size and diretion obtained above can be implemented as follows:
+The step size and direction obtained above can be implemented as follows:
 \begin{align}
@@
 We have implemented the condensed-space IPM by adapting our code base
-in MadNLP.jl, which implements the IPM as in Ipopt in Julia Language.
+in MadNLP.jl, a port of Ipopt in Julia Language.
 A key feature of MadNLP is that the IPM is implemented with
@@
 
-We compared four different configurations of NLP soultion frameworks:
+We compared four different configurations of NLP solution frameworks:
 \begin{align}
@@
 \ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
-system. The initial symbolic factorization is performed using the KLU
+system. The initial symbolic factorization is performed on the CPU using the KLU
 package \cite{davis2010algorithm}, and the subsequent numerical
@@
 routines implemented in AMPL or JuMP.jl. This efficiency arises from
-SIMDiff.jl obtaining the derivative functions specifcially compiled
+SIMDiff.jl obtaining the derivative functions specifically compiled
 for the type of model, which is optimized for the distinctive
@@
 evaluations are \ref{eqn:pattern-1}
-operations, which are highly effective on GPUs.
+operations, highly effective on GPUs.
 
@@
 
-\paragraph*{Obtaining Higher Numerical Precisions}
+\paragraph*{Obtaining Higher Numerical Precision}
 While we have focused on the IPM, other constrained optimization
@@
 
-\bibliographystyle{IEEEtran}
+\bibliographystyle{ieeetr}
 \bibliography{main}

### 2023-07-28 paper/main.tex
second pass

@@
   This paper introduces a novel computational framework for solving
-alternating current (AC) optimal power flow (OPF) problems using
+alternating current optimal power flow (ACOPF) problems using
 graphics processing units (GPUs). While GPUs have demonstrated
@@
 without numerical pivoting, which in the past has hampered the parallelization of
-the algorithm. By combining these two strategies, we can perform the
+the IPM algorithm. By combining these two strategies, we can perform the
 majority of operations on GPUs while keeping the data residing in the
@@
 
-While graphics processing units (GPUs) have demonstrated remarkable
-capabilities in various computing domains, their adoption in
-large-scale constrained nonlinear programming (NLP) regimes, such as
-alternating current (AC) optimal power flow (OPF) problems, has faced
-some limitations. One of the primary challenges arises from the
+The adoption of GPUs in the mathematical programming community has remained
+limited. Notably, nonlinear programming (NLP) remains dependent on
+algorithms developed in the 1990s offering limited room for parallelism.
+One of the primary challenges arises from the
 automatic differentiation (AD) of sparse model equations and the
@@
 \cite{anitescu2021targeting}. While GPU computation can trivially
-accelerate several parts of the optimization process, especially
-various internal computations within the optimization solver, the
+accelerate several parts of the optimization process --- especially
+various internal computations within the optimization solver --- the
 sluggish data transfer between host and device memory hampers the
-ad-hoc implementation of GPU accelerations (Fig. \ref{fig:simd}). To
-fully leverage the potential that modern GPU hardware has to offer, it
+ad-hoc implementation of GPU accelerations (Fig. \ref{fig:memory}). To
+fully leverage the potential offered by modern GPU hardware, it
 becomes imperative to have a comprehensive computational framework for
@@
 
+\subsection{Contributions}
 This paper presents our approach to implement a comprehensive
@@
 
+\paragraph*{SIMD abstraction of nonlinear programs}
 While derivative evaluation can be generally cheaper than linear
@@
 
+\paragraph*{Condensed-space interior-point method}
 Linear algebra operations, especially sparse indefinite matrix
@@
 We present numerical benchmark results to showcase the efficiency of
-our method, utilizing our implementations: MadNLP.jl and
+our method, utilizing our two packages: MadNLP.jl and
 SIMDiff.jl. The solution of the KKT system is performed using the
@@
 compare it against standard CPU approaches using the data available in
-pglib-opf \cite{wachter2006implementation}.  Our benchmark results
+pglib-opf \cite{babaeinejadsarookolaee2019power}.  Our benchmark results
 demonstrate that our proposed computational framework has significant
@@
 when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
-Specifically, when running on NVIDIA GPUs, our method achieves a
-notable 4x speedup compared to our solver running on CPU, in the case
-of the largest instance. Moreover for the same instance, our approach
+Notably, when running on NVIDIA GPUs, our method achieves a
+4x speedup compared to our solver running on CPU for
+the largest instance. Moreover for the same instance, our approach
 surpasses the performance of existing tools (such as Ipopt interfaced
@@
 
-\paragraph*{Related Work}
+\subsection{Related Work}
 Several recent works have explored the use of GPUs for large-scale
@@
 
-\paragraph*{Organization}
+\subsection{Organization}
 The paper is organized as follows. In the remainder of the current
@@
 Numerous solution algorithms have been developed in the NLP
-literature. In terms of strategies to deal with inequality
+literature to solve \eqref{eqn:cpt}. In terms of strategies to deal with inequality
 constraints, the NLP solution algorithms can be broadly classified
@@
 with the optimal solution in a combinatorial manner, while IPMs
-replace inequality constraints with smooth barrier functions, enabling
-the use of Newton-type second-order algorithms. IPMs are known to be
+replace inequality constraints with smooth barrier functions.
+IPMs are known to be
 more scalable for problems with a large number of constraints and
@@
 
-Since the successful implementation of the open-source IPM solver in
+Since the successful implementation of the open-source IPM solver
 Ipopt, many subsequent implementations of NLP solvers
@@
 1) Given the current primal-dual iterate $(x^{(\ell)},y^{(\ell)},
-z^{\flat(\ell)},z^{\sharp(\ell)})$, the AD package evaluates:
+z^{\flat(\ell)},z^{\sharp(\ell)})$, the AD package evaluates the first-
+and second-order derivatives:
 \begin{align*}
@@
 
-3) The optimization solver employs the filter line search procedure to
-determine the step size. The primal-dual iterate is updated by
+3) The optimization solver employs a filter line search procedure to
+determine the step size~\cite{wachter2006implementation}. The primal-dual iterate is updated by
 applying the determined step size and direction. This process is
-repeated until convergence criteria (typically based on the residual
-to the first-order optimality conditions) are satisfied.
-
-\subsection{Automatic Differentiation}
+repeated until satisfaction of the convergence criteria (typically based on the residual
+to the first-order optimality conditions).
+
+\subsection{Automatic Differentiation (AD)}
 Numerical differentiation of computer programs can be achieved through
@@
 In contrast, AD differentiates computer
-programs by inspecting the computation graph and applying chain rules,
+programs directly by inspecting the computation graph and applying chain rules,
 to evaluate derivatives efficiently and accurately.
@@
 scientific computing domain, including NLP and
-machine learning. For large-scale optimization, such as AC OPFs, AD
+machine learning. For large-scale optimization problems, such as AC OPFs, AD
 tools are often implemented as part of domain-specific modeling
@@
 efficient ways to implement automatic differentiation. Through the use
-of the multiple dispatch paradigm, \textit{any Julia function},
+of the multiple dispatch paradigm~\cite{bezanson2017julia} \textit{any Julia function} ---
 including commonly used operations like addition, multiplication,
-trigonometric and exponential functions, among others, can be easily
+trigonometric and exponential functions (among others) --- can be easily
 overloaded. Multiple dispatch allows functions to be dynamically
@@
 architectures to efficiently solve large-scale NLPs, like AC OPF problems.
-However, adapting the NLP solution algorithms,
-such as IPM, to GPUs presents challenges due to the fundamental
+However, adapting a NLP solution algorithms
+such as IPM to GPUs presents challenges due to the fundamental
 differences between GPU and CPU programming paradigms. While CPUs
@@
 lockstep execution across multiple threads, and in turn, prevent
-efficient implementations on GPUs. On the other hand, when the
+efficient implementations on GPUs. On the contrary, when the
 algorithm's structure allows for efficient parallelization, the SIMD
@@
 simultaneously manipulate the same device memory location. While
-\ref{eqn:pattern-2} and \ref{eqn:pattern-3} are less effective as they
-require such manipulations, they still can be significantly
+\ref{eqn:pattern-2} and \ref{eqn:pattern-3} are less effective,
+they still can be significantly
 faster than operations in CPUs, as substantial part of the operation
@@
 Implementing kernel functions for the above computational patterns in
-the Julia Language is straightforward, as it provides excellent
+the Julia Language is straightforward, as Julia provides excellent
 high-level interfaces for array and kernel programming for GPU
@@
 equations. However, such repetitive structure is not well exploited
-during AD in the standard NLP modeling paradigms.  In
+in the standard NLP modeling paradigms. In
 fact, without the SIMD abstraction, it is difficult for the AD package
@@
 be expensive, as the sparsity should be analyzed for potentially
-millions of objective terms and constraints. However, oftentimes
+millions of objective and constraints terms. However, oftentimes
 these analyses are applied for the same computational patterns, and
@@
 more amenable to parallel implementation than indefinite LBL$^\top$
-factorization, which is standard in the state-of-the-art IPM
-algorithms but requires the use of pivoting.
+factorization (standard in the state-of-the-art IPM
+algorithms but requires the use of pivoting).
 
@@
 
-Due to the fact that the solutions of NLPs often have multiple active
-constraints, and the presence of active constraints can arbitrarily
-increase the conditioning of the KKT System, the system in
-\eqref{eqn:very-long-condensed} can be severely
-ill-conditioned. As such, a single triangular solve may not provide a
+As we approach towards the solution, multiple constraints become active:
+in the diagonal matrices $\Sigma_x$ and $\Sigma_s$, the terms associated
+to the active (resp. inactive) variables $(x, s)$ go to infinity
+(resp. $0$).
+As such, the presence of active constraints can arbitrarily
+increase the conditioning of the KKT system, leading to
+an ill-conditioned KKT system in \eqref{eqn:very-long-condensed}.
+As a consequence, a single triangular solve may not provide a
 sufficiently accurate step direction. Accordingly, iterative
@@
 We have implemented the condensed-space IPM by adapting our code base
-in MadNLP.jl, a port of Ipopt in Julia Language.
-A key feature of MadNLP is that the IPM is implemented with
-high levels of abstractions, while the specific handling of the data
+in MadNLP.jl, a port of Ipopt in Julia.
+A key feature of MadNLP is that the IPM is implemented with a
+high levels of abstraction, while the specific handling of the data
 structures within the KKT systems is carried out by data-type specific
 kernel functions. This design allows us to apply the mathematical
-operations equivalent to Ipopt to different data structures, such as
+operations equivalent to Ipopt to different KKT data structures, such as
 {\tt SparseKKTSystem}, {\tt DenseKKTSystem}, {\tt
@@
 non-condensed KKT system, as in \eqref{eqn:kkt-indefinite}. In
-\ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
+\ref{config-1}, we use the {\tt cuSOLVER} library to solve the condensed KKT
 system. The initial symbolic factorization is performed on the CPU using the KLU
 package \cite{davis2010algorithm}, and the subsequent numerical
-factorization and triangular solves are performed by cuSOLVER with the
+factorization and triangular solves are performed by {\tt cuSOLVER} with the
 fixed pivot sequence from KLU.  Software and hardware details of each
@@
 repository \cite{babaeinejadsarookolaee2019power}. The external
-packages are called from Julia Language, through thin wrapper
+packages are called from Julia, through thin wrapper
 packages, such as Ipopt.jl and AmplNLWriters.jl. A tolerance of
@@
 
+\paragraph{Convergence pattern}
 First, when comparing the solvers' performance in terms of the
@@
 
+\paragraph{Performance of SIMD AD}
 Next, we discuss the effectiveness of parallel AD on GPUs. We observe
 that even on CPUs, SIMDiff.jl is substantially faster than the AD
-routines implemented in AMPL or JuMP.jl. This efficiency arises from
-SIMDiff.jl obtaining the derivative functions specifically compiled
-for the type of model, which is optimized for the distinctive
-computational pattern in the model. When comparing SIMDiff.jl running
-on CPUs and GPUs, we observe a further speedup of up to almost 20x for
+routines implemented in AMPL or JuMP.jl. Indeed,
+SIMDiff.jl generates derivative functions specifically compiled
+for the type of model, including optimization for the distinctive
+computational pattern found in the model. When comparing SIMDiff.jl running
+on CPUs and GPUs, we observe a further speedup on the GPU of up to almost 20x for
 large instances (e.g., case24465\_goc); remarkably, this is 50 times
-faster than JuMP.jl. This demonstrates that parallelization of AD
+faster than JuMP.jl. This demonstrates that parallelizing AD
 brings significant computational gain.
@@
 
+\paragraph{Performance of the condensed KKT strategy}
 We next discuss the linear solver time. While the speedup achieved by
@@
 example, for case24464\_goc, the GPU linear solver was slower than the
-CPUs. The investigation of under which circumstances cuSOLVERRF is
+CPUs. The investigation of under which circumstances {\tt cuSOLVERRF} is
 more effective warrants further research.
@@
 Overall, our GPU implementation exhibits significant speedup across
-all components: derivative evaluation, linear algebra, and solver
-internal computation, resulting in substantial gains in total solution
+all components (derivative evaluation, linear algebra, and solver
+internal computation) resulting in substantial gains in total solution
 time. The results indicate that GPUs become more effective for
@@
 case30000\_goc, our GPU implementation is 4 times faster than our CPU
-implementation and approximately 10 times faster than the solver
-configuration with state-of-the-art tools (Ipopt, JuMP.jl, and
-Ma27). This demonstrates that GPUs can bring significant computational
+implementation and approximately 10 times faster than
+state-of-the-art tools (Ipopt, JuMP.jl, and Ma27).
+This demonstrates that GPUs can bring significant computational
 gains for large-scale AC OPF problems, enabling the solution of
@@
 outperforms one of the state-of-the-art CPU-based implementations by
-approximately 10 times. These results, along with our packages
+a factor of 10. These results, along with our packages
 MadNLP.jl and SIMDiff.jl, showcase a significant advancement in our

### 2023-07-28 paper/main.tex
rename reduced KKT as augmented KKT

@@
     \end{bmatrix}
-  }_{M_{\text{red}}}
+  }_{M_{\text{aug}}}
   \begin{bmatrix}
@@
 are always diagonal, so their computation is cheap.
-Also, note that the reduced system in \eqref{eqn:very-long-reduced}
+Also, note that the augmented system in \eqref{eqn:very-long-reduced}
 corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
@@
 The key advantage of the inequality relaxation strategy is that it
-imposes additional structure to the reduced KKT system, allowing us to
+imposes additional structure to the augmented KKT system, allowing us to
 further reduce the dimension of the problem. In particular, the
@@
 descent direction, we need a condition that
-$\text{inertia}(M_\text{red}) = (n+5m,0,m)$. Here, inertia refers to
+$\text{inertia}(M_\text{aug}) = (n+5m,0,m)$. Here, inertia refers to
 the tuple of positive, zero, and negative eigenvalues. Accordingly, we
@@
 \begin{align*}
-  &\text{inertia}(M_\text{red}) = (n+5m,0,m)\\
+  &\text{inertia}(M_\text{aug}) = (n+5m,0,m)\\
   &\iff \text{inertia}(M_\text{cond}) = (n,0,0).
@@
 \paragraph*{Alternative Linear Solvers}
-While cuSOLVERRF has been effective for solving the condensed KKT
+While {\tt cuSOLVERRF} has been effective for solving the condensed KKT
 systems using LU factorization, Cholesky factorization holds promise

### 2023-07-28 paper/main.tex
further changes before mihai edit

--- 2023-07-28 paper/main.tex
@@
-\input{preamble}
+% \input{preamble}
 \title{Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods
@@
 
-\subsection{Contributions}
 This paper presents our approach to implement a comprehensive
@@
 
-\paragraph*{SIMD abstraction of nonlinear programs}
 While derivative evaluation can be generally cheaper than linear
-algebra operations, our numerical results on AC OPF problems
-show that AD often constitutes more than half of the total solver time when using
-off-the-shelf AD implementations like JuMP.jl~\cite{dunning2017jump} or AMPL~\cite{fourer1990modeling}. Instead,
-our method leverages a specialized AD
-implementation based on the SIMD abstraction of NLPs. This abstraction
-allows to preserve the parallelizable structure within the model
-equations, facilitating efficient derivative evaluations on the
-GPU. The AC power flow model is particularly well-suited for this
-abstraction as it involves repetitive expressions for each component
-type (e.g., buses, lines, generators), and the number of computational
-patterns does not increase with the network's size. Numerical results
-reported in this paper demonstrate that our proposed strategies can
-achieve over 20 times speedup by running on GPU. In comparison to
-general AD implementations on CPUs (such as AMPL and JuMP.jl), our
-GPU-based  differentiation method can be approximately 50 times faster.
-
-\paragraph*{Condensed-space interior-point method}
+algebra operations, our numerical results on AC OPF problems show that
+AD often constitutes more than half of the total solver time when
+using off-the-shelf AD implementations like
+JuMP.jl~\cite{dunning2017jump} or
+AMPL~\cite{fourer1990modeling}. Instead, our method leverages a
+specialized AD implementation based on the SIMD abstraction of
+NLPs. This abstraction allows to preserve the parallelizable structure
+within the model equations, facilitating efficient derivative
+evaluations on the GPU. The AC power flow model is particularly
+well-suited for this abstraction as it involves repetitive expressions
+for each component type (e.g., buses, lines, generators), and the
+number of computational patterns does not increase with the network's
+size. Numerical results reported in this paper demonstrate that our
+proposed strategies can achieve over 20 times speedup by running on
+GPU. In comparison to general AD implementations on CPUs (such as AMPL
+and JuMP.jl), our GPU-based  differentiation method can be approximately 50 times faster.
+
 Linear algebra operations, especially sparse indefinite matrix
@@
 
-\subsection{Related Work}
+\paragraph*{Contributions}
+We present, for the first time, a nonlinear optimization framework
+that can run entirely on GPU, with all the performance-critical data
+arrays residing exclusively on GPUs. Additionally, we introduce the
+concept of SIMD abstraction for NLP problems, which results in an
+efficient implementation of GPU-accelerated automatic differentiation,
+leading to a 50x speedup compared to existing tools for large-scale
+OPFs. Furthermore, we propose the condensed IPM with inequality
+relaxation strategy for the first time, enabling the treatment of the
+KKT systems of equality-constrained NLPs without numerical pivoting,
+thus allowing the solution of sparse, large-scale NLPs (with a
+prominent example being AC OPFs) on GPUs.
+
+\paragraph*{Related Work}
 Several recent works have explored the use of GPUs for large-scale
@@
 \cite{pacaud2023parallel,pacaud2022feasible,pacaud2023accelerating,cole2023exploiting}.
+In parallel, approaches based on Lagrangian decomposition and
+batched nonlinear optimzition solver based on TRON algorithm \cite{lin1999newton}
+has been investigated \cite{kim2022accelerated,kim2021leveraging}.
 An NLP solver for high-performance computers (HPC) with GPU
 accelerators called HiOP has been under development
-\cite{hiop_techrep}, with a similar scope as our solver
-MadNLP.jl.
-
-\subsection{Organization}
+\cite{hiop_techrep}, with a similar scope as our solver MadNLP.jl.
+The implementation of derivative evaluations with the exploitation of
+repeated structures within model equations has, to the best of our
+knowledge, been first introduced in Gravity \cite{Gravity}. This was
+achieved through the introduction of so-called template constraints,
+and multi-threaded derivate evaluation has been implemented therein.
+However, it is important to note that their differentiation approach
+is based on symbolic differentiation.
+
+\paragraph*{Organization}
 The paper is organized as follows. In the remainder of the current
@@
 \cite{rosetta-opf}, and the test cases are obtained from the pglib-opf
-repository \cite{babaeinejadsarookolaee2019power}. The external
-packages are called from Julia, through thin wrapper
+repository \cite{babaeinejadsarookolaee2019power}. We have selected
+the goc and pegase cases, as they contain large-scale instances.  The
+external packages are called from Julia, through thin wrapper
 packages, such as Ipopt.jl and AmplNLWriters.jl. A tolerance of

### 2023-07-28 paper/main.tex
changes

@@
 constraints to inequalities and condensing the Karush-Kuhn-Tucker
-system into a much smaller positive definite matrix. This strategy
+system into a much smaller positive definite system. This strategy
 offers the key advantage of being able to factorize the KKT matrix
@@
 concept of SIMD abstraction for NLP problems, which results in an
-efficient implementation of GPU-accelerated automatic differentiation,
-leading to a 50x speedup compared to existing tools for large-scale
+efficient implementation of GPU-accelerated parallel AD, leading to a
+50x speedup compared to existing tools for large-scale
 OPFs. Furthermore, we propose the condensed IPM with inequality
 relaxation strategy for the first time, enabling the treatment of the
-KKT systems of equality-constrained NLPs without numerical pivoting,
-thus allowing the solution of sparse, large-scale NLPs (with a
-prominent example being AC OPFs) on GPUs.
+KKT systems of NLPs without numerical pivoting, thus allowing the
+solution of sparse, large-scale NLPs (with a prominent example being
+AC OPFs) on GPUs.
 
@@
 augmented Lagrangian outer iteration and the treatment of linear
-systems using a preconditioned conjugate gradient method.  Another
-recent development is the hybrid (direct-iterative) KKT system solver
-specifically designed for GPUs \cite{regev2023hykkt}.  Before the
-introduction of the condensed space IPM with an inequality relaxation
+systems using a preconditioned conjugate gradient method. Prior to the
+introduction of the sparse condensed-space IPM with an inequality relaxation
 strategy, the authors have investigated the use of reduction
@@
 dense form on the GPU
-\cite{pacaud2023parallel,pacaud2022feasible,pacaud2023accelerating,cole2023exploiting}.
+\cite{pacaud2023parallel,pacaud2022feasible,pacaud2023accelerating}.
 In parallel, approaches based on Lagrangian decomposition and
-batched nonlinear optimzition solver based on TRON algorithm \cite{lin1999newton}
+batched with batched TRON solver \cite{lin1999newton}
 has been investigated \cite{kim2022accelerated,kim2021leveraging}.
@@
 \cite{hiop_techrep}, with a similar scope as our solver MadNLP.jl.
+ Another recent development is the hybrid (direct-iterative) KKT
+system solver specifically designed for GPUs \cite{regev2023hykkt}.
 The implementation of derivative evaluations with the exploitation of
@@
 is based on symbolic differentiation.
+The idea of condensed-space IPM (without inequlaity relaxation
+strategy) is not new, but they have been used primarily in more
+specific contexts, where the increased nonzero entries in the KKT system
+is less of a concern, such as in the dense form model predictive control
+problems \cite{jerez2012sparse,cole2023exploiting}.
 
@@
 \ref{sec:prelim}, we provide general preliminary knowledge on
-numerical optimization and GPU computing. In Section \ref{sec:simd},
-we present the SIMD abstraction of NLPs and their advantages in terms
-of implementing parallel AD. Section \ref{sec:ipm} presents the
-optimization algorithm under study, the condensed-space IPM with an
-inequality relaxation strategy. Section \ref{sec:num} presents the
-numerical results, comparing our approach with other state-of-the-art
-solution methods on the CPUs. Finally, conclusions and future outlooks
-are given in Section \ref{sec:conc}.
+numerical optimization, automatic differentiation, and GPU computing.
+In Section \ref{sec:simd}, we present the SIMD abstraction of NLPs and
+their advantages in terms of implementing parallel AD. Section
+\ref{sec:ipm} presents the optimization algorithm under study, the
+condensed-space IPM with an inequality relaxation strategy. Section
+\ref{sec:num} presents the numerical results, comparing our approach
+with other state-of-the-art solution methods on the CPUs. Finally,
+conclusions and future outlooks are given in Section \ref{sec:conc}.
 
@@
 while $A>(\geq) B$ denotes a component-wise inequality. We use the
-following convention for any symbol $x$: $X:=\diag(x)$.
+convention $X:=\diag(x)$ for any symbol $x$.
 
@@
 scientific computing domain, including NLP and
-machine learning. For large-scale optimization problems, such as AC OPFs, AD
-tools are often implemented as part of domain-specific modeling
-languages, such as JuMP, CasADi, and AMPL (optimization) and Tensorflow, Torch and
-Flux (machine learning).
+machine learning. For large-scale optimization problems, such as AC
+OPFs, AD tools are often implemented as part of domain-specific
+modeling languages. Examples of such modeling languages include JuMP,
+CasADi, and AMPL (optimization) and Tensorflow, Torch and, Flux
+(machine learning).
 
@@
 efficient ways to implement automatic differentiation. Through the use
-of the multiple dispatch paradigm~\cite{bezanson2017julia} \textit{any Julia function} ---
-including commonly used operations like addition, multiplication,
-trigonometric and exponential functions (among others) --- can be easily
-overloaded. Multiple dispatch allows functions to be dynamically
-dispatched based on the run-time type, a crucial feature for
-implementing differentiable programming. Several AD implementations
-have been developed in Julia Language, such as ReverseDiff.jl,
-ForwardDiff.jl, Zygote.jl, and JuMP.jl. While these tools are general
-and useful for various applications, they are not optimized for
-evaluating derivatives of AC OPF problems, as they are not designed to
-exploit the parallelizable structures in the model.
+of the multiple dispatch paradigm~\cite{bezanson2017julia} \textit{any
+Julia function}---including commonly used operations like addition,
+multiplication, trigonometric and exponential functions (among
+others)---can be easily overloaded. Multiple dispatch allows functions
+to be dynamically dispatched based on the run-time type, a crucial
+feature for implementing differentiable programming. Several AD
+implementations have been developed in Julia Language, such as
+ReverseDiff.jl, ForwardDiff.jl, Zygote.jl, and JuMP.jl. While these
+tools are general and useful for various applications, they are not
+optimized for evaluating derivatives of AC OPF problems, as they are
+not designed to exploit the parallelizable structures in the model,
+while preserving the desired sparsity.
 
@@
 architectures to efficiently solve large-scale NLPs, like AC OPF problems.
-However, adapting a NLP solution algorithms
-such as IPM to GPUs presents challenges due to the fundamental
+However, adapting a NLP solution algorithms,
+such as IPM, to GPUs presents challenges due to the fundamental
 differences between GPU and CPU programming paradigms. While CPUs
@@
 they still can be significantly
-faster than operations in CPUs, as substantial part of the operation
+faster than operations on CPUs, as substantial part of the operation
 can still be parallelized by the use of buffers. In simple cases, the
@@
 {\tt map} and {\tt mapreduce} programming models. However, in more
-complex cases, especially for \ref{eqn:pattern-3}, the
-implementation may require pre-inspection of memory write-access
-patterns and the use of custom kernels.
+complex cases, especially for \ref{eqn:pattern-3}, the implementation
+may require pre-inspection of memory write-access patterns and the use
+of custom kernels.
 
@@
 well as the application of optimization algorithms, are based on the
-computational patterns mentioned above. For example, an AC OPF model can
-be implemented with 15 different computational patterns (see Section~\ref{sec:simd:ad}), most of which
-fall within the aforementioned categories. Furthermore, the
-computation within optimization solvers, such as forming the
-left-hand-side for the KKT systems, computing the $\|\cdot\|_\infty$
-norm of the constraint violation, and assembling the condensed KKT
-system, can be carried out by using these computational patterns as well. The
-only exception is the factorization of the sparse KKT matrix, which
-requires more sophisticated implementations.
+computational patterns mentioned above. For example, an AC OPF model
+can be implemented with 15 different computational patterns (see
+Section~\ref{sec:simd:ad}), all of which fall within the
+aforementioned categories. Furthermore, the computation within
+optimization solvers, such as forming the left-hand-side for the KKT
+systems, computing the $\|\cdot\|_\infty$ norm of the constraint
+violation, and assembling the condensed KKT system, can be carried out
+by using these computational patterns as well. The only exception is
+the factorization of the sparse KKT matrix, which requires more
+sophisticated implementations.
 
@@
 \eqref{eqn:prob} is expressed by the computational patterns in Section
-\ref{sec:gpu}. In particular, the objective function falls within \ref{eqn:pattern-1},
-the first term in the constraint falls within \ref{eqn:pattern-2}, and the second term in the constraint falls within \ref{eqn:pattern-3}.
+\ref{sec:gpu}. In particular, the objective function falls within \ref{eqn:pattern-2},
+the first term in the constraint falls within \ref{eqn:pattern-1}, and the second term in the constraint falls within \ref{eqn:pattern-3}.
 Accordingly, the evaluation and differentiation of the
@@
 (14-15) in/out flows contributions to the power balance
-equations. However, such repetitive structure is not well exploited
-in the standard NLP modeling paradigms. In
-fact, without the SIMD abstraction, it is difficult for the AD package
-to detect the parallelizable structure within the model, as it will
-require the full inspection of the computational graph. By preserving
-the repetitive structures in the model, one can facilitate the
-implementation of parallelized derivative computations.
+equations. However, such repetitive structure is not well exploited in
+the standard NLP modeling paradigms. In fact, without the SIMD
+abstraction, it is difficult for the AD package to detect the
+parallelizable structure within the model, as it will require the full
+inspection of the computational graph over all expressions.  By
+preserving the repetitive structures in the model, the repetitive
+structure can be directly available in AD implementation.
 

### 2023-07-28 paper/main.tex
done

--- 2023-07-28 paper/main.tex
@@
 in AC OPF has been limited due to challenges associated with porting
-sparse automatic differentiation (AD) and sparse linear solver
+sparse automatic differentiation (AD) and spars e linear solver
 routines to GPUs. We aim to address these issues with two key
@@
 corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
-in the original version of the algorithm, we did not introduce the
-slack variables, so it does not have the additional structure imposed
+in the original version of the algorithm \cite{pacaud2023accelerating}, we did not introduce the
+slack variables, so it did not have the additional structure imposed
 by the slack variables.
@@
 A key feature of MadNLP is that the IPM is implemented with a
-high levels of abstraction, while the specific handling of the data
+high level of abstraction, while the specific handling of the data
 structures within the KKT systems is carried out by data-type specific

### 2023-07-28 paper/main.tex
before merging with Mihai's edit

--- 2023-07-28 paper/main.tex
@@
 GPU. In comparison to general AD implementations on CPUs (such as AMPL
-and JuMP.jl), our GPU-based  differentiation method can be approximately 50 times faster.
+and JuMP.jl), our GPU-based differentiation method can be
+approximately 500 times faster.
 
@@
 concept of SIMD abstraction for NLP problems, which results in an
-efficient implementation of GPU-accelerated parallel AD, leading to a
-50x speedup compared to existing tools for large-scale
-OPFs. Furthermore, we propose the condensed IPM with inequality
-relaxation strategy for the first time, enabling the treatment of the
-KKT systems of NLPs without numerical pivoting, thus allowing the
-solution of sparse, large-scale NLPs (with a prominent example being
-AC OPFs) on GPUs.
+efficient implementation of GPU-accelerated parallel AD. Furthermore,
+we propose the condensed IPM with inequality relaxation strategy for
+the first time, enabling the treatment of the KKT systems of NLPs
+without numerical pivoting, thus allowing the solution of sparse,
+large-scale NLPs (with a prominentexample being AC OPFs) on GPUs.
 
@@
 stability while achieving satisfactory convergence behavior. Thus, our
-solver sets the tolerance to $10^{-4}$ by default when using condensed
+solver sets the $\epsilon_\text{tol}$ to $10^{-4}$ by default when using condensed
 IPM.
@@
     \begin{aligned}
-      - z_s^\flat  + z_s^\sharp  &= 0\\
-      Z^\flat_x (x-x^\flat) - \mu\boldsymbol{1} &= 0\\
+      - z_s^\flat  + z_s^\sharp + y &= 0,\\
+      Z^\flat_x (x-x^\flat) - \mu\boldsymbol{1} &= 0,\\
       Z^\flat_s (s-s^\flat) - \mu\boldsymbol{1}&= 0,
@@
 \eqref{eqn:relaxed}. The Newton step for solving the nonlinear
-equations in \eqref{eqn:first} can be computed by solving the unreduced KKT system in \eqref{eqn:very-long-eqn}.
-\begin{figure*}[h!]
+equations in \eqref{eqn:first} can be computed by solving the system in Fig. \ref{fig:very-long-eqn}.
+\begin{figure*}[t]
   % \begin{strip}
   % \hline
-  \begin{align}\label{eqn:very-long-eqn}
+  \begin{align*}
     \underbrace{
@@
     \end{bmatrix}
-  \end{align}
-% \hline
-% \end{strip}
+  \end{align*}
+  \caption{Full KKT system}\label{fig:very-long-eqn} 
 \end{figure*}
@@
 
-Now, we observe that a significant portion of the system in
-\eqref{eqn:very-long-eqn} can be eliminated by exploiting the block
+Now, we observe that a significant portion of the system in Fig.
+\eqref{fig:very-long-eqn} can be eliminated by exploiting the block
 structure, leading to an equivalent system stated in a smaller
@@
 \end{align}
-where the dependence on the evaluation point is suppressed for concise notation, and:
+where:
 \begin{align*}
@@
 matrix in the left-hand-side of \eqref{eqn:very-long-condensed}, with
-dimension $n \times n$.
-In general NLPs, the condensation strategy
-can arbitrarily increase the density of the KKT system, but for AC OPF
-problems, the condensed KKT system remains sparse.
-This is because the maximum number of nonzeros per row in constraint
-Jacobian is bounded by the graph degree, which should be reasonably
-small in practice.
+dimension $n \times n$.  Although we call
+\eqref{eqn:very-long-condensed} a condensed KKT system,
+$M_\text{cond}$ is not necessarily a dense matrix. In fact, in the
+case of AC OPF problems, $M_\text{cond}$ is still highly sparse, as
+$W$ and $A$ are graph-induced banded systems.  Thus, exploiting
+sparsity is still necessary to enable scalable computations.  In
+general NLPs, however, the condensation strategy can arbitrarily
+increase the density of the KKT system. Thus, the condensed-space IPM
+strategy needs to be used with caution.
 
@@
 algorithms but requires the use of pivoting).
-
-Although we call \eqref{eqn:very-long-condensed} a condensed KKT system,
-$M_\text{cond}$ is not necessarily a dense matrix. In fact, in the case of
-AC OPF problems, $M_\text{cond}$ is still highly sparse, as $W$
-and $A$ are graph-induced banded systems. Thus, exploiting sparsity is
-still necessary to enable scalable computations.
 
@@
 constraint violation measured by $\|\cdot\|_\infty$) is shown in Table
-\ref{tbl:quality}. Figure \ref{fig:speedup} visually represents the
+\ref{tbl:quality}. Fig. \ref{fig:speedup} visually represents the
 speedup brought by GPUs, by comparing the timing results of
@@
 
-\paragraph{Convergence pattern}
+\paragraph*{Convergence pattern}
 First, when comparing the solvers' performance in terms of the
@@
 
-\paragraph{Performance of SIMD AD}
+\paragraph*{Performance of AD}
 Next, we discuss the effectiveness of parallel AD on GPUs. We observe
 that even on CPUs, SIMDiff.jl is substantially faster than the AD
-routines implemented in AMPL or JuMP.jl. Indeed,
-SIMDiff.jl generates derivative functions specifically compiled
-for the type of model, including optimization for the distinctive
-computational pattern found in the model. When comparing SIMDiff.jl running
-on CPUs and GPUs, we observe a further speedup on the GPU of up to almost 20x for
-large instances (e.g., case24465\_goc); remarkably, this is 50 times
-faster than JuMP.jl. This demonstrates that parallelizing AD
-brings significant computational gain.
-This is because most operations in the derivative
-evaluations are \ref{eqn:pattern-1}
-operations, highly effective on GPUs.
-
-\paragraph{Performance of the condensed KKT strategy}
+routines implemented in AMPL or JuMP.jl. Indeed, SIMDiff.jl generates
+derivative functions specifically compiled for the type of model,
+including optimization for the distinctive computational pattern found
+in the model. When comparing SIMDiff.jl running on CPUs and GPUs, we
+observe a further speedup on the GPU of up to almost 20x for large
+instances (e.g., case24465\_goc); remarkably, this is 500 times faster
+than JuMP.jl. This demonstrates that adopting SIMD abstraction and
+parallelizing AD brings significant computational gain.
+
+\paragraph*{Performance of the Condensed-Space IPM}
 We next discuss the linear solver time. While the speedup achieved by
@@
 operations is consistently greater than the speedup in linear
-solvers. However, due to the frequent use of \ref{eqn:pattern-2}
-operations, the speedup in solver internal operations is lesser than
-that of derivative evaluations.
+solvers. However, due to the frequent use of \ref{eqn:pattern-2} and
+\ref{eqn:pattern-3} operations, the speedup in solver internal
+operations is lesser than that of derivative evaluations.
 
@@
 gains for large-scale AC OPF problems, enabling the solution of
-previously inconceivable problems due to the limitations of existing
-CPU-based solution tools.
+previously inconceivable problems due to the limitations of CPU-based
+solution tools.
+
+\paragraph*{Known Limiataions}
+The key limitation of our method is the decreased precision of the
+final solution. Reliable convergence is achieved only up to a
+tolerance of $10^{-4}$. We have observed that the condensation of the
+KKT system worsens the conditioning of the already ill-conditioned
+augmented KKT system, resulting in even higher condition numbers,
+particularly when the solution is almost converged. For instance, in
+the case of case118\_ieee, the condition number of the KKT system at
+the solution (with tolerance set to $10^{-4}$) is $9.43\times 10^{11}$
+for the augmented KKT system and $3.15\times 10^{14}$ for the
+condensed KKT system. As a consequence, the achievable precision of
+the solution is reduced. Further investigation into the possibility of
+obtaining higher precision will be necessary in the future.
 

### 2023-07-28 paper/main.tex

@@
 are always diagonal, so their computation is cheap.
-Also, note that the reduced system in \eqref{eqn:very-long-reduced}
+Also, note that the augmented system in \eqref{eqn:very-long-reduced}
 corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
-in the original version of the algorithm, we did not introduce the
-slack variables, so it does not have the additional structure imposed
+in the original version of the algorithm \cite{pacaud2023accelerating}, we did not introduce the
+slack variables, so it did not have the additional structure imposed
 by the slack variables.

### 2023-07-28 paper/main.tex
math typo

@@
   q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
-  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
+  q_s&:=p_s + (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
   q_y&:=p_y.
@@
 \begin{align}\label{eqn:very-long-condensed}
-  (\underbrace{W + \delta_wI + \Sigma_x + A^{\top} D A}_{M_\text{cond}} ) \Delta x = q_x + A^\top (C q_y +  Dq_s ),
+  (\underbrace{W + \delta_wI + \Sigma_x + A^{\top} D A}_{M_\text{cond}} ) \Delta x = q_x + A^\top (C q_s +  Dq_y ),
 \end{align}

### 2023-07-28 paper/main.tex
ran grammarly

@@
 in AC OPF has been limited due to challenges associated with porting
-sparse automatic differentiation (AD) and spars e linear solver
+sparse automatic differentiation (AD) and sparse linear solver
 routines to GPUs. We aim to address these issues with two key
@@
 \begin{IEEEkeywords}
-  nonlinear programming, automatic differentiation, GPU computing, optimal power flow
+  nonlinear programming, automatic differentiation, GPU computing,
+optimal power flow
 \end{IEEEkeywords}
@@
 
-This paper presents our approach to implement a comprehensive
-computational framework for solving large-scale AC OPF problems on
+This paper presents our approach to implementing a comprehensive
+a computational framework for solving large-scale AC OPF problems on
 NVIDIA GPUs, along with the associated software implementations:
@@
 for each component type (e.g., buses, lines, generators), and the
-number of computational patterns does not increase with the network's
+the number of computational patterns does not increase with the network's
 size. Numerical results reported in this paper demonstrate that our
@@
 SIMDiff.jl. The solution of the KKT system is performed using the
-external {\tt cuSOLVER} library. To assess the performance of our method, we
+external cuSOLVER library. To assess the performance of our method, we
 compare it against standard CPU approaches using the data available in
@@
 4x speedup compared to our solver running on CPU for
-the largest instance. Moreover for the same instance, our approach
+the largest instance. Moreover, for the same instance, our approach
 surpasses the performance of existing tools (such as Ipopt interfaced
@@
 efficient implementation of GPU-accelerated parallel AD. Furthermore,
-we propose the condensed IPM with inequality relaxation strategy for
+we propose the condensed IPM with an inequality relaxation strategy for
 the first time, enabling the treatment of the KKT systems of NLPs
 without numerical pivoting, thus allowing the solution of sparse,
-large-scale NLPs (with a prominentexample being AC OPFs) on GPUs.
+large-scale NLPs (with a prominent example being AC OPFs) on GPUs.
 
@@
 
-\subsection{Numerical Optimization}\label{sec:numopt}
+This section covers three essential background topics: numerical
+optimization, AD, and GPU computing.\subsection{Numerical Optimization}\label{sec:numopt}
 We consider NLPs of the following form:
@@
 applying the determined step size and direction. This process is
-repeated until satisfaction of the convergence criteria (typically based on the residual
+repeated until the satisfaction of the convergence criteria (typically based on the residual
 to the first-order optimality conditions).
@@
 architectures to efficiently solve large-scale NLPs, like AC OPF problems.
-However, adapting a NLP solution algorithms,
+However, adapting an NLP solution algorithm,
 such as IPM, to GPUs presents challenges due to the fundamental
@@
 algorithm's structure allows for efficient parallelization, the SIMD
-parallelism in GPUs can offer orders of magnitudes speedup.
+parallelism in GPUs can offer orders of magnitude speedup.
 
@@
 highly efficient derivative computation code, specifically compiled
-for each computational pattern in the model. These derivative
-evaluation code can be run over the data in various GPU array formats,
-implemented via array and kernel programming in Julia Language. In
+for each computational pattern in the model. These derivative evaluation codes can be run over the data in various GPU array formats,
+and implemented via array and kernel programming in Julia Language. In
 turn, SIMDiff.jl has the capability to efficiently evaluate first and
-second order derivatives using GPU accelerators.
+second-order derivatives using GPU accelerators.
 
@@
 millions of objective and constraints terms. However, oftentimes
-these analyses are applied for the same computational patterns, and
-the time and memory spent for sparsity analysis can be significantly
+these analyses are applied to the same computational patterns, and
+the time and memory spent on sparsity analysis can be significantly
 reduced if the repetitive structures are exploited.
@@
 accomplished by applying sparsity analysis for the instruction for
-each computational pattern and expand the obtained sparsity pattern
+each computational pattern and expanding the obtained sparsity pattern
 over the data over which the instruction is executed. Specifically,
@@
 obtained parameterized sparsity pattern is materialized once the data
-array is given. This process drastically saves in computational cost
+array is given. This process drastically saves computational costs
 for the sparsity analysis.
@@
 $p^{(\ell)}_x,\cdots p^{(\ell)}_{z_s^\sharp}$ are defined by the left-hand-sides of the equations in
-\eqref{eqn:first}. In sequel, we shall drop the superscript
+\eqref{eqn:first}. In the sequel, we shall drop the superscript
 $(\cdot)^{(\ell)}$ for concise notation.
@@
 employ inertia correction methods to modify the augmented KKT system
-so that the desired conditions on the inertia is satisfied.
+so that the desired conditions on the inertia are satisfied.
 
@@
 
-As we approach towards the solution, multiple constraints become active:
+As we approach the solution, multiple constraints become active:
 in the diagonal matrices $\Sigma_x$ and $\Sigma_s$, the terms associated
-to the active (resp. inactive) variables $(x, s)$ go to infinity
+with the active (resp. inactive) variables $(x, s)$ go to infinity
 (resp. $0$).
@@
 search procedure employed here determines the step size by performing
-backtracking line search until a trial point satisfying sufficient
-progress conditions are satisfied and acceptable by the
+a backtracking line search until a trial point satisfying sufficient
+progress conditions is satisfied and acceptable by the
 filter. Furthermore, to enhance the convergence behavior, various
@@
 {\tt SparseKKTSystem}, {\tt DenseKKTSystem}, {\tt
-DenseCondensedKKTSystem}, etc, whose data are stored either on host or
-device memory. For the implementation of the condensed-space IPM
+DenseCondensedKKTSystem}, etc, whose data are stored either on the host or device memory. For the implementation of the condensed-space IPM
 presented in this paper, we have added a new type of KKT system called
@@
 non-condensed KKT system, as in \eqref{eqn:kkt-indefinite}. In
-\ref{config-1}, we use the {\tt cuSOLVER} library to solve the condensed KKT
+\ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
 system. The initial symbolic factorization is performed on the CPU using the KLU
 package \cite{davis2010algorithm}, and the subsequent numerical
-factorization and triangular solves are performed by {\tt cuSOLVER} with the
+factorization and triangular solves are performed by cuSOLVER with the
 fixed pivot sequence from KLU.  Software and hardware details of each
@@
 example, for case24464\_goc, the GPU linear solver was slower than the
-CPUs. The investigation of under which circumstances {\tt cuSOLVERRF} is
+CPUs. The investigation of under which circumstances cuSOLVER is
 more effective warrants further research.
@@
 challenges posed by ill-conditioning remains to be an important aspect
-for future work. In the following paragraphs, we discuss some
+of future work. In the following paragraphs, we discuss some
 remaining open questions and future outlooks.
@@
 \paragraph*{Alternative Linear Solvers}
-While {\tt cuSOLVERRF} has been effective for solving the condensed KKT
+While cuSOLVER has been effective for solving the condensed KKT
 systems using LU factorization, Cholesky factorization holds promise
@@
 incorporating cross-architecture linear solvers, we envision
-supporting broader class of GPU accelerators in the future.
+supporting a broader class of GPU accelerators in the future.
 

### 2023-08-22 coin-or/main.tex
coin-or synopsis

\documentclass{article}

\input{preamble}
\title{ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
\author{Sungho Shin$^\dag$, Francois Pacaud$^\ddag$, and Mihai Anitescu$^\dag$} 
\date{\small
  $^\dag$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^\ddag$Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France
}
\begin{document}
\maketitle

We are pleased to submit our nominations for the COIN-OR cup, highlighting our contributions of ExaModels and MadNLP in the domain of computational infrastructure for operations research, in the particular area of nonlinear optimization. Our open-source software packages represent a significant advance in addressing the challenges of efficiently solving large-scale nonlinear optimziation problems by harnessing the capabilities of modern GPU hardware. In this document, we highlight the key challenges and opportunities we are facing for the development of optimization software in 2023, especially in the context of the advent of accelerated computing. We also explain how ExaModels and MadNLP jointly address the key technical challanges and harness the benefits that modern accelerated computing has to offer. We also highlight several performance improvement we have achieved in terms of solving large-scale, real-world nonolinear optimziation problems. Finally, we conclude by discussing how our development sets the stage for the next-generation nonlinear optimization solver development and opens up new possibilities in various applications.

\paragraph{Challenges and Significance}
The computational landscape of today presents us with immense challenges and potential opportunities, especially with Graphics Processing Units (GPUs) at our disposal. Notably, NVIDIA GPUs have enabled great stride in the scalabiltiy of deep learning and achieved remarkable success in building large-scale AI models. Also, in the public doamin, most of the computational power for the next generation exascale HPC systems, such as Frontier and Aurora, come from GPU accelerators. While the GPUs have enabled great success in different scientific computing areas, such as machine learning and high-fidelity simulations, utilizing GPUs within nonlinear programming (NLP) algorithms has been hindered by challenges like sparse automatic differentiation (AD) and sparse linear solver routines, which do not seamlessly translate to GPU architecture. The key issue is that the conventional sparse AD and matrix factorization algoirthms are built for CPUs, and they heavily rely on the serial computation.

While GPU computation can trivially accelerate several parts of the optimization process---especially various internal computations within the optimization solver---the sluggish data transfer between host and device memory hampers the ad-hoc implementation of GPU accelerations (Fig. \ref{fig:memory}). To fully leverage the potential offered by modern GPU hardware, it becomes imperative to have a comprehensive computational framework for optimization on GPUs. That is, we need an AD/algebraic modeling framework, sparse linear solvers, and NLP solvers that can operate entirely on the GPU. Specifically, for the best performance, both the problem data and the solver's intermediate computational data must be exclusively resident within the device memory, with the majority of operations executed on the GPU. This necessitates the development of {\it a comprehensive nonlinear optimization solution framework}, which performs all the necessary computations---automatic differentiation, nonlinear optimization, and linear algebra---exclusively on GPUs. ExaModels and MadNLP rise to these challenges, carving out a pathway to fully harness the power of GPUs for operaitons research and nonlinear optimization.

\begin{figure}[t]
  \centering
  \scalebox{.87}{
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]

    \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
    \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
    \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};

    \node[align=center] at (1.25, 3.75) {Data};
    \node[align=center] at (6.75, 3.75) {Data};

    % Host Memory
    \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};

    % Device Memory
    \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};

    % Arrow with Dashed Line
    \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
    \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
    \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  }
  \includegraphics[width=.48\textwidth]{speedup-sol.pdf}
  \label{fig:simd}
  \caption{A schematic of host (CPU) and device (GPU) memory structure (left) and speedup achieved for AC optimal power flow problems (right).}\label{fig:memory}
\end{figure}

\paragraph{Developed Software Tools}
We have developed a comprehesnive nonlinear optimization framework by implementing our algebraic modeling/automatic differentiation tool ExaModels and nonlinear optimization solver MadNLP, while linear algebra computation is currently performed by the external cuSOLVER library.

{\it ExaModels}: SIMD Abstraction for NLPs: ExaModels pioneers the implementation of a groundbreaking single-instruction, multiple-data (SIMD) abstraction for NLPs. This revolutionary approach unlocks the potential for efficient parallel automatic differentiation on GPUs. The essence of this innovation lies in preserving the parallelizable structure within model equations, transforming derivative evaluations into streamlined operations. By enabling efficient computation of derivatives on GPUs, ExaModels offers an exceptional performance boost for power systems analysis.

The optimization models are most of the times implemented by the researchers and practitioners who do not necessarily understand the operations 
ExaModels provides a new platform for scalable nonlinear optimization. Our numerical results suggest that the conventional algebraic modeling systems are limited in terms of efficiently computing the derivates of the model equations (Table \ref{tab:num}), and it does not allow performing the operations on GPUs (Table \ref{tbl:portabiltiy}). ExaModels provide a framework that allows the user to conveniently inform the AD backend the parallelizable structure. Writing the models in this way will enable the GPU compatibility

{\it MadNLP}: Condensed-Space IPM with Inequality Relaxation: MadNLP introduces an equally transformative approach through its condensed-space interior-point method (IPM) with an inequality relaxation strategy. This strategy deftly overcomes the challenges posed by sparse matrix factorization on GPUs. The relaxation of equality constraints and condensation of the Karush-Kuhn-Tucker (KKT) system establishes positive definiteness, paving the way for the utilization of highly efficient linear solvers. MadNLP thus unveils a remarkable solution to traditionally intractable problems in GPU-based optimization.

While MadNLP has started as a port of Ipopt on Julia Language, now marking as the fourth year since its initial development, MadNLP has become a lot more versatile solver than the previously developed nonlinear optimization solvers. The key features include (1) being able to solve dense optimzation problems, (2) ability to handle different forms of KKT systems, (3) being able to handle diverse array data types (most notably, device arrays), and (4) exploit various Hessian approximation strategies (BFGS variants and limited-memory BFGS methods).

\paragraph{Performance Highlights}

The significance of these contributions is not merely theoretical. Empirical results speak volumes about the impact of ExaModels and MadNLP. When applied to AC OPF problems, the performance enhancements achieved are nothing short of extraordinary. The GPU-based solutions attain speedups of over 20 times compared to CPU-based counterparts, showcasing the profound potential of these software packages in power systems analysis. Their prowess surpasses even established tools interfaced with CPUs, marking a paradigm shift in computational efficiency.

\cite{shin2023accelerating}

\begin{table}[t]
  \begin{center}
    \include{table}
    \vspace{-1em}
    \footnotesize
    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  \caption{Numerical Performance of ExaModels and MadNLP}
  \end{center}
  \label{tab:num}
\end{table}

\paragraph{Portability}

While NVIDIA seems to be most capable in computing power and mature in terms of the implementation of the low-level operations (especially, the CUDA libraries), we aim to support all the other alternative architectures. This is because quite a few next-generaiton exascale high-performance computing architectures will be based on AMD and Intel accelerator devices. Also, as an open-source software, we aim to be the most generic software tool, not the most efficient one. By supporting diverse platforms, we hope to spark the attention of the chip makers in the operations research area and encourage better support for the software infrastructure for implmenting operations research software.

Table \ref{tbl:portabiltiy} summarizes the current state of the portability of our tools and other existing open-source software tools. Notably, ExaModels.jl already support all major architectures, whereas MadNLP.jl only support NVIDIA GPUs. Nonetheless, we have clear path forward in supporting all the major accelerator architectures, including NVIDIA, AMD, and Intel GPUs, while Apple GPUs are somewhat limited due to their inability to perform double precision operations. This is due to the fact that all of our GPU kernels are implemented via KernelAbstractions.jl, a julia package enabling portable implementation of GPU kernels. In this way, the GPU kernels for diverse accelerator architectures can be run based on a sam source code. Using this strategy, recently ExaModels.jl has demonstrated its capabiltiy to run on 4 different accelerator architectures (multi-thread CPUs and NVIDIA, AMD, and Intel GPUs). In the coming years, we plan to exand MadNLP's support for various accelerator architectures.

However, at the same time, it is absolutely crucial to have efficient and reliable, and preferrably also portable sparse linear sovler routines. Currently, we rely on CUSOLVER library for solving the linear systems. While vendor-implmented libraries are expected to have the best performance via high level of optimization, it would be beneficial for the community to have the open-source portable linear solvers. By showcasing the capability of solving high-stake nonlinear optimziation problems on GPUs, we can attract the attention of the numerical linear algebra community so that we can encourage the development of capable open-source sparse linear solvers. 
1
\begin{table}[t]
  \begin{center}
    \scalebox{0.72}{
      \begin{tabular}{|c|c|cccccc|}
        \hline
        &&CPU (single) &CPU (multi)& NVIDIA & AMD & Intel& Apple\\
        \hline
        \multirow{3}{*}{Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf JuMP}& \xmark& \cmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\ 
        \hline
        \multirow{2}{*}{Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{The current status on the portability of the nonlinear optimization frameworks.}
  \label{tbl:portabiltiy}
\end{table}

\paragraph{Impacts in Various Applications}

{\it Energy Infrastructures.} The resillience We have recently showcased the capability of MadNLP with multiple GPUs for solving extremely-large scale security-constrained AC OPF problems.

{\it 1}

{\it Machine Learning Surrogate Models.}

\paragraph{Closing Remarks}
ExaModels and MadNLP embody the spirit of innovation and progress that the COIN-OR cup seeks to recognize. By addressing critical challenges and introducing transformative strategies, these packages extend the frontiers of computational infrastructure for operations research. We are confident that their exceptional contributions will resonate within the operations research community and beyond, redefining the future of power systems analysis and optimization. It is with great enthusiasm that we nominate ExaModels and MadNLP for the COIN-OR cup, recognizing their exceptional impact on the field.

\bibliography{main}

\end{document}

### 2023-08-22 paper/main.tex
paper submitted versions

@@
-% \input{preamble}
+% \documentclass{IEEEtran4PSCC} % will be used for submission
+% \author{
+%   \IEEEauthorblockN{Sungho Shin and Mihai Anitescu}
+%   \IEEEauthorblockA{
+%     Mathematics and Computer Science Division\\
+%     Argonne National Laboratory\\
+%     Lemont, IL, USA\\
+%     sshin@anl.gov, anitescu@mcs.anl.gov}
+%   \and
+%   \IEEEauthorblockN{François Pacaud}
+%   \IEEEauthorblockA{Centre Automatique et Systèmes\\
+%     Mines Paris - PSL \\
+%     Paris, France\\
+%     francois.pacaud@minesparis.psl.eu}
+% }
+% \psccfooter{%
+%         \parbox{\textwidth}{\hrulefill \\ \small{23rd Power Systems Computation Conference} \hfill \begin{minipage}{0.2\textwidth}\centering \vspace*{4pt} \includegraphics[scale=0.06]{PSCC_logo.png}\\\small{PSCC 2024} \end{minipage} \hfill \small{Paris, France --- June 4 -- 7, 2024}}%
+% }
+
+\documentclass{IEEEtran} % for ArXiv submission
+\author{
+  Sungho Shin, François Pacaud, Mihai Anitescu
+  \thanks{
+    This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research (ASCR) under Contract DE-AC02-06CH11347.
+  }
+  \thanks{
+    Sungho Shin and Mihai Anitescu are with Mathematics and Computer Science Division,  Argonne National Laboratory, Lemont, IL, USA. e-mail: \url{sshin@anl.gov}, \url{anitescu@mcs.anl.gov}
+  }
+  \thanks{
+    François Pacaud is with Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France. e-mail: \url{francois.pacaud@minesparis.psl.eu}
+  }
+}
+
+\hyphenation{op-tical net-works semi-conduc-tor}
+
+% Set footer
+\makeatletter
+\let\old@ps@headings\ps@headings
+\let\old@ps@IEEEtitlepagestyle\ps@IEEEtitlepagestyle
+\def\psccfooter#1{%
+    \def\ps@headings{%
+        \old@ps@headings%
+        \def\@oddfoot{\strut\hfill#1\hfill\strut}%
+        \def\@evenfoot{\strut\hfill#1\hfill\strut}%
+    }%
+    \def\ps@IEEEtitlepagestyle{%
+        \old@ps@IEEEtitlepagestyle%
+        \def\@oddfoot{\strut\hfill#1\hfill\strut}%
+        \def\@evenfoot{\strut\hfill#1\hfill\strut}%
+    }%
+    \ps@headings%
+}
+\makeatother
+
+\usepackage{amsmath,amssymb,amsthm,amsfonts} % assumes amsmath package installed
+\usepackage[linktocpage=true,colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
+% \usepackage[letterpaper,margin=1in]{geometry}
+
+\newtheorem{definition}{Definition}
+\newtheorem{assumption}{Assumption}
+\newtheorem{theorem}{Theorem}
+\newtheorem{conjecture}{Conjecture}
+\newtheorem{lemma}{Lemma}
+\newtheorem{proposition}{Proposition}
+\newtheorem{remark}{Remark}
+
+\newcommand{\bx}{\boldsymbol{x}}
+\newcommand{\be}{\boldsymbol{e}}
+\newcommand{\blambda}{\boldsymbol{\lambda}}
+\newcommand{\bLambda}{\boldsymbol{\Lambda}}
+\newcommand{\bu}{\boldsymbol{u}}
+\newcommand{\bw}{\boldsymbol{w}}
+\newcommand{\by}{\boldsymbol{y}}
+\newcommand{\bz}{\boldsymbol{z}}
+\newcommand{\bV}{\boldsymbol{V}}
+\newcommand{\bX}{\boldsymbol{X}}
+\newcommand{\bY}{\boldsymbol{Y}}
+\newcommand{\bZ}{\boldsymbol{Z}}
+\newcommand{\bv}{\boldsymbol{v}}
+\newcommand{\bxi}{\boldsymbol{\xi}}
+\newcommand{\bpi}{\boldsymbol{\pi}}
+\newcommand{\bphi}{\boldsymbol{\phi}}
+\newcommand{\bbeta}{\boldsymbol{\eta}}
+\newcommand{\bpsi}{\boldsymbol{\psi}}
+\newcommand{\bzeta}{\boldsymbol{\zeta}}
+\newcommand{\bmu}{\boldsymbol{\mu}}
+\newcommand{\bq}{\boldsymbol{q}}
+\newcommand{\bQ}{\boldsymbol{Q}}
+\newcommand{\bK}{\boldsymbol{K}}
+\newcommand{\bP}{\boldsymbol{P}}
+\newcommand{\bS}{\boldsymbol{S}}
+\newcommand{\bT}{\boldsymbol{T}}
+\newcommand{\bF}{\boldsymbol{F}}
+\newcommand{\bG}{\boldsymbol{G}}
+\newcommand{\bd}{\boldsymbol{d}}
+\newcommand{\bp}{\boldsymbol{p}}
+\newcommand{\bff}{\boldsymbol{f}}
+\newcommand{\bc}{\boldsymbol{c}}
+\newcommand{\bg}{\boldsymbol{g}}
+\newcommand{\bh}{\boldsymbol{h}}
+\newcommand{\bA}{\boldsymbol{A}}
+\newcommand{\bL}{\boldsymbol{L}}
+\newcommand{\ba}{\boldsymbol{a}}
+\newcommand{\bb}{\boldsymbol{b}}
+\newcommand{\bB}{\boldsymbol{B}}
+\newcommand{\bC}{\boldsymbol{C}}
+\newcommand{\bE}{\boldsymbol{E}}
+\newcommand{\bH}{\boldsymbol{H}}
+\newcommand{\bR}{\boldsymbol{R}}
+\newcommand{\bn}{\boldsymbol{n}}
+\newcommand{\bm}{\boldsymbol{m}}
+\newcommand{\br}{\boldsymbol{r}}
+\newcommand{\bl}{\boldsymbol{l}}
+\newcommand{\bI}{\boldsymbol{I}}
+\newcommand{\osigma}{\overline{\sigma}}
+\newcommand{\usigma}{\underline{\sigma}}
+\newcommand{\oosigma}{\overline{\osigma}}
+\newcommand{\uusigma}{\underline{\usigma}}
+\newcommand{\olambda}{\overline{\lambda}}
+\newcommand{\ulambda}{\underline{\lambda}}
+\newcommand{\oolambda}{\overline{\olambda}}
+\newcommand{\uulambda}{\underline{\ulambda}}
+\newcommand{\bzero}{\boldsymbol{0}}
+\newcommand{\dist}{\text{\normalfont dist}}
+\newcommand{\st}{\mathop{\text{\normalfont s.t.}}}
+\newcommand{\diag}{\mathop{\text{\normalfont diag}}}
+\newcommand{\amin}{\mathop{\text{\normalfont argmin}}}
+\newcommand{\ReH}{\mathop{\text{\normalfont ReH}}}
+\newcommand{\bbZ}{\mathbb{Z}}
+\newcommand{\cG}{\mathcal{G}}
+\newcommand{\cV}{\mathcal{V}}
+\newcommand{\cW}{\mathcal{W}}
+\newcommand{\cA}{\mathcal{A}}
+\newcommand{\cB}{\mathcal{B}}
+\newcommand{\cL}{\mathcal{L}}
+\newcommand{\cE}{\mathcal{E}}
+\newcommand{\cD}{\mathcal{D}}
+\newcommand{\cP}{\mathcal{P}}
+\newcommand{\cQ}{\mathcal{Q}}
+\newcommand{\cK}{\mathcal{K}}
+\newcommand{\cM}{\mathcal{M}}
+\newcommand{\cN}{\mathcal{N}}
+\newcommand{\cI}{\mathcal{I}}
+\newcommand{\cJ}{\mathcal{J}}
+\newcommand{\cT}{\mathcal{T}}
+\newcommand{\interior}{\mathop{\text{\normalfont interior}}}
+\newcommand{\relint}{\mathop{\text{\normalfont relint}}}
+\newcommand{\vertices}{\mathop{\text{\normalfont vertices}}}
+\sloppy
+\usepackage{graphicx}
+
+\usepackage{enumitem}
+\usepackage{mathtools}
+\allowdisplaybreaks
+\renewcommand{\theenumi}{(\alph{enumi})}
+% \usepackage{parskip}
+
+\usepackage{tikz}
+\usepackage{lscape}
+\usepackage{multirow}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage{cuted}
+\setlength\stripsep{3pt plus 1pt minus 1pt}
+
+\usepackage{listings}
+\lstset{
+basicstyle=\ttfamily\footnotesize,
+columns=fullflexible, % make sure to use fixed-width font, CM typewriter is NOT fixed width
+numbers=left,
+numberstyle=\ttfamily,
+stepnumber=1,
+numbersep=-2pt,
+numberfirstline=true,
+numberblanklines=true,
+tabsize=4,
+lineskip=-1.5pt,
+extendedchars=true,
+breaklines=true,
+keywordstyle=\color{Blue}\bfseries,
+identifierstyle=, % using emph or index keywords
+commentstyle=\sffamily\color{OliveGreen},
+stringstyle=\color{Maroon},
+showstringspaces=false,
+showtabs=false,
+upquote=false,
+texcl=true % interpet comments as LaTeX
+}
+
+\lstdefinelanguage{Julia}%
+  {morekeywords={abstract,break,case,catch,const,continue,do,else,elseif,%
+      end,export,false,for,function,immutable,import,importall,if,in,%
+      macro,module,otherwise,quote,return,switch,true,try,type,typealias,%
+      using,while},%
+   sensitive=true,%
+   alsoother={$},%
+   morecomment=[l]\#,%
+   morecomment=[n]{\#=}{=\#},%
+   morestring=[s]{"}{"},%
+   morestring=[m]{'}{'},%
+}[keywords,comments,strings]%
+
+\lstset{%
+  language         = Julia,
+  basicstyle       = \ttfamily\small,
+  keywordstyle     = \bfseries\color{blue},
+  stringstyle      = \color{magenta},
+  commentstyle     = \color{ForestGreen},
+  showstringspaces = false,
+}
+
+\usepackage{cite}
+
 \title{Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods
@@
 for solving large-scale AC OPF problems to a moderate accuracy, our
-implementations—MadNLP.jl and SIMDiff.jl—running on NVIDIA GPUs
+implementations—MadNLP.jl and ExaModels.jl—running on NVIDIA GPUs
 achieve an order of magnitude speedup compared to state-of-the-art
@@
 \cite{anitescu2021targeting}. While GPU computation can trivially
-accelerate several parts of the optimization process --- especially
-various internal computations within the optimization solver --- the
+accelerate several parts of the optimization process---especially
+various internal computations within the optimization solver---the
 sluggish data transfer between host and device memory hampers the
@@
 NVIDIA GPUs, along with the associated software implementations:
-SIMDiff.jl \cite{simdiff}, an algebraic modeling/AD tool, and MadNLP.jl \cite{madnlp}, an
+ExaModels.jl \cite{simdiff}, an algebraic modeling/AD tool, and MadNLP.jl \cite{madnlp}, an
 NLP solver. Our approach incorporates two novel strategies: (i) a
@@
 Parallelizing this operation has been considered to be challenging,
-primarily due to the need for numerical pivoting, which is sequential
-in nature. However, when the matrix can be factorized without
+primarily due to the need for numerical pivoting, which requires irregular
+memory accesses and data movement \cite{swirydowicz2022linear}. However, when the matrix can be factorized without
 numerical pivoting, a significant part of the operation can be
@@
 our method, utilizing our two packages: MadNLP.jl and
-SIMDiff.jl. The solution of the KKT system is performed using the
+ExaModels.jl. The solution of the KKT system is performed using the
 external cuSOLVER library. To assess the performance of our method, we
@@
 \paragraph*{Contributions}
-We present, for the first time, a nonlinear optimization framework
-that can run entirely on GPU, with all the performance-critical data
-arrays residing exclusively on GPUs. Additionally, we introduce the
-concept of SIMD abstraction for NLP problems, which results in an
-efficient implementation of GPU-accelerated parallel AD. Furthermore,
-we propose the condensed IPM with an inequality relaxation strategy for
-the first time, enabling the treatment of the KKT systems of NLPs
-without numerical pivoting, thus allowing the solution of sparse,
-large-scale NLPs (with a prominent example being AC OPFs) on GPUs.
+We present, for the first time, a sparse nonlinear optimization
+solution framework that can run entirely on GPU, with all the
+performance-critical data arrays residing exclusively on
+GPUs. Additionally, we introduce the concept of SIMD abstraction for
+NLP problems, which results in an efficient implementation of
+GPU-accelerated parallel AD. Furthermore, we propose the condensed IPM
+with an inequality relaxation strategy for the first time, enabling
+the treatment of the KKT systems of NLPs without numerical pivoting,
+thus allowing the solution of sparse, large-scale NLPs (with a
+prominent example being AC OPFs) on GPUs.
 
@@
 of the AD and optimization capabilities in our
-tools, MadNLP.jl and SIMDiff.jl, are implemented in the Julia Language,
+tools, MadNLP.jl and ExaModels.jl, are implemented in the Julia Language,
 by leveraging its kernel and array programming capabilities.
@@
 sparse AD of the model equations. The abstraction and AD are
-implemented as part of our algebraic modeling language SIMDiff.jl.
+implemented as part of our algebraic modeling language ExaModels.jl.
 
@@
 To implement the SIMD abstraction in the modeling environment, the
-algebraic modeling interface in SIMDiff.jl enforces the users to
-specify the model equations in an {\tt Iterable} data type in
+algebraic modeling interface in ExaModels.jl enforces the users to
+specify the model equations in an {\tt Generator} data type in
 Julia Language. This composite data type consists of an instruction (a Julia
@@
 
-Using the multiple dispatch feature of Julia, SIMDiff.jl generates
+Using the multiple dispatch feature of Julia, ExaModels.jl generates
 highly efficient derivative computation code, specifically compiled
@@
 and implemented via array and kernel programming in Julia Language. In
-turn, SIMDiff.jl has the capability to efficiently evaluate first and
+turn, ExaModels.jl has the capability to efficiently evaluate first and
 second-order derivatives using GPU accelerators.
@@
 
-SIMDiff.jl exploits the SIMD abstraction of the model equations to
+ExaModels.jl exploits the SIMD abstraction of the model equations to
 save the computational cost spent for sparsity analysis.  This is
@@
 \begin{align}
-  \label{config-1}\tag{Config 1} \bullet\;&\text{MadNLP.jl + SIMDiff.jl + cuSOLVER (GPU)}\\
-  \label{config-2}\tag{Config 2} \bullet\;&\text{MadNLP.jl + SIMDiff.jl + Ma27 (CPU)}\\
-  \label{config-3}\tag{Config 3} \bullet\;&\text{Ipopt + AMPL + Ma27 (CPU)}\\
-  \label{config-4}\tag{Config 4} \bullet\;&\text{Ipopt + JuMP.jl + Ma27 (CPU)}.
+  \label{config-1}\tag{Config 1} \bullet\;&\text{MadNLP.jl+ExaModels.jl+cuSOLVER (GPU)}\\
+  \label{config-2}\tag{Config 2} \bullet\;&\text{MadNLP.jl+ExaModels.jl+Ma27 (CPU)}\\
+  \label{config-3}\tag{Config 3} \bullet\;&\text{Ipopt+AMPL+Ma27 (CPU)}\\
+  \label{config-4}\tag{Config 4} \bullet\;&\text{Ipopt+JuMP.jl+Ma27 (CPU)}.
 \end{align}
@@
 problem is formulated using the model from the rosetta-opf project
-\cite{rosetta-opf}, and the test cases are obtained from the pglib-opf
+\cite{rosetta-opf}, which is based on the models in PowerModels.jl \cite{8442948}, and the test cases are obtained from the pglib-opf
 repository \cite{babaeinejadsarookolaee2019power}. We have selected
@@
     \hline
-    & {\textbf{MadNLP.jl + SIMDiff.jl + cuSOLVER}}
-    & {\textbf{MadNLP.jl + SIMDiff.jl + Ma27}}
-    & {\textbf{Ipopt + AMPL + Ma27}}
-    & {\textbf{Ipopt + JuMP.jl + Ma27}}\\
-    &\textbf{(gpu)} &\textbf{(cpu)} &\textbf{(cpu)}& \textbf{(cpu)}\\
+    & {\textbf{MadNLP+ExaModels+cuSOLVER}}
+    & {\textbf{MadNLP+ExaModels+Ma27}}
+    & {\textbf{Ipopt+AMPL+Ma27}}
+    & {\textbf{Ipopt+JuMP+Ma27}}\\
+    &\textbf{(GPU)} &\textbf{(CPU)} &\textbf{(CPU)}& \textbf{(CPU)}\\
     \hline
@@
     \hline
-    \textbf{Derivative Evaluations} & \multicolumn{2}{c|}{SIMDiff.jl (dev)$^*$} &  AMPL Solver Library & JuMP.jl (v1.12.0)\\
+    \textbf{Derivative Evaluations} & \multicolumn{2}{c|}{ExaModels.jl (dev)$^*$} &  AMPL Solver Library & JuMP.jl (v1.12.0)\\
     \hline
@@
   \caption{Numerical Results}
-  \input{table.tex}\\
+  \input{./include/table.tex}\\
   $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.
@@
   \label{tbl:quality}
-  \input{table-2.tex}\\
+  \input{./include/table-2.tex}\\
 \end{table*}
@@
 Next, we discuss the effectiveness of parallel AD on GPUs. We observe
-that even on CPUs, SIMDiff.jl is substantially faster than the AD
-routines implemented in AMPL or JuMP.jl. Indeed, SIMDiff.jl generates
+that even on CPUs, ExaModels.jl is substantially faster than the AD
+routines implemented in AMPL or JuMP.jl. Indeed, ExaModels.jl generates
 derivative functions specifically compiled for the type of model,
 including optimization for the distinctive computational pattern found
-in the model. When comparing SIMDiff.jl running on CPUs and GPUs, we
+in the model. When comparing ExaModels.jl running on CPUs and GPUs, we
 observe a further speedup on the GPU of up to almost 20x for large
@@
 a factor of 10. These results, along with our packages
-MadNLP.jl and SIMDiff.jl, showcase a significant advancement in our
+MadNLP.jl and ExaModels.jl, showcase a significant advancement in our
 capabilities in dealing with large-scale optimization problems in

### 2023-08-22 coin-or/main.tex
coin-or letter added

@@
-\documentclass{article}
-
 \input{preamble}
+\usepackage{wrapfig}
 \title{ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
@@
 
-We are pleased to submit our nominations for the COIN-OR cup, highlighting our contributions of ExaModels and MadNLP in the domain of computational infrastructure for operations research, in the particular area of nonlinear optimization.
-Our open-source software packages represent a significant advance in addressing the challenges of efficiently solving large-scale nonlinear optimziation problems by harnessing the capabilities of modern GPU hardware.
-In this document, we highlight the key challenges and opportunities we are facing for the development of optimization software in 2023, especially in the context of the advent of accelerated computing.
-We also explain how ExaModels and MadNLP jointly address the key technical challanges and harness the benefits that modern accelerated computing has to offer.
-We also highlight several performance improvement we have achieved in terms of solving large-scale, real-world nonolinear optimziation problems.
-Finally, we conclude by discussing how our development sets the stage for the next-generation nonlinear optimization solver development and opens up new possibilities in various applications.
+We are delighted to submit our nominations for the COIN-OR Cup,
+highlighting our contributions of {\bf ExaModels} and {\bf MadNLP} in the field of
+computational infrastructure for operations research, specifically in
+the area of nonlinear programming (NLP). Our open-source software
+packages represent a significant breakthrough in efficiently solving
+large-scale NLPs by leveraging the capabilities of modern GPU
+hardware.  In this document, we address the primary challenges and
+opportunities for optimization software development in 2023,
+particularly within the context of accelerated computing. We elucidate
+how ExaModels and MadNLP collectively address technical obstacles and
+capitalize on the advantages offered by modern accelerated
+computing. Furthermore, we emphasize the remarkable performance
+enhancements achieved in solving real-world nonlinear optimization
+problems at scale. A noteworthy example of such progress is our recent
+accomplishment in accelerating alternating current (AC) optimal power
+flow (OPF) problems by up to ten times compared to state-of-the-art
+tools. Finally, we discuss how our development lays the foundation for
+the advancement in various topics in operations research and unlocks
+new possibilities.
 
-\paragraph{Challenges and Significance}
-The computational landscape of today presents us with immense challenges and potential opportunities, especially with Graphics Processing Units (GPUs) at our disposal.
-Notably, NVIDIA GPUs have enabled great stride in the scalabiltiy of deep learning and achieved remarkable success in building large-scale AI models.
-Also, in the public doamin, most of the computational power for the next generation exascale HPC systems, such as Frontier and Aurora, come from GPU accelerators.
-While the GPUs have enabled great success in different scientific computing areas, such as machine learning and high-fidelity simulations, utilizing GPUs within nonlinear programming (NLP) algorithms has been hindered by challenges like sparse automatic differentiation (AD) and sparse linear solver routines, which do not seamlessly translate to GPU architecture.
-The key issue is that the conventional sparse AD and matrix factorization algoirthms are built for CPUs, and they heavily rely on the serial computation.
+\paragraph{Challenges and Opportunities}
+In recent years, the scalability of deep learning has experienced
+significant advancements, primarily due to the utilization of NVIDIA
+GPUs, which have demonstrated remarkable success in constructing
+large-scale AI models. Additionally, in the public domain, GPU
+accelerators have provided a substantial portion of the computational
+power for next-generation leadership HPC systems such as Frontier and
+Aurora, achieving the milestone of one exaflop per second.  While GPUs
+have proven highly effective in various scientific computing areas
+like machine learning and high-fidelity simulations, integrating them
+into constrained nonlinear programming (NLP) algorithms has encountered
+considerable challenges. Specifically, the seamless translation of
+sparse automatic differentiation (AD) and sparse linear solver
+routines to GPU architecture has been recognized to be particularly
+challenging.  This is because conventional sparse AD and matrix
+factorization algorithms are designed for CPUs and heavily rely on
+serial computation, making them unsuitable for the single-instruction,
+multiple-data (SIMD) parallelism employed by GPU architectures (see
+Figure \ref{fig:simd}, left).
 
-While GPU computation can trivially accelerate several parts of the optimization process---especially various internal computations within the optimization solver---the sluggish data transfer between host and device memory hampers the ad-hoc implementation of GPU accelerations (Fig.
-\ref{fig:memory}).
-To fully leverage the potential offered by modern GPU hardware, it becomes imperative to have a comprehensive computational framework for optimization on GPUs.
-That is, we need an AD/algebraic modeling framework, sparse linear solvers, and NLP solvers that can operate entirely on the GPU.
-Specifically, for the best performance, both the problem data and the solver's intermediate computational data must be exclusively resident within the device memory, with the majority of operations executed on the GPU.
-This necessitates the development of {\it a comprehensive nonlinear optimization solution framework}, which performs all the necessary computations---automatic differentiation, nonlinear optimization, and linear algebra---exclusively on GPUs.
-ExaModels and MadNLP rise to these challenges, carving out a pathway to fully harness the power of GPUs for operaitons research and nonlinear optimization.
+While GPU computation can accelerate several parts of the optimization
+process straightforwardly, the slow data transfer between host and
+device memory hinders the ad-hoc implementation of GPU accelerations
+(Figure \ref{fig:simd}, right). To fully exploit the potential offered
+by modern GPU hardware, it is crucial to develop a comprehensive
+computational framework for optimization on GPUs. This requires
+keeping the problem data and the solver's intermediate computational
+data reside exclusively within device memory, with the majority of
+operations executed on the GPU. Consequently, the development of a
+comprehensive NLP solution framework that performs
+all necessary computations, including AD, optimization, and linear
+algebra, exclusively on GPUs becomes imperative.  To address these
+challenges, ExaModels and MadNLP offer a comprehensive solution
+framework for NLP. They enable AD and NLP solution
+exclusively on GPUs, while linear algebra computation is performed by
+the external library (e.g., cuSOLVER), allowing for the full
+utilization of GPUs.
 
@@
   \centering 
+  \scalebox{.82}{
+    \begin{tikzpicture}[remember picture, scale=.9, font=\small]
+    \node[draw,align=left] (I) at (0,2.75) {
+      $\bullet$ Instruction $g(\cdot,\cdot)$\\
+      $\bullet$ Pointer to array $x$
+    };
+    \node[draw] at (-4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (A\x) at (0,-\x/2) {$s_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (0,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \node[draw] at (4,0) {
+      \tikz{
+        \foreach\x in {1,2,3,4,5,7}{
+          \node (C\x) at (0,-\x/2) {$y_{\x}$};
+        }
+        \node at (0,-2.9) {$\vdots$};
+      }
+    };
+    \foreach\x in {1,2,3,4,5,7}{
+      \draw[->] (A\x.east) -- (B\x.west);
+      \draw[->] (B\x.east) -- (C\x.west);
+      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
+    }
+    \node at (-4,-2.5) {Input Data};
+    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
+    \node at (4,-2.5) {Output Data};
+  \end{tikzpicture} 
+  }
   \scalebox{.87}{
@@
   }
-  \includegraphics[width=.48\textwidth]{speedup-sol.pdf}
+  % \includegraphics[width=.48\textwidth]{speedup-sol.pdf} 
   \label{fig:simd}
-  \caption{A schematic of host (CPU) and device (GPU) memory structure (left) and speedup achieved for AC optimal power flow problems (right).}\label{fig:memory}
+  \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}\label{fig:memory}
 \end{figure}
 
-\paragraph{Developed Software Tools}
-We have developed a comprehesnive nonlinear optimization framework by implementing our algebraic modeling/automatic differentiation tool ExaModels and nonlinear optimization solver MadNLP, while linear algebra computation is currently performed by the external cuSOLVER library.
+\paragraph{ExaModels}
 
-{\it ExaModels}: SIMD Abstraction for NLPs: ExaModels pioneers the implementation of a groundbreaking single-instruction, multiple-data (SIMD) abstraction for NLPs.
-This revolutionary approach unlocks the potential for efficient parallel automatic differentiation on GPUs.
-The essence of this innovation lies in preserving the parallelizable structure within model equations, transforming derivative evaluations into streamlined operations.
-By enabling efficient computation of derivatives on GPUs, ExaModels offers an exceptional performance boost for power systems analysis.
+ExaModels is an algebraic modeling and AD system embedded in the Julia
+Language, specialized for the SIMD abstraction of nonlinear
+programs. This system is designed based on the repetitive structure
+commonly found in large-scale NLPs. For example, AC OPF problems,
+which may have millions of variables and constraints, can be expressed
+using just 15 computational patterns. By enforcing the creation of
+models through specifying repetitive patterns, ExaModels preserves the
+parallelizable structure within model equations and facilitates
+parallel automatic differentiation. Writing models in ExaModels
+automatically enables GPU compatibility, which is especially
+beneficial as optimization models are often implemented by domain
+experts who may not fully understand the internals of algebraic
+modeling systems and NLP solvers.  This makes ExaModels a powerful
+modeling platform for scalable computations.
 
-The optimization models are most of the times implemented by the researchers and practitioners who do not necessarily understand the operations 
-ExaModels provides a new platform for scalable nonlinear optimization.
-Our numerical results suggest that the conventional algebraic modeling systems are limited in terms of efficiently computing the derivates of the model equations (Table \ref{tab:num}), and it does not allow performing the operations on GPUs (Table \ref{tbl:portabiltiy}).
-ExaModels provide a framework that allows the user to conveniently inform the AD backend the parallelizable structure.
-Writing the models in this way will enable the GPU compatibility
+By enabling efficient computation of derivatives on GPUs, ExaModels
+provides exceptional performance, surpassing existing tools like AMPL
+(AD backend of Pyomo) and JuMP by more than two orders of
+magnitude. Our numerical results demonstrate that conventional
+algebraic modeling systems struggle to efficiently compute derivatives
+of model equations (Table \ref{tab:num}), and they are incompatible
+with GPU operations (Table \ref{tbl:portability}). In contrast,
+ExaModels provides a convenient framework for users to inform the AD
+backend about the parallelizable structure, and in turn, efficient,
+parallelized AD can be performed in the backend.
 
-{\it MadNLP}: Condensed-Space IPM with Inequality Relaxation: MadNLP introduces an equally transformative approach through its condensed-space interior-point method (IPM) with an inequality relaxation strategy.
-This strategy deftly overcomes the challenges posed by sparse matrix factorization on GPUs.
-The relaxation of equality constraints and condensation of the Karush-Kuhn-Tucker (KKT) system establishes positive definiteness, paving the way for the utilization of highly efficient linear solvers.
-MadNLP thus unveils a remarkable solution to traditionally intractable problems in GPU-based optimization.
+\paragraph{MadNLP}
 
-While MadNLP has started as a port of Ipopt on Julia Language, now marking as the fourth year since its initial development, MadNLP has become a lot more versatile solver than the previously developed nonlinear optimization solvers.
-The key features include (1) being able to solve dense optimzation problems, (2) ability to handle different forms of KKT systems, (3) being able to handle diverse array data types (most notably, device arrays), and (4) exploit various Hessian approximation strategies (BFGS variants and limited-memory BFGS methods).
+MadNLP is a versatile NLP solver capable of
+handling various problem data structures. By leveraging the modularity
+provided by the key features of the Julia Language, such as multiple
+dispatch and just-in-time compilation, MadNLP offers great flexibility
+in handling different types of data structures while retaining the
+sophistication of the algorithm. Importantly, with this approach, we can avoid
+re-implementing interior point solver every time for each data types,
+which in turn ensures that the operations performed are mathematically
+equivalent to those in the mature, extensively tested existing code
+base. MadNLP has started as a port of Ipopt (an open-source nonlinear
+optimziation solver running on CPUs) in the Julia Language, but over
+the past four years, it has expanded beyond its initial scope and now
+offers a range of capabilities not supported by state-of-the-art NLP
+solvers. These include solving dense optimization problems, handling
+different numerical precisions, working with different forms of KKT
+systems, managing diverse array data types (including GPU device
+arrays), and utilizing various Hessian approximation strategies, such
+as different variants of BFGS methods.
+
+Recently, we have discovered that the condensed-space interior point
+method (IPM) with an inequality relaxation strategy is particularly
+effective for solving large-scale NLPs, like AC optimal power flow, up
+to moderate precisions. This strategy efficiently addresses the
+challenges posed by sparse matrix factorization on GPUs, as the
+condensed KKT systems can be factorized without numerical pivoting,
+which has previously hindered GPU utilization in the large-scale
+optimization regime.  Furthermore, MadNLP offers diverse approaches
+for handling KKT systems with special structures, via reduction
+strategies (citation needed) and Schur complement decomposition
+strategies (citation needed).
 
@@
 
-The significance of these contributions is not merely theoretical.
-Empirical results speak volumes about the impact of ExaModels and MadNLP.
-When applied to AC OPF problems, the performance enhancements achieved are nothing short of extraordinary.
-The GPU-based solutions attain speedups of over 20 times compared to CPU-based counterparts, showcasing the profound potential of these software packages in power systems analysis.
-Their prowess surpasses even established tools interfaced with CPUs, marking a paradigm shift in computational efficiency.
-
-\cite{shin2023accelerating}
+Our numerical results, summarized in Table \ref{tab:num} (for detailed
+results, please refer to \cite{shin2023accelerating}), clearly
+demonstrate the significant potential of using GPUs. Specifically,
+when applied to AC OPF problems, which are one of the most important
+applications of NLP, we have achieved significant performance
+enhancements. The GPU-based solutions achieve speedups of over 10
+times compared to CPU-based state-of-the-art tools. Notably, ExaModels
+enables derivative evaluations that are up to two orders of magnitude
+faster, and the use of the condensed-space IPM strategy in MadNLP
+greatly improves linear algebra computation speed. Furthermore, we
+have also explored the utilization of these methods in reduced-space
+OPF \cite{} and model predictive control (MPC) \cite{}, where we have
+achieved similar degrees of speedup.
 
@@
     \footnotesize
-    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  \caption{Numerical Performance of ExaModels and MadNLP}
+    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
   \end{center}
@@
 
-While NVIDIA seems to be most capable in computing power and mature in terms of the implementation of the low-level operations (especially, the CUDA libraries), we aim to support all the other alternative architectures.
-This is because quite a few next-generaiton exascale high-performance computing architectures will be based on AMD and Intel accelerator devices.
-Also, as an open-source software, we aim to be the most generic software tool, not the most efficient one.
-By supporting diverse platforms, we hope to spark the attention of the chip makers in the operations research area and encourage better support for the software infrastructure for implmenting operations research software.
+While NVIDIA currently leads in terms of computing power and mature
+implementation of low-level operations, especially with CUDA
+libraries, we aim to provide support for alternative architectures as
+well. This is crucial because many next-generation exascale
+high-performance computing architectures will be based on AMD and
+Intel accelerator devices. Furthermore, by supporting diverse
+platforms, we hope to drive the attention of chip makers towards the
+operations research field and encourage better support for software
+infrastructure in implementing operations research software.
 
-Table \ref{tbl:portabiltiy} summarizes the current state of the portability of our tools and other existing open-source software tools.
-Notably, ExaModels.jl already support all major architectures, whereas MadNLP.jl only support NVIDIA GPUs.
-Nonetheless, we have clear path forward in supporting all the major accelerator architectures, including NVIDIA, AMD, and Intel GPUs, while Apple GPUs are somewhat limited due to their inability to perform double precision operations.
-This is due to the fact that all of our GPU kernels are implemented via KernelAbstractions.jl, a julia package enabling portable implementation of GPU kernels.
-In this way, the GPU kernels for diverse accelerator architectures can be run based on a sam source code.
-Using this strategy, recently ExaModels.jl has demonstrated its capabiltiy to run on 4 different accelerator architectures (multi-thread CPUs and NVIDIA, AMD, and Intel GPUs).
-In the coming years, we plan to exand MadNLP's support for various accelerator architectures.
+Table \ref{tbl:portability} provides an overview of the current state
+of portability for our tools and other existing open-source software
+tools. ExaModels.jl already supports all major architectures, while
+MadNLP.jl currently only supports NVIDIA GPUs. However, we have a
+clear roadmap to extend support to all major accelerator
+architectures, including NVIDIA, AMD, and Intel GPUs. The portability
+is enabled by KernelAbstractions.jl, a Julia package that facilitates
+the portable implementation of GPU kernels.
 
-However, at the same time, it is absolutely crucial to have efficient and reliable, and preferrably also portable sparse linear sovler routines.
-Currently, we rely on CUSOLVER library for solving the linear systems.
-While vendor-implmented libraries are expected to have the best performance via high level of optimization, it would be beneficial for the community to have the open-source portable linear solvers.
-By showcasing the capability of solving high-stake nonlinear optimziation problems on GPUs, we can attract the attention of the numerical linear algebra community so that we can encourage the development of capable open-source sparse linear solvers.
-1
 \begin{table}[t]
@@
         \hline
-        \multirow{3}{*}{Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
-        &{\bf JuMP}& \xmark& \cmark& \xmark& \xmark& \xmark& \xmark\\
+        \multirow{3}{*}{Algebraic Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
+        &{\bf JuMP}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
         &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\ 
         \hline
-        \multirow{2}{*}{Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
+        \multirow{2}{*}{NLP Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
         & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
@@
   \end{center}
-  \caption{The current status on the portability of the nonlinear optimization frameworks.}
+  \caption{The current status on the portability of the NLP frameworks.}
   \label{tbl:portabiltiy}
@@
 
-\paragraph{Impacts in Various Applications}
+\paragraph{Expected Impacts in Applications}
 
-{\it Energy Infrastructures.} The resillience We have recently showcased the capability of MadNLP with multiple GPUs for solving extremely-large scale security-constrained AC OPF problems.
+\begin{itemize}
+\item \textit{Energy Infrastructures}: Enhancing the scalability of
+optimization methods is crucial for improving the resilience of energy
+system operations. We have demonstrated the ability of MadNLP with
+multiple GPUs to solve large-scale security-constrained AC OPF
+problems effectively. This showcases the potential of ExaModels and
+MadNLP to enhance the scalability of NLP by utilizing GPU
+capabilities.
 
-{\it 1}
+\item \textit{Model Predictive Control}: MPC has been widely used for
+optimization-based control in various settings, such as chemical
+processes, energy systems, and robotics. However, the effectiveness of
+MPC is often limited by the efficiency of optimization solvers,
+especially when short sampling times are required. Our results have
+shown that MadNLP running on GPUs can solve such problems with high
+efficiency \cite{}. Moving forward, we plan to implement the extension
+packages for control problems so that control community can take
+advantage of accelerated NLP frameworks.
 
-{\it Machine Learning Surrogate Models.}
+\item \textit{Machine Learning Surrogate Models}: An exciting future
+direction is the integration of neural network surrogate models into
+constrained optimization algorithms. GPU acceleration plays a crucial
+role in these applications, considering the complexity of real-world
+AI models. MadNLP offers the necessary capabilities through the native
+handling of device arrays as well as dense KKT systems with various
+quasi-Newton strategies. We intend to explore these problems further
+and provide extensions for surrogate modeling with GPU acceleration.
 
-\paragraph{Closing Remarks}
-ExaModels and MadNLP embody the spirit of innovation and progress that the COIN-OR cup seeks to recognize.
-By addressing critical challenges and introducing transformative strategies, these packages extend the frontiers of computational infrastructure for operations research.
-We are confident that their exceptional contributions will resonate within the operations research community and beyond, redefining the future of power systems analysis and optimization.
-It is with great enthusiasm that we nominate ExaModels and MadNLP for the COIN-OR cup, recognizing their exceptional impact on the field.
+\end{itemize}
+
+\paragraph{Closing Remarks} ExaModels and MadNLP provide computational
+infrastructures for operations research that align with the era of
+accelerated computing, which we believe the COIN-OR cup aims to
+recognize. By addressing critical challenges and introducing new
+strategies, these packages significantly extend the boundaries of
+computational infrastructure for operations research. We are confident
+that these contributions will resonate within the operations research
+community and beyond, opening up new possibilities in applications
+such as energy systems, control, and machine learning. With great
+enthusiasm, we nominate ExaModels and MadNLP for the COIN-OR cup, in
+recognition of their impact on the field.
 
 \bibliography{main}
-
 \end{document}

### 2023-08-22 coin-or/main.tex
references added

@@
   }
-  % \includegraphics[width=.48\textwidth]{speedup-sol.pdf} 
+  \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}
   \label{fig:simd}
-  \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}\label{fig:memory}
 \end{figure}
@@
 for handling KKT systems with special structures, via reduction
-strategies (citation needed) and Schur complement decomposition
-strategies (citation needed).
+strategies \cite{pacaud2023accelerating} and Schur complement decomposition
+strategies \cite{pacaud2023parallel}.
 
@@
 have also explored the utilization of these methods in reduced-space
-OPF \cite{} and model predictive control (MPC) \cite{}, where we have
-achieved similar degrees of speedup.
+OPF \cite{pacaud2023accelerating} and model predictive control (MPC)
+\cite{cole2023exploiting}, where we have achieved similar degrees of
+speedup.
 
@@
     \footnotesize
-    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
+    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  
   \end{center}
+  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
   \label{tab:num}
@@
   \caption{The current status on the portability of the NLP frameworks.}
-  \label{tbl:portabiltiy}
+  \label{tbl:portability}
 \end{table}
@@
 multiple GPUs to solve large-scale security-constrained AC OPF
-problems effectively. This showcases the potential of ExaModels and
-MadNLP to enhance the scalability of NLP by utilizing GPU
-capabilities.
+problems effectively \cite{pacaud2023parallel}. This showcases the
+potential of ExaModels and MadNLP to enhance the scalability of NLP by
+utilizing GPU capabilities.
 
@@
 shown that MadNLP running on GPUs can solve such problems with high
-efficiency \cite{}. Moving forward, we plan to implement the extension
-packages for control problems so that control community can take
-advantage of accelerated NLP frameworks.
+efficiency \cite{cole2023exploiting}. Moving forward, we plan to
+implement the extension packages for control problems so that control
+community can take advantage of accelerated NLP frameworks.
 
@@
 direction is the integration of neural network surrogate models into
-constrained optimization algorithms. GPU acceleration plays a crucial
-role in these applications, considering the complexity of real-world
-AI models. MadNLP offers the necessary capabilities through the native
+constrained optimization algorithms (a proof-of-concept presented in
+\cite{shin2023constrained}). GPU acceleration plays a crucial role in
+these applications, considering the complexity of real-world AI
+models. MadNLP offers the necessary capabilities through the native
 handling of device arrays as well as dense KKT systems with various
@@
 
+\bibliographystyle{plain} 
 \bibliography{main}

### 2023-08-22 coin-or/main.tex
coin-or nomination added

@@
 We are delighted to submit our nominations for the COIN-OR Cup,
-highlighting our contributions of {\bf ExaModels} and {\bf MadNLP} in the field of
-computational infrastructure for operations research, specifically in
-the area of nonlinear programming (NLP). Our open-source software
-packages represent a significant breakthrough in efficiently solving
+highlighting our contributions of {\bf ExaModels} \cite{examodels} and
+{\bf MadNLP} \cite{madnlp} in the field of computational
+infrastructure for operations research, specifically in the area of
+nonlinear programming (NLP). Our open-source software packages
+represent a significant breakthrough in efficiently solving
 large-scale NLPs by leveraging the capabilities of modern GPU
@@
 
-\paragraph{Challenges and Opportunities}
-In recent years, the scalability of deep learning has experienced
-significant advancements, primarily due to the utilization of NVIDIA
+\paragraph{Challenges and Opportunities} In recent years, the
+scalability of deep learning has experienced significant advancements,
+primarily due to the utilization of NVIDIA
 GPUs, which have demonstrated remarkable success in constructing
@@
 like machine learning and high-fidelity simulations, integrating them
-into constrained nonlinear programming (NLP) algorithms has encountered
+into constrained NLP algorithms has encountered
 considerable challenges. Specifically, the seamless translation of

### 2023-08-23 coin-or/main.tex
name spells fixed
[coin-or] minor modifications

@@
 \usepackage{wrapfig}
-\title{ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
-\author{Sungho Shin$^\dag$, Francois Pacaud$^\ddag$, and Mihai Anitescu$^\dag$} 
+\title{ExaModels and MadNLP: An Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
+\author{Sungho Shin$^\dag$, François Pacaud$^\ddag$, and Mihai Anitescu$^\dag$}
 \date{\small
@@
 accomplishment in accelerating alternating current (AC) optimal power
-flow (OPF) problems by up to ten times compared to state-of-the-art
+flow (OPF) problems by up to a factor of ten compared to state-of-the-art
 tools. Finally, we discuss how our development lays the foundation for
@@
 
-\paragraph{Challenges and Opportunities} In recent years, the
+\paragraph{Challenges and Opportunities.} In recent years, the
 scalability of deep learning has experienced significant advancements,
@@
 GPUs, which have demonstrated remarkable success in constructing
-large-scale AI models. Additionally, in the public domain, GPU
+large-scale AI models. Additionally, in the high-performance computing (HPC) domain, GPU
 accelerators have provided a substantial portion of the computational
@@
 like machine learning and high-fidelity simulations, integrating them
-into constrained NLP algorithms has encountered
+into nonlinear optimization solvers has encountered
 considerable challenges. Specifically, the seamless translation of
@@
 computational framework for optimization on GPUs. This requires
-keeping the problem data and the solver's intermediate computational
-data reside exclusively within device memory, with the majority of
+keeping both the problem data and the solver's intermediate computational
+data exclusively within device memory, with the majority of
 operations executed on the GPU. Consequently, the development of a
 comprehensive NLP solution framework that performs
-all necessary computations, including AD, optimization, and linear
-algebra, exclusively on GPUs becomes imperative.  To address these
+all necessary computations --- including AD, optimization, and linear
+algebra --- exclusively on GPUs becomes imperative.  To address these
 challenges, ExaModels and MadNLP offer a comprehensive solution
-framework for NLP. They enable AD and NLP solution
+framework for NLP. They respectively enable AD and NLP solution
 exclusively on GPUs, while linear algebra computation is performed by
-the external library (e.g., cuSOLVER), allowing for the full
+external libraries (e.g., cuSOLVER), allowing for the full
 utilization of GPUs.
@@
 
-\paragraph{ExaModels}
+\paragraph{ExaModels.}
 
@@
 commonly found in large-scale NLPs. For example, AC OPF problems,
-which may have millions of variables and constraints, can be expressed
+which may encompass millions of variables and constraints, can be expressed
 using just 15 computational patterns. By enforcing the creation of
-models through specifying repetitive patterns, ExaModels preserves the
+models through the specification of repetitive patterns, ExaModels preserves the
 parallelizable structure within model equations and facilitates
@@
 algebraic modeling systems struggle to efficiently compute derivatives
-of model equations (Table \ref{tab:num}), and they are incompatible
-with GPU operations (Table \ref{tbl:portability}). In contrast,
+of model equations (Table~\ref{tab:num}), and they are incompatible
+with GPU operations (Table~\ref{tbl:portability}). In contrast,
 ExaModels provides a convenient framework for users to inform the AD
-backend about the parallelizable structure, and in turn, efficient,
-parallelized AD can be performed in the backend.
+backend about the parallelizable structure, enabling efficient
+parallelized AD in the backend.
 
-\paragraph{MadNLP}
+\paragraph{MadNLP.}
 
@@
 base. MadNLP has started as a port of Ipopt (an open-source nonlinear
-optimziation solver running on CPUs) in the Julia Language, but over
+optimization solver running on CPUs) in the Julia Language, but over
 the past four years, it has expanded beyond its initial scope and now
@@
 solvers. These include solving dense optimization problems, handling
-different numerical precisions, working with different forms of KKT
+different numerical precisions, working with different forms of Karush-Kuhn-Tucker (KKT)
 systems, managing diverse array data types (including GPU device
@@
 effective for solving large-scale NLPs, like AC optimal power flow, up
-to moderate precisions. This strategy efficiently addresses the
+to moderate accuracy. This strategy efficiently addresses the
 challenges posed by sparse matrix factorization on GPUs, as the
@@
 
-While NVIDIA currently leads in terms of computing power and mature
-implementation of low-level operations, especially with CUDA
+While NVIDIA is currently the leader in terms of computing power and mature
+implementation of low-level operations on GPUs, especially with CUDA
 libraries, we aim to provide support for alternative architectures as
-well. This is crucial because many next-generation exascale
-high-performance computing architectures will be based on AMD and
-Intel accelerator devices. Furthermore, by supporting diverse
-platforms, we hope to drive the attention of chip makers towards the
-operations research field and encourage better support for software
-infrastructure in implementing operations research software.
+well. This is crucial as many next-generation exascale HPC
+architectures will be based on AMD and Intel accelerator devices.
+Furthermore, by supporting diverse platforms, we hope to drive the attention
+of chip makers towards the operations research field and encourage better
+support for software infrastructure in implementing operations research software.
 
@@
 of portability for our tools and other existing open-source software
-tools. ExaModels.jl already supports all major architectures, while
+tools. ExaModels.jl already supports all major architectures at
+the exception of Apple Metal, while
 MadNLP.jl currently only supports NVIDIA GPUs. However, we have a
@@
         \hline
-        &&CPU (single) &CPU (multi)& NVIDIA & AMD & Intel& Apple\\
+        &&CPU (single) &CPU (multi)& NVIDIA GPU & AMDGPU  & Intel GPU& Apple Metal\\
         \hline
@@
 optimization methods is crucial for improving the resilience of energy
-system operations. We have demonstrated the ability of MadNLP with
-multiple GPUs to solve large-scale security-constrained AC OPF
-problems effectively \cite{pacaud2023parallel}. This showcases the
+system operations. We have demonstrated the ability of MadNLP
+to solve large-scale security-constrained AC OPF
+problems effectively with multiple GPUs~\cite{pacaud2023parallel}. This showcases the
 potential of ExaModels and MadNLP to enhance the scalability of NLP by
@@
 processes, energy systems, and robotics. However, the effectiveness of
-MPC is often limited by the efficiency of optimization solvers,
+MPC is often limited by the latency of optimization solvers,
 especially when short sampling times are required. Our results have
@@
 efficiency \cite{cole2023exploiting}. Moving forward, we plan to
-implement the extension packages for control problems so that control
-community can take advantage of accelerated NLP frameworks.
+implement extension packages for control problems to enable
+the control community to take advantage of accelerated NLP frameworks.
 
@@
 
-\paragraph{Closing Remarks} ExaModels and MadNLP provide computational
-infrastructures for operations research that align with the era of
+\paragraph{Closing Remarks.} ExaModels and MadNLP provide computational
+infrastructures for operations research that align with the new era of
 accelerated computing, which we believe the COIN-OR cup aims to

### 2023-10-27 coin-or/main.tex
coin-or and AIChE talk

@@
 \usepackage{wrapfig}
-\title{ExaModels and MadNLP: An Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
-\author{Sungho Shin$^\dag$, François Pacaud$^\ddag$, and Mihai Anitescu$^\dag$}
+\title{
+  \vspace{-.35in}
+  ExaModels and MadNLP: An Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs
+  \vspace{-.15in}
+}
+\author{
+  Sungho Shin$^\dag$, François Pacaud$^\ddag$, and Mihai Anitescu$^\dag$
+}
 \date{\small
+  \vspace{-.1in}
   $^\dag$Mathematics and Computer Science Division, Argonne National Laboratory\\
   $^\ddag$Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France
+  \vspace{-.1in}
 }
@@
 represent a significant breakthrough in efficiently solving
-large-scale NLPs by leveraging the capabilities of modern GPU
-hardware.  In this document, we address the primary challenges and
+large-scale NLPs by leveraging the capabilities of modern GPUs.
+In this document, we address the primary challenges and
 opportunities for optimization software development in 2023,
-particularly within the context of accelerated computing. We elucidate
-how ExaModels and MadNLP collectively address technical obstacles and
+particularly in the context of accelerated computing. We elucidate
+how ExaModels and MadNLP address technical obstacles and
 capitalize on the advantages offered by modern accelerated
@@
 have proven highly effective in various scientific computing areas
-like machine learning and high-fidelity simulations, integrating them
+like machine learning (ML) and simulations, integrating them
 into nonlinear optimization solvers has encountered
@@
 comprehensive NLP solution framework that performs
-all necessary computations --- including AD, optimization, and linear
-algebra --- exclusively on GPUs becomes imperative.  To address these
+all necessary computations---including AD, optimization, and linear
+algebra---exclusively on GPUs becomes imperative.  To address these
 challenges, ExaModels and MadNLP offer a comprehensive solution
@@
   \centering
-  \scalebox{.82}{
+  \scalebox{.7}{
     \begin{tikzpicture}[remember picture, scale=.9, font=\small]
@@
   }
-  \scalebox{.87}{
+  \scalebox{.75}{
   \begin{tikzpicture}[remember picture, scale=.9, font=\small]
@@
 }
+\vspace{-.075in}
 \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}
+\vspace{-.075in}
 \label{fig:simd}
@@
 condensed KKT systems can be factorized without numerical pivoting,
-which has previously hindered GPU utilization in the large-scale
-optimization regime.  Furthermore, MadNLP offers diverse approaches
-for handling KKT systems with special structures, via reduction
-strategies \cite{pacaud2023accelerating} and Schur complement decomposition
-strategies \cite{pacaud2023parallel}.
+which has previously hindered GPU utilization.  Furthermore, MadNLP
+offers diverse approaches for handling KKT systems with special
+structures, via reduction strategies \cite{pacaud2023accelerating} and
+Schur complement decomposition strategies \cite{pacaud2023parallel}.
 
@@
 
-Our numerical results, summarized in Table \ref{tab:num} (for detailed
-results, please refer to \cite{shin2023accelerating}), clearly
-demonstrate the significant potential of using GPUs. Specifically,
-when applied to AC OPF problems, which are one of the most important
-applications of NLP, we have achieved significant performance
-enhancements. The GPU-based solutions achieve speedups of over 10
-times compared to CPU-based state-of-the-art tools. Notably, ExaModels
-enables derivative evaluations that are up to two orders of magnitude
-faster, and the use of the condensed-space IPM strategy in MadNLP
-greatly improves linear algebra computation speed. Furthermore, we
-have also explored the utilization of these methods in reduced-space
-OPF \cite{pacaud2023accelerating} and model predictive control (MPC)
+Our numerical results for AC optimal power flow problems, summarized
+in Table \ref{tab:num} (for detailed results, please refer to
+\cite{shin2023accelerating}), clearly demonstrate the significant
+potential of using GPUs. Specifically, when applied to AC OPF
+problems, which are one of the most important applications of NLP, we
+have achieved significant performance enhancements. The GPU-based
+solutions achieve speedups of over 10 times compared to CPU-based
+state-of-the-art tools. Notably, ExaModels enables derivative
+evaluations that are up to two orders of magnitude faster, and the use
+of the condensed-space IPM strategy in MadNLP greatly improves linear
+algebra computation speed. Furthermore, we have also explored the
+utilization of these methods in reduced-space OPF
+\cite{pacaud2023accelerating} and model predictive control (MPC)
 \cite{cole2023exploiting}, where we have achieved similar degrees of
@@
     \footnotesize
-    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.
   \end{center}
-  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
+  \vspace{-.075in}
+  \caption{Solution times (sec) for solving AC OPF problems}
+  \vspace{-.075in}
   \label{tab:num}
@@
   \end{center}
+  \vspace{-.075in}
   \caption{The current status on the portability of the NLP frameworks.}
+  \vspace{-.075in}
   \label{tbl:portability}
@@
 optimization methods is crucial for improving the resilience of energy
-system operations. We have demonstrated the ability of MadNLP
-to solve large-scale security-constrained AC OPF
-problems effectively with multiple GPUs~\cite{pacaud2023parallel}. This showcases the
-potential of ExaModels and MadNLP to enhance the scalability of NLP by
-utilizing GPU capabilities.
+system operations as the multi-period and stochastic formulations for
+energy networks are large-scale. We have recently demonstrated the
+ability of MadNLP to solve large-scale security-constrained AC OPF
+problems effectively with multiple GPUs~\cite{pacaud2023parallel}.
+This showcases the potential of accelerated computing, enabled by
+ExaModels and MadNLP, to contribute to enhancing the grid efficiency
+and resillience.
 
@@
 implement extension packages for control problems to enable
-the control community to take advantage of accelerated NLP frameworks.
+the control community to take advantage of better operations research
+software with accelerated computing capabilities.
 
@@
 direction is the integration of neural network surrogate models into
-constrained optimization algorithms (a proof-of-concept presented in
-\cite{shin2023constrained}). GPU acceleration plays a crucial role in
-these applications, considering the complexity of real-world AI
-models. MadNLP offers the necessary capabilities through the native
-handling of device arrays as well as dense KKT systems with various
-quasi-Newton strategies. We intend to explore these problems further
-and provide extensions for surrogate modeling with GPU acceleration.
+constrained optimization algorithms (proof-of-concept in
+\cite{shin2023constrained}). The development of the new award-winning
+software OMLT represents the increasing interest in incorporating ML
+models into optimization problems \cite{ceccon2022omlt}. GPU
+acceleration will play a crucial role in these applications,
+considering the complexity of real-world AI models. MadNLP offers the
+necessary capabilities through the native handling of device arrays as 
+well as dense KKT systems with various quasi-Newton strategies. We
+intend to explore these problems further and provide extensions for
+surrogate modeling with GPU acceleration.
 

### 2023-10-27 paper/cover/main.tex
coin-or and AIChE talk

\documentclass[onecolumn,oneside,11pt]{article} 
\usepackage[left=1in,right=1in,top=1cm,bottom=0.35in,%
includeheadfoot,headheight=99pt,headsep=0.2in]{geometry}
\usepackage{times,graphicx,cite,fancyhdr,amsmath,amssymb,captdef}
\usepackage{marvosym}
\renewcommand{\headrulewidth}{0pt}
%----------------  page style -------------------
\pagestyle{fancy}
\fancyhead{}    %clear the head
\fancyhead[L]{
  \begin{minipage}[t]{1.8in}
    \hspace{-0.6in}
    \includegraphics[height=0.6in]{anl.png}
  \end{minipage}
}
\fancyhead[C]{}
\fancyhead[R]{%
  \begin{minipage}[t]{1.7in}
    \begin{flushleft}
      {\scriptsize
        \vspace{-1.1in}
        {\bf Sungho Shin}\\
        \vspace{0.08in}
        {\bf Mathematics \& Computer Science Division}\\
        Argonne National Laboratory \\
        9700 South Cass Avenue\\
        Lemont, IL 60439 \\
        \vspace{0.08in}
        \Telefon: 1-608-448-5155 \\
        \Email: sshin@mcs.anl.gov}
    \end{flushleft}
  \end{minipage}
}
\fancyfoot{}    %clear the foot
\fancyfoot[C]{
  \footnotesize{A U.S. Department of Energy laboratory managed by UChicago
    Argonne, LLC}
}
% ==============================================================================
\setlength{\parskip}{2ex}
\setlength{\parindent}{0pt}

\begin{document}

\begin{flushleft}
  \today
\end{flushleft}

Dear PSCC 2024 Editorial Board:

I am pleased to submit our conference paper titled ``Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods'' for consideration for publication in the PSCC 2024 conference.

This manuscript, authored by Sungho Shin, François Pacaud, and Mihai Anitescu, presents a novel framework for solving alternating current (AC) optimal power flow (OPF) problems on graphics processing units (GPUs). While solving ACOPF problems on GPUs has garnered significant interest recently, the implementation of efficient solution algorithms has been hampered by the lack of efficient automatic differentiation tools and sparse linear solvers. We introduce two novel algorithmic approaches to address these challenges: single instruction, multiple data  abstraction of nonlinear programs and condensed-space interior-point methods. These new methods enable us to efficiently solve large-scale ACOPF problems efficiently on GPUs. We present numerical results, along with our open-source packages ExaModels.jl and MadNLP.jl, demonstrating that large-scale AC OPF instances can be solved up to 10 times faster than with the state-of-the-art tools running on CPUs.

We kindly request permission to publish the current version of the manuscript, which spans 10 pages. While we believe the paper is presented in a concise manner; the concepts discussed (SIMD abstraction of nonlinear programs and condensed-space interior point method) necessitate ample space for a thorough explanation. Furthermore, presenting the numerical results to showcase these new capabilities requires room to include tables displaying the data. Reducing the paper length to 7 pages  would entail a significant loss of content, potentially diminishing the paper's overall impact. If further justification is needed for the additional pages, please do not hesitate to contact us.

Thank you for considering our submission. We eagerly anticipate hearing from you regarding the outcome of the review process.

Sincerely,

\vspace{1cm}
Sungho Shin\\
Postdoctoral Appointee\\
Mathematics and Computer Science Division\\
Argonne National Laboratory\\
sshin@anl.gov

\end{document}

### 2023-10-27 paper/main.tex
coin-or and AIChE talk

@@
-% \documentclass{IEEEtran4PSCC} % will be used for submission
-% \author{
-%   \IEEEauthorblockN{Sungho Shin and Mihai Anitescu}
-%   \IEEEauthorblockA{
-%     Mathematics and Computer Science Division\\
-%     Argonne National Laboratory\\
-%     Lemont, IL, USA\\
-%     sshin@anl.gov, anitescu@mcs.anl.gov}
-%   \and
-%   \IEEEauthorblockN{François Pacaud}
-%   \IEEEauthorblockA{Centre Automatique et Systèmes\\
-%     Mines Paris - PSL \\
-%     Paris, France\\
-%     francois.pacaud@minesparis.psl.eu}
-% }
-% \psccfooter{%
-%         \parbox{\textwidth}{\hrulefill \\ \small{23rd Power Systems Computation Conference} \hfill \begin{minipage}{0.2\textwidth}\centering \vspace*{4pt} \includegraphics[scale=0.06]{PSCC_logo.png}\\\small{PSCC 2024} \end{minipage} \hfill \small{Paris, France --- June 4 -- 7, 2024}}%
-% }
-
-\documentclass{IEEEtran} % for ArXiv submission
+\documentclass{IEEEtran4PSCC} % will be used for submission
 \author{
-  Sungho Shin, François Pacaud, Mihai Anitescu
-  \thanks{
-    This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research (ASCR) under Contract DE-AC02-06CH11347.
+  \IEEEauthorblockN{Sungho Shin and Mihai Anitescu}
+  \IEEEauthorblockA{
+    Mathematics and Computer Science Division\\
+    Argonne National Laboratory,
+    Lemont, IL, USA\\
+    sshin@anl.gov, anitescu@mcs.anl.gov}
+  \and
+  \IEEEauthorblockN{François Pacaud}
+  \IEEEauthorblockA{Centre Automatique et Systèmes\\
+    Mines Paris - PSL,
+    Paris, France\\
+    francois.pacaud@minesparis.psl.eu}
 }
-  \thanks{
-    Sungho Shin and Mihai Anitescu are with Mathematics and Computer Science Division,  Argonne National Laboratory, Lemont, IL, USA. e-mail: \url{sshin@anl.gov}, \url{anitescu@mcs.anl.gov}
-  }
-  \thanks{
-    François Pacaud is with Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France. e-mail: \url{francois.pacaud@minesparis.psl.eu}
-  }
-}
-
-\hyphenation{op-tical net-works semi-conduc-tor}
-
 % Set footer
@@
 
+\psccfooter{%
+        \parbox{\textwidth}{\hrulefill \\ \small{23rd Power Systems Computation Conference} \hfill \begin{minipage}{0.2\textwidth}\centering \vspace*{4pt} \includegraphics[scale=0.06]{PSCC_logo.png}\\\small{PSCC 2024} \end{minipage} \hfill \small{Paris, France --- June 4 -- 7, 2024}}%
+}
+
+% \documentclass{IEEEtran} % for ArXiv submission
+% \author{
+%   Sungho Shin, François Pacaud, Mihai Anitescu
+%   \thanks{
+%     This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research (ASCR) under Contract DE-AC02-06CH11347.
+%   }
+%   \thanks{
+%     Sungho Shin and Mihai Anitescu are with Mathematics and Computer Science Division,  Argonne National Laboratory, Lemont, IL, USA. e-mail: \url{sshin@anl.gov}, \url{anitescu@mcs.anl.gov}
+%   }
+%   \thanks{
+%     François Pacaud is with Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France. e-mail: \url{francois.pacaud@minesparis.psl.eu}
+%   }
+% }
+
+\hyphenation{op-tical net-works semi-conduc-tor}
+
+% % Set footer
+% \makeatletter
+% \let\old@ps@headings\ps@headings
+% \let\old@ps@IEEEtitlepagestyle\ps@IEEEtitlepagestyle
+% \def\psccfooter#1{%
+%     \def\ps@headings{%
+%         \old@ps@headings%
+%         \def\@oddfoot{\strut\hfill#1\hfill\strut}%
+%         \def\@evenfoot{\strut\hfill#1\hfill\strut}%
+%     }%
+%     \def\ps@IEEEtitlepagestyle{%
+%         \old@ps@IEEEtitlepagestyle%
+%         \def\@oddfoot{\strut\hfill#1\hfill\strut}%
+%         \def\@evenfoot{\strut\hfill#1\hfill\strut}%
+%     }%
+%     \ps@headings%
+% }
+% \makeatother
+
+\usepackage{graphicx}
+\usepackage{enumitem}
+\usepackage{mathtools}
+\usepackage{tikz}
+\usepackage{lscape}
+\usepackage{multirow}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage{cuted}
 \usepackage{amsmath,amssymb,amsthm,amsfonts} % assumes amsmath package installed
 \usepackage[linktocpage=true,colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
-% \usepackage[letterpaper,margin=1in]{geometry}
-
-\newtheorem{definition}{Definition}
-\newtheorem{assumption}{Assumption}
-\newtheorem{theorem}{Theorem}
-\newtheorem{conjecture}{Conjecture}
-\newtheorem{lemma}{Lemma}
-\newtheorem{proposition}{Proposition}
-\newtheorem{remark}{Remark}
+
+% \newtheorem{definition}{Definition}
+% \newtheorem{assumption}{Assumption}
+% \newtheorem{theorem}{Theorem}
+% \newtheorem{conjecture}{Conjecture}
+% \newtheorem{lemma}{Lemma}
+% \newtheorem{proposition}{Proposition}
+% \newtheorem{remark}{Remark}
 
@@
 \sloppy
-\usepackage{graphicx}
-
-\usepackage{enumitem}
-\usepackage{mathtools}
 \allowdisplaybreaks
@@
 
-\usepackage{tikz}
-\usepackage{lscape}
-\usepackage{multirow}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\usepackage{cuted}
 \setlength\stripsep{3pt plus 1pt minus 1pt}
@@
 
-\date{\small
-  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
-  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
-}
 \begin{document}
 \maketitle
+
 \begin{abstract}
-  This paper introduces a novel computational framework for solving
-alternating current optimal power flow (ACOPF) problems using
-graphics processing units (GPUs). While GPUs have demonstrated
-remarkable performance in various computing domains, their application
-in AC OPF has been limited due to challenges associated with porting
-sparse automatic differentiation (AD) and sparse linear solver
-routines to GPUs. We aim to address these issues with two key
-strategies. First, we utilize a single-instruction, multiple-data
-(SIMD) abstraction of nonlinear programs (NLP). This approach enables
-the specification of model equations while preserving their
-parallelizable structure, and in turn, facilitates the implementation
-of AD routines that can exploit such structure. Second, we employ a
-condensed-space interior-point method (IPM) with an inequality
-relaxation strategy. This technique involves relaxing equality
-constraints to inequalities and condensing the Karush-Kuhn-Tucker
-system into a much smaller positive definite system. This strategy
-offers the key advantage of being able to factorize the KKT matrix
-without numerical pivoting, which in the past has hampered the parallelization of
-the IPM algorithm. By combining these two strategies, we can perform the
-majority of operations on GPUs while keeping the data residing in the
-device memory only. Comprehensive numerical benchmark results showcase
-the substantial computational advantage of our approach. Remarkably,
-for solving large-scale AC OPF problems to a moderate accuracy, our
-implementations—MadNLP.jl and ExaModels.jl—running on NVIDIA GPUs
-achieve an order of magnitude speedup compared to state-of-the-art
-tools running on contemporary CPUs.
+  This paper introduces a framework for solving alternating current optimal power flow (ACOPF) problems using graphics processing units (GPUs).
+While GPUs have demonstrated remarkable performance in various computing domains, their application in ACOPF has been limited due to challenges associated with porting sparse automatic differentiation (AD) and sparse linear solver routines to GPUs.
+We  address these issues with two key strategies.
+First, we utilize a single-instruction, multiple-data  abstraction of nonlinear programs.
+This approach enables the specification of model equations while preserving their parallelizable structure and, in turn, facilitates the parallel AD implementation.
+Second, we employ a condensed-space interior-point method (IPM) with an inequality relaxation.
+This technique involves condensing the Karush--Kuhn--Tucker (KKT) system into a positive definite system.
+This strategy offers the key advantage of being able to factorize the KKT matrix without numerical pivoting, which has hampered the parallelization of the IPM algorithm.
+By combining these strategies, we can perform the majority of operations on GPUs while keeping the data residing in the device memory only.
+Comprehensive numerical benchmark results showcase the advantage of our approach.
+Remarkably, our implementations—MadNLP.jl and ExaModels.jl—running on NVIDIA GPUs achieve an order of magnitude speedup compared with state-of-the-art tools running on contemporary CPUs.
 \end{abstract}
@@
 \end{IEEEkeywords}
+
 \section{Introduction}
 
-The adoption of GPUs in the mathematical programming community has remained
+The adoption of graphics processing units (GPUs) in the mathematical programming community has remained
 limited. Notably, nonlinear programming (NLP) remains dependent on
@@
 sluggish data transfer between host and device memory hampers the
-ad-hoc implementation of GPU accelerations (Fig. \ref{fig:memory}). To
-fully leverage the potential offered by modern GPU hardware, it
-becomes imperative to have a comprehensive computational framework for
-optimization on GPUs. That is, we need an AD/algebraic
+ad hoc implementation of GPU accelerations. To
+fully leverage the potential offered by modern GPU hardware, a comprehensive computational framework for
+optimization on GPUs is imperative. That is, we need an AD/algebraic
 modeling framework, sparse linear solvers, and NLP solvers that can
@@
 
-\begin{figure}[t]
-  \centering
-  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
-
-    \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
-    \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
-    \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};
-
-    \node[align=center] at (1.25, 3.75) {Data};
-    \node[align=center] at (6.75, 3.75) {Data};
-
-    % Host Memory
-    \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};
-
-    % Device Memory
-    \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};
-
-    % Arrow with Dashed Line
-    \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
-    \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
-    \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;
-
-    \def\rows{10}
-    \def\cols{10}
-    \def\elementwidth{2}
-    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
-
-      % Draw Grid
-      \foreach \i in {0,...,\rows} {
-        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
-      }
-      \foreach \i in {0,...,\cols} {
-        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
-      }
-    }
-  \end{tikzpicture}
-  \caption{A schematic description of host (CPU) and device (GPU) memory.}\label{fig:memory}
-\end{figure}
+% \begin{figure}[t]
+%   \centering
+%   \begin{tikzpicture}[remember picture, scale=.9, font=\small]
+
+%     \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
+%     \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
+%     \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};
+
+%     \node[align=center] at (1.25, 3.75) {Data};
+%     \node[align=center] at (6.75, 3.75) {Data};
+
+%     % Host Memory
+%     \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};
+
+%     % Device Memory
+%     \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};
+
+%     % Arrow with Dashed Line
+%     \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
+%     \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
+%     \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;
+
+%     \def\rows{10}
+%     \def\cols{10}
+%     \def\elementwidth{2}
+%     \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
+
+%       % Draw Grid
+%       \foreach \i in {0,...,\rows} {
+%         \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
+%       }
+%       \foreach \i in {0,...,\cols} {
+%         \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
+%       }
+%     }
+%   \end{tikzpicture}
+%   \caption{A schematic description of host (CPU) and device (GPU) memory.}\label{fig:memory}
+%   \vspace{-.15in}
+% \end{figure}
 
 This paper presents our approach to implementing a comprehensive
-a computational framework for solving large-scale AC OPF problems on
+computational framework for solving large-scale alternating current optimal power flow (ACOPF) problems on
 NVIDIA GPUs, along with the associated software implementations:
@@
 single-instruction, multiple-data (SIMD) abstraction of nonlinear
-programs (NLPs), enabling streamlined parallel AD on GPUs, and (ii) a
+programs, enabling streamlined parallel AD on GPUs, and (ii) a
 condensed-space interior-point method (IPM) with an inequality
@@
 While derivative evaluation can be generally cheaper than linear
-algebra operations, our numerical results on AC OPF problems show that
+algebra operations, our numerical results on ACOPF problems show that
 AD often constitutes more than half of the total solver time when
-using off-the-shelf AD implementations like
+using off-the-shelf AD implementations such as
 JuMP.jl~\cite{dunning2017jump} or
@@
 specialized AD implementation based on the SIMD abstraction of
-NLPs. This abstraction allows to preserve the parallelizable structure
+NLPs. This abstraction allows us to preserve the parallelizable structure
 within the model equations, facilitating efficient derivative
 evaluations on the GPU. The AC power flow model is particularly
-well-suited for this abstraction as it involves repetitive expressions
-for each component type (e.g., buses, lines, generators), and the
+well suited for this abstraction because it involves repetitive expressions
+for each component type (e.g., buses, lines, generators) and the
 the number of computational patterns does not increase with the network's
@@
 proposed strategies can achieve over 20 times speedup by running on
-GPU. In comparison to general AD implementations on CPUs (such as AMPL
+the GPU. Compared with general AD implementations on CPUs (such as AMPL
 and JuMP.jl), our GPU-based differentiation method can be
@@
 factorization, are typically the bottleneck in NLP solution methods.
-Parallelizing this operation has been considered to be challenging,
-primarily due to the need for numerical pivoting, which requires irregular
+Parallelizing this operation has been considered challenging,
+primarily because of the need for numerical pivoting, which requires irregular
 memory accesses and data movement \cite{swirydowicz2022linear}. However, when the matrix can be factorized without
@@
 numerical pivoting. This strategy relaxes equality constraints by
-permitting small violations, which enables expressing the
-Karush-Kuhn-Tucker (KKT) system entirely in the primal space through
+permitting small violations that enable expressing the
+Karush--Kuhn--Tucker (KKT) system entirely in the primal space through
 the condensation procedure. Although this strategy is not new
 \cite{nocedal2006numerical}, it has traditionally been considered less
-efficient than the standard full-space method due to increased nonzero
-entries in the KKT system. However, when implemented on GPUs, it
+efficient than the standard full-space method because of increased nonzero
+entries in the KKT system. When implemented on GPUs, however, it
 offers the key advantage of ensuring positive definiteness in the
@@
 Although this method is susceptible to numerical stability issues due to
-increased condition number in the KKT system, our results demonstrate
+the increased condition number in the KKT system, our results demonstrate
 that the solver is robust enough to solve problems with a relative
@@
 We present numerical benchmark results to showcase the efficiency of
-our method, utilizing our two packages: MadNLP.jl and
-ExaModels.jl. The solution of the KKT system is performed using the
+our method, utilizing our two packages MadNLP.jl and
+ExaModels.jl. The  KKT system is solved using the
 external cuSOLVER library. To assess the performance of our method, we
-compare it against standard CPU approaches using the data available in
+compare it with standard CPU approaches using the data available in
 pglib-opf \cite{babaeinejadsarookolaee2019power}.  Our benchmark results
 demonstrate that our proposed computational framework has significant
-potential for accelerating the solution of AC OPF problems, especially
+potential for accelerating the solution of ACOPF problems, especially
 when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
 Notably, when running on NVIDIA GPUs, our method achieves a
-4x speedup compared to our solver running on CPU for
+4x speedup compared with our solver running on CPUs for
 the largest instance. Moreover, for the same instance, our approach
@@
 We present, for the first time, a sparse nonlinear optimization
-solution framework that can run entirely on GPU, with all the
+solution framework that can run entirely on GPUs, with all the
 performance-critical data arrays residing exclusively on
@@
 thus allowing the solution of sparse, large-scale NLPs (with a
-prominent example being AC OPFs) on GPUs.
+prominent example being ACOPFs) on GPUs.
 
@@
 introduction of the sparse condensed-space IPM with an inequality relaxation
-strategy, the authors have investigated the use of reduction
+strategy, the authors  investigated the use of reduction
 strategies (state variable elimination) to treat KKT systems in a
@@
 batched with batched TRON solver \cite{lin1999newton}
-has been investigated \cite{kim2022accelerated,kim2021leveraging}.
-An NLP solver for high-performance computers (HPC) with GPU
-accelerators called HiOP has been under development
-\cite{hiop_techrep}, with a similar scope as our solver MadNLP.jl.
+have been investigated \cite{kim2022accelerated,kim2021leveraging}.
+An NLP solver for high-performance computers  with GPU
+accelerators, called HiOP, has been under development
+\cite{hiop_techrep}, with a  scope similar to that of our solver MadNLP.jl.
  Another recent development is the hybrid (direct-iterative) KKT
@@
 The implementation of derivative evaluations with the exploitation of
-repeated structures within model equations has, to the best of our
-knowledge, been first introduced in Gravity \cite{Gravity}. This was
+repeated structures within model equations was, to the best of our
+knowledge,  first introduced in Gravity \cite{Gravity}. This was
 achieved through the introduction of so-called template constraints,
-and multi-threaded derivate evaluation has been implemented therein.
-However, it is important to note that their differentiation approach
+and multithreaded derivative evaluation has been implemented therein.
+We note, however,  that their differentiation approach
 is based on symbolic differentiation.
-The idea of condensed-space IPM (without inequlaity relaxation
-strategy) is not new, but they have been used primarily in more
+The idea of condensed-space IPM (without inequality relaxation
+strategy) is not new, but it has been used primarily in more
 specific contexts, where the increased nonzero entries in the KKT system
-is less of a concern, such as in the dense form model predictive control
+are less of a concern, such as in the dense form model predictive control
 problems \cite{jerez2012sparse,cole2023exploiting}.
 
-\paragraph*{Organization}
-The paper is organized as follows. In the remainder of the current
-section, we introduce the mathematical notation. In Section
-\ref{sec:prelim}, we provide general preliminary knowledge on
-numerical optimization, automatic differentiation, and GPU computing.
-In Section \ref{sec:simd}, we present the SIMD abstraction of NLPs and
-their advantages in terms of implementing parallel AD. Section
-\ref{sec:ipm} presents the optimization algorithm under study, the
-condensed-space IPM with an inequality relaxation strategy. Section
-\ref{sec:num} presents the numerical results, comparing our approach
-with other state-of-the-art solution methods on the CPUs. Finally,
-conclusions and future outlooks are given in Section \ref{sec:conc}.
+% \paragraph*{Organization}
+% The paper is organized as follows. In the remainder of the current
+% section, we introduce the mathematical notation. In Section
+% \ref{sec:prelim}, we provide general preliminary knowledge on
+% numerical optimization, automatic differentiation, and GPU computing.
+% In Section \ref{sec:simd}, we present the SIMD abstraction of NLPs and
+% their advantages in terms of implementing parallel AD. Section
+% \ref{sec:ipm} presents the optimization algorithm under study, the
+% condensed-space IPM with an inequality relaxation strategy. Section
+% \ref{sec:num} presents the numerical results, comparing our approach
+% with other state-of-the-art solution methods on the CPUs. Finally,
+% conclusions and future outlooks are given in Section \ref{sec:conc}.
 
@@
 We denote the set of real numbers and the set of integers by
-$\mathbb{R}$ and $\mathbb{I}$. We let $[M]:=\{1,2,\cdots,M\}$. We let
+$\mathbb{R}$ and $\mathbb{I}$, respectively. We let $[M]:=\{1,2,\cdots,M\}$. We let
 $[v_i]_{i\in[M]}:=[v_1;v_2;\cdots,v_M]$.  A vector of ones with an
@@
 with an appropriate size is denoted by $I$. For matrices $A$ and $B$,
-$A\succ(\succeq) B$ indicates that $A-B$ is positive (semi)-definite
-while $A>(\geq) B$ denotes a component-wise inequality. We use the
+$A\succ(\succeq) B$ indicates that $A-B$ is positive (semi-)definite
+while $A>(\geq) B$ denotes a componentwise inequality. We use the
 convention $X:=\diag(x)$ for any symbol $x$.
@@
 
-This section covers three essential background topics: numerical
-optimization, AD, and GPU computing.\subsection{Numerical Optimization}\label{sec:numopt}
+\subsection{Numerical Optimization}\label{sec:numopt}
 We consider NLPs of the following form:
@@
     \min_{x^\flat \leq x \leq x^\sharp}\;  f(x) \quad \st\;
-     g(x) =0 \, .
+     g(x) =0 .
 \end{equation}
@@
 constraints, the NLP solution algorithms can be broadly classified
-into active-set methods and interior-point methods (IPMs) \cite{nocedal2006numerical}.
+into active-set methods and interior-point methods \cite{nocedal2006numerical}.
 Active-set methods aim to find the set of active constraints associated
@@
 orchestrates the operation of these tools to drive the solution
-iterate towards the stationary point of the optimization problem.
+iterate toward the stationary point of the optimization problem.
 
@@
 reference for the IPM implementation. Below, we outline the overall
-computational procedure employed within the NLP solution frameworks:
-
-1) Given the current primal-dual iterate $(x^{(\ell)},y^{(\ell)},
+computational procedure employed within the NLP solution frameworks.
+
+(1) Given the current primal-dual iterate $(x^{(\ell)},y^{(\ell)},
 z^{\flat(\ell)},z^{\sharp(\ell)})$, the AD package evaluates the first-
@@
 
-2) The following sparse indefinite system (known as the KKT system) is
+(2) The following sparse indefinite system (known as the KKT system) is
 solved using sparse indefinite factorization (typically, via sparse
-LBL$^\top$ factorization) and triangular solve routines.
+LBL$^\top$ factorization) and triangular solve routines:
 \begin{align}\label{eqn:kkt-indefinite}
@@
 
-3) The optimization solver employs a filter line search procedure to
+(3) The optimization solver employs a filter line search procedure to
 determine the step size~\cite{wachter2006implementation}. The primal-dual iterate is updated by
 applying the determined step size and direction. This process is
-repeated until the satisfaction of the convergence criteria (typically based on the residual
+repeated until  satisfaction of the convergence criteria (typically based on the residual
 to the first-order optimality conditions).
 
-\subsection{Automatic Differentiation (AD)}
+\subsection{Automatic Differentiation}
 Numerical differentiation of computer programs can be achieved through
-three different methods: (i) finite difference method, (ii) symbolic
-differentiation, and (iii) AD. The finite difference method suffers
+three different methods:  finite difference,  symbolic
+differentiation, and  AD. The finite difference method suffers
 from numerical rounding errors, and its computational complexity grows
@@
 derivatives. While this method can differentiate functions up to high
-numerical precision, it suffers from "expression swelling" effect and
+numerical precision, it suffers from "expression swelling"  and
 struggles to compute the derivatives of long nested expressions in a
@@
 scientific computing domain, including NLP and
-machine learning. For large-scale optimization problems, such as AC
-OPFs, AD tools are often implemented as part of domain-specific
+machine learning. For large-scale optimization problems, such as ACOPFs, AD tools are often implemented as part of domain-specific
 modeling languages. Examples of such modeling languages include JuMP,
-CasADi, and AMPL (optimization) and Tensorflow, Torch and, Flux
+CasADi, and AMPL (optimization) and TensorFlow, Torch, and Flux
 (machine learning).
 
-There are two alternative ways of propagating derivatives through the
-recursive application of chain rules: (i) forward-mode and (ii)
-reverse-mode, which operate in opposite directions (respectively, from leaves to
+ Derivatives can be propagated through the
+recursive application of chain rules in two ways, forward mode and 
+reverse mode, which operate in opposite directions (respectively, from leaves to
 root and from root to leaves). Reverse-mode automatic
@@
 
-The Julia Language, our language of choice, offers convenient and
+The Julia language, our language of choice, offers convenient and
 efficient ways to implement automatic differentiation. Through the use
 of the multiple dispatch paradigm~\cite{bezanson2017julia} \textit{any
-Julia function}---including commonly used operations like addition,
-multiplication, trigonometric and exponential functions (among
-others)---can be easily overloaded. Multiple dispatch allows functions
-to be dynamically dispatched based on the run-time type, a crucial
+Julia function}---including commonly used operations such as addition,
+multiplication, trigonometric and exponential functions---can be easily overloaded. Multiple dispatch allows functions
+to be dynamically dispatched based on the runtime type, a crucial
 feature for implementing differentiable programming. Several AD
-implementations have been developed in Julia Language, such as
+implementations have been developed in the Julia language, such as
 ReverseDiff.jl, ForwardDiff.jl, Zygote.jl, and JuMP.jl. While these
-tools are general and useful for various applications, they are not
-optimized for evaluating derivatives of AC OPF problems, as they are
+tools are general and useful for various applications, however, they are not
+optimized for evaluating derivatives of ACOPF problems, since they are
 not designed to exploit the parallelizable structures in the model,
@@
 \subsection{GPU Computing}\label{sec:gpu}
-With the increasing prevalence of GPU in various scientific computing
+With the increasing prevalence of GPUs in various scientific computing
 domains, there has been growing interest in leveraging these emerging
-architectures to efficiently solve large-scale NLPs, like AC OPF problems.
+architectures to efficiently solve large-scale NLPs, such as ACOPF problems.
 However, adapting an NLP solution algorithm,
@@
 same instruction simultaneously on hundreds of threads using the SIMD
-paradigm (see Fig. \ref{fig:simd}). The SIMD parallelism works well
+paradigm. SIMD parallelism works well
 for algorithms that can be decomposed into simple instructions running
@@
 paradigm. For example, branching in the control flow can hinder
-lockstep execution across multiple threads, and in turn, prevent
+lockstep execution across multiple threads and, in turn, prevent
 efficient implementations on GPUs. On the contrary, when the
@@
 they still can be significantly
-faster than operations on CPUs, as substantial part of the operation
+faster than operations on CPUs, since a substantial part of the operation
 can still be parallelized by the use of buffers. In simple cases, the
 implementation of these operations can be performed with the standard
-{\tt map} and {\tt mapreduce} programming models. However, in more
-complex cases, especially for \ref{eqn:pattern-3}, the implementation
-may require pre-inspection of memory write-access patterns and the use
+{\tt map} and {\tt mapreduce} programming models. In more
+complex cases, however, especially for \ref{eqn:pattern-3}, the implementation
+may require preinspection of memory write-access patterns and the use
 of custom kernels.
@@
 well as the application of optimization algorithms, are based on the
-computational patterns mentioned above. For example, an AC OPF model
+computational patterns mentioned above. For example, an ACOPF model
 can be implemented with 15 different computational patterns (see
@@
 aforementioned categories. Furthermore, the computation within
-optimization solvers, such as forming the left-hand-side for the KKT
+optimization solvers, such as forming the left-hand side for the KKT
 systems, computing the $\|\cdot\|_\infty$ norm of the constraint
@@
 Implementing kernel functions for the above computational patterns in
-the Julia Language is straightforward, as Julia provides excellent
+the Julia language is straightforward, since Julia provides excellent
 high-level interfaces for array and kernel programming for GPU
@@
 of the AD and optimization capabilities in our
-tools, MadNLP.jl and ExaModels.jl, are implemented in the Julia Language,
+tools MadNLP.jl and ExaModels.jl are implemented in  Julia,
 by leveraging its kernel and array programming capabilities.
 
-\begin{figure}[t]
-  \centering
-  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
-    \node[draw,align=left] (I) at (0,2.75) {
-      $\bullet$ Instruction $g(\cdot,\cdot)$\\
-      $\bullet$ Pointer to array $x$
-    };
-    \node[draw] at (-4,0) {
-      \tikz{
-        \foreach\x in {1,2,3,4,5,7}{
-          \node (A\x) at (0,-\x/2) {$s_{\x}$};
-        }
-        \node at (0,-2.9) {$\vdots$};
-      }
-    };
-    \node[draw] at (0,0) {
-      \tikz{
-        \foreach\x in {1,2,3,4,5,7}{
-          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
-        }
-        \node at (0,-2.9) {$\vdots$};
-      }
-    };
-    \node[draw] at (4,0) {
-      \tikz{
-        \foreach\x in {1,2,3,4,5,7}{
-          \node (C\x) at (0,-\x/2) {$y_{\x}$};
-        }
-        \node at (0,-2.9) {$\vdots$};
-      }
-    };
-    \foreach\x in {1,2,3,4,5,7}{
-      \draw[->] (A\x.east) -- (B\x.west);
-      \draw[->] (B\x.east) -- (C\x.west);
-      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
-    }
-    \node at (-4,-2.5) {Input Data};
-    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
-    \node at (4,-2.5) {Output Data};
-  \end{tikzpicture}
-  \label{fig:simd}
-  \caption{A schematic description of SIMD parallelism}
-\end{figure}
+% \begin{figure}[t]
+%   \centering
+%   \begin{tikzpicture}[remember picture, scale=.9, font=\small]
+%     \node[draw,align=left] (I) at (0,2.75) {
+%       $\bullet$ Instruction $g(\cdot,\cdot)$\\
+%       $\bullet$ Pointer to array $x$
+%     };
+%     \node[draw] at (-4,0) {
+%       \tikz{
+%         \foreach\x in {1,2,3,4,5,7}{
+%           \node (A\x) at (0,-\x/2) {$s_{\x}$};
+%         }
+%         \node at (0,-2.9) {$\vdots$};
+%       }
+%     };
+%     \node[draw] at (0,0) {
+%       \tikz{
+%         \foreach\x in {1,2,3,4,5,7}{
+%           \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
+%         }
+%         \node at (0,-2.9) {$\vdots$};
+%       }
+%     };
+%     \node[draw] at (4,0) {
+%       \tikz{
+%         \foreach\x in {1,2,3,4,5,7}{
+%           \node (C\x) at (0,-\x/2) {$y_{\x}$};
+%         }
+%         \node at (0,-2.9) {$\vdots$};
+%       }
+%     };
+%     \foreach\x in {1,2,3,4,5,7}{
+%       \draw[->] (A\x.east) -- (B\x.west);
+%       \draw[->] (B\x.east) -- (C\x.west);
+%       \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
+%     }
+%     \node at (-4,-2.5) {Input Data};
+%     \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
+%     \node at (4,-2.5) {Output Data};
+%   \end{tikzpicture}
+%   \label{fig:simd}
+%   \caption{A schematic description of SIMD parallelism}
+%   \vspace{-.15in}
+% \end{figure}
 
@@
 Accordingly, the evaluation and differentiation of the
-model equations in \eqref{eqn:prob} is amenable to SIMD parallelism.
+model equations in \eqref{eqn:prob} are amenable to SIMD parallelism.
 
 To implement the SIMD abstraction in the modeling environment, the
-algebraic modeling interface in ExaModels.jl enforces the users to
-specify the model equations in an {\tt Generator} data type in
-Julia Language. This composite data type consists of an instruction (a Julia
+algebraic modeling interface in ExaModels.jl requires the users to
+specify the model equations in an {\tt Generator} data type in the
+Julia language. This composite data type consists of an instruction (a Julia
 function) and data (a host or device array) over which the instruction
 is executed. This naturally facilitates maintaining the NLP model
-information in the form of SIMD abstraction in \eqref{eqn:prob}, and
+information in the form of SIMD abstraction in \eqref{eqn:prob} and
 facilitates the model evaluation and differentiation on GPU
@@
 \label{sec:simd:ad}
-Many physics-based models, such as AC OPF, have a highly repetitive
+Many physics-based models, such as ACOPF, have a highly repetitive
 structure. One of the manifestations of it is that the mathematical
@@
 AC OPF model. These patterns arise from (1) generation cost, (2) reference
-bus voltage angle constraint, (3-6) active and reactive power flow (from and to),
-(7) voltage angle difference constraint, (8-9) apparent
-power flow limits (from and to), (10-11) power balance equations,
-(12-13) generators' contributions to the power balance equations, and
-(14-15) in/out flows contributions to the power balance
+bus voltage angle constraint, (3--6) active and reactive power flow (from and to),
+(7) voltage angle difference constraint, (8--9) apparent
+power flow limits (from and to), (10--11) power balance equations,
+(12--13) generators' contributions to the power balance equations, and
+(14--15) in/out flow contributions to the power balance
 equations. However, such repetitive structure is not well exploited in
 the standard NLP modeling paradigms. In fact, without the SIMD
-abstraction, it is difficult for the AD package to detect the
-parallelizable structure within the model, as it will require the full
+abstraction it is difficult for the AD package to detect the
+parallelizable structure within the model, because it will require the full
 inspection of the computational graph over all expressions.  By
 preserving the repetitive structures in the model, the repetitive
-structure can be directly available in AD implementation.
+structure can be directly available in the AD implementation.
 
@@
 highly efficient derivative computation code, specifically compiled
-for each computational pattern in the model. These derivative evaluation codes can be run over the data in various GPU array formats,
-and implemented via array and kernel programming in Julia Language. In
-turn, ExaModels.jl has the capability to efficiently evaluate first and
+for each computational pattern in the model. These derivative evaluation codes can be run over the data in various GPU array formats
+and implemented via array and kernel programming in the Julia Language. In
+turn, ExaModels.jl has the capability to efficiently evaluate first- and
 second-order derivatives using GPU accelerators.
@@
 problems, the initial sparsity analysis of nonlinear expressions can
-be expensive, as the sparsity should be analyzed for potentially
-millions of objective and constraints terms. However, oftentimes
+be expensive, since the sparsity should be analyzed for potentially
+millions of objective and constraints terms. Often, however, 
 these analyses are applied to the same computational patterns, and
@@
 obtained parameterized sparsity pattern is materialized once the data
-array is given. This process drastically saves computational costs
+array is given. This process saves significant computational costs
 for the sparsity analysis.
@@
     \min_{\big[\substack{x^\flat\\s^\flat}\big]\leq \big[\substack{x\\s}\big] \leq \big[\substack{x^\sharp\\s^\sharp}\big]}\;
-      f(x)\quad \st\;  g(x) - s = 0 \, .
+      f(x)\quad \st\;  g(x) - s = 0.
 \end{equation}
-In our implementation, we set $s^{\flat},s^\sharp$ as
+In our implementation we set $s^{\flat},s^\sharp$ as
 $-\epsilon_{\text{tol}}\boldsymbol{1}$ and $+\epsilon_{\text{tol}}\boldsymbol{1}$, respectively,
@@
 For condensed-space IPM, we cannot maintain the same level of precision
-due to the increased condition number of the KKT system.
+because of the increased condition number of the KKT system.
 We have found that setting $\epsilon_{\text{tol}}$ to be
@@
 
-The IPM replaces the equality and inequality-constrained problem in \eqref{eqn:relax} with an equality-constrained  barrier subproblem:
+The IPM replaces the equality- and inequality-constrained problem in \eqref{eqn:relax} with an equality-constrained  barrier subproblem:
 \begin{subequations}\label{eqn:barrier}
@@
 \eqref{eqn:relaxed}. The Newton step for solving the nonlinear
-equations in \eqref{eqn:first} can be computed by solving the system in Fig. \ref{fig:very-long-eqn}.
+equations in \eqref{eqn:first} can be computed by solving the system in \eqref{fig:very-long-eqn}.
 \begin{figure*}[t]
@@
   %   \hline
-  \begin{align*}
+  \vspace{-.2in}
+  \begin{align}
     \underbrace{
@@
     \end{bmatrix}
-  \end{align*}
-  \caption{Full KKT system}\label{fig:very-long-eqn} 
+    \label{fig:very-long-eqn} 
+  \end{align}
+  % \caption{Full KKT system}\label
+  \vspace{-.3in}
 \end{figure*}
 Here, we recall the definitions of $W^{(\ell)}$ and $A^{(\ell)}$ from Section \ref{sec:numopt}, and
-$p^{(\ell)}_x,\cdots p^{(\ell)}_{z_s^\sharp}$ are defined by the left-hand-sides of the equations in
-\eqref{eqn:first}. In the sequel, we shall drop the superscript
+$p^{(\ell)}_x,\cdots p^{(\ell)}_{z_s^\sharp}$ are defined by the left-hand sides of the equations in
+\eqref{eqn:first}. In the sequel, we will drop the superscript
 $(\cdot)^{(\ell)}$ for concise notation.
 
-Now, we observe that a significant portion of the system in Fig.
-\eqref{fig:very-long-eqn} can be eliminated by exploiting the block
+Now, we observe that a significant portion of the system in \eqref{fig:very-long-eqn} can be eliminated by exploiting the block
 structure, leading to an equivalent system stated in a smaller
@@
 the strict interior of the feasible set. This allows for eliminating
-the lower-right 4x4 block, resulting in:
+the lower-right 4x4 block, resulting in
 \begin{align}\label{eqn:very-long-reduced}
@@
 \end{align}
-where:
+where
 \begin{align*}
@@
 The key advantage of the inequality relaxation strategy is that it
-imposes additional structure to the augmented KKT system, allowing us to
+imposes additional structure on the augmented KKT system, allowing us to
 further reduce the dimension of the problem. In particular, the
@@
 \end{align*}
-and the dual and slack step directions can be recovered by:
+and the dual and slack step directions can be recovered by
 \begin{align}
@@
 Therefore, the only sparse matrix that needs to be factorized is the
-matrix in the left-hand-side of \eqref{eqn:very-long-condensed}, with
+matrix in the left-hand side of \eqref{eqn:very-long-condensed}, with
 dimension $n \times n$.  Although we call
@@
 $M_\text{cond}$ is not necessarily a dense matrix. In fact, in the
-case of AC OPF problems, $M_\text{cond}$ is still highly sparse, as
+case of ACOPF problems, $M_\text{cond}$ is still highly sparse, since
 $W$ and $A$ are graph-induced banded systems.  Thus, exploiting
@@
 GPUs is that the matrix in \eqref{eqn:very-long-condensed} is positive
-definite upon the application of standard inertia correction
+definite upon  application of the standard inertia correction
 method. Typically, to guarantee that the computed step direction is a
@@
 
-By Sylvester's law, we have:
+By Sylvester's law, we have
 \begin{align*}
@@
 Cholesky factorization or LU refactorization), which is significantly
-more amenable to parallel implementation than indefinite LBL$^\top$
+more amenable to parallel implementation than is indefinite LBL$^\top$
 factorization (standard in the state-of-the-art IPM
@@
 an ill-conditioned KKT system in \eqref{eqn:very-long-condensed}.
-As a consequence, a single triangular solve may not provide a
+Consequently, a single triangular solve may not provide a
 sufficiently accurate step direction. Accordingly, iterative
@@
 
-The step size can be determined using the line search procedure.
-Although there are numerous alternative approaches, we follow the
+The step size can be determined by using the line search procedure.
+Although  numerous alternative approaches exist, we follow the
 filter line search method implemented in the Ipopt solver. The line
@@
 progress conditions is satisfied and acceptable by the
-filter. Furthermore, to enhance the convergence behavior, various
+filter. Furthermore, in order to enhance the convergence behavior, various
 additional strategies are implemented, such as the second-order
@@
 $\text{residual}(\cdot)$ is a scaled version of the residual to the
-first-order conditions in \eqref{eqn:first}. Finally, we summarize our
+first-order conditions in \eqref{eqn:first}. We summarize our
 condensed-space IPM in Algorithm \ref{alg:con-ipm}.
@@
     \STATE Solve the condensed KKT system \eqref{eqn:very-long-condensed} with $\delta_w=\delta_c=0$ to compute the primal step $\Delta x$ and recover the dual steps $\Delta y, \Delta z_x^\flat, \Delta z_x^\sharp, \Delta z_s^\flat, \Delta z_s^\sharp$ by \eqref{eqn:recover-1} and \eqref{eqn:recover-2}.
-    \STATE Determine the need for regularization and if necessary, recompute the step directions with proper choices of $\delta_w,\delta_c>0$.
+    \STATE Determine the need for regularization and, if necessary, recompute the step directions with proper choices of $\delta_w,\delta_c>0$.
     \STATE Choose step sizes $\alpha,\alpha_z>0$ via line search.
@@
 operations equivalent to Ipopt to different KKT data structures, such as
-{\tt SparseKKTSystem}, {\tt DenseKKTSystem}, {\tt
-DenseCondensedKKTSystem}, etc, whose data are stored either on the host or device memory. For the implementation of the condensed-space IPM
+{\tt SparseKKTSystem}, {\tt DenseKKTSystem}, and {\tt
+DenseCondensedKKTSystem},  whose data are stored either on the host or in device memory. For the implementation of the condensed-space IPM
 presented in this paper, we have added a new type of KKT system called
@@
 method against state-of-the-art methods on CPUs for solving standard
-AC OPF problems.
+ACOPF problems.
 
@@
 \ref{config-2}, MadNLP.jl applies IPM based on the indefinite,
-non-condensed KKT system, as in \eqref{eqn:kkt-indefinite}. In
+noncondensed KKT system, as in \eqref{eqn:kkt-indefinite}. In
 \ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
-system. The initial symbolic factorization is performed on the CPU using the KLU
+system. The initial symbolic factorization is performed on the CPU by using the KLU
 package \cite{davis2010algorithm}, and the subsequent numerical
@@
 fixed pivot sequence from KLU.  Software and hardware details of each
-configuration are illustrated in Table \ref{tbl:settings}. The AC OPF
-problem is formulated using the model from the rosetta-opf project
+configuration are illustrated in Table \ref{tbl:settings}. The ACOPF
+problem is formulated by using the model from the rosetta-opf project
 \cite{rosetta-opf}, which is based on the models in PowerModels.jl \cite{8442948}, and the test cases are obtained from the pglib-opf
 repository \cite{babaeinejadsarookolaee2019power}. We have selected
-the goc and pegase cases, as they contain large-scale instances.  The
+the goc and pegase cases because they contain large-scale instances.  The
 external packages are called from Julia, through thin wrapper
@@
 constraint violation measured by $\|\cdot\|_\infty$) is shown in Table
-\ref{tbl:quality}. Fig. \ref{fig:speedup} visually represents the
+\ref{tbl:quality}. Figure \ref{fig:speedup} visually represents the
 speedup brought by GPUs, by comparing the timing results of
@@
   \label{tbl:settings}
+  \vspace{-.1in}
 \end{table*}
@@
 \begin{figure}[t]
-  \includegraphics[width=.45\textwidth]{speedup-sol.pdf}
+  \centering
+  \includegraphics[width=.4\textwidth]{speedup-sol.pdf}
+  \vspace{-.1in}
   \caption{Speedup achieved by using GPUs.}
   \label{fig:speedup}
+  \vspace{-.2in}
 \end{figure}
 
-\paragraph*{Convergence pattern}
-First, when comparing the solvers' performance in terms of the
-interior point method (IPM) iteration counts, MadNLP.jl is as
+\paragraph*{Convergence Pattern}
+When comparing the solvers' performance in terms of the
+IPM iteration counts, MadNLP.jl is as
 efficient as the state-of-the-art solver Ipopt. The IPM iteration
@@
 suggests that running mathematically equivalent operations on GPUs (by
-using shared code base in high-level abstractions) can yield a similar
+using a shared code base in high-level abstractions) can yield a similar
 degree of effectiveness in terms of IPM convergence.
@@
 linear solvers is only moderate, this has a high impact on the overall
-speedup, as linear solver time constitutes a significant portion of
+speedup because the linear solver time constitutes a significant portion of
 the total solution time. For large instances, approximately 4x speedup
-can be achieved, though this can vary depending on the instances. For
+can be achieved, although this can vary depending on the instances. For
 example, for case24464\_goc, the GPU linear solver was slower than the
@@
 operations is consistently greater than the speedup in linear
-solvers. However, due to the frequent use of \ref{eqn:pattern-2} and
-\ref{eqn:pattern-3} operations, the speedup in solver internal
-operations is lesser than that of derivative evaluations.
+solvers. Because of the frequent use of \ref{eqn:pattern-2} and
+\ref{eqn:pattern-3} operations,  however, the speedup in solver internal
+operations is less than that of derivative evaluations.
 
@@
 
-\paragraph*{Known Limiataions}
+\paragraph*{Known Limitations}
 The key limitation of our method is the decreased precision of the
@@
 for the augmented KKT system and $3.15\times 10^{14}$ for the
-condensed KKT system. As a consequence, the achievable precision of
+condensed KKT system. Consequently, the achievable precision of
 the solution is reduced. Further investigation into the possibility of
-obtaining higher precision will be necessary in the future.
+obtaining higher precision will be necessary.
 
@@
 framework that can run entirely on GPUs. Our method has demonstrated
-promising results, achieving a 4x speedup when compared to CPU
-implementations for large-scale AC OPF problems. Notably, our approach
+promising results, achieving a 4x speedup when compared with CPU
+implementations for large-scale ACOPF problems. Notably, our approach
 outperforms one of the state-of-the-art CPU-based implementations by
@@
 resulting in decreased final solution accuracy. Addressing the
-challenges posed by ill-conditioning remains to be an important aspect
+challenges posed by ill-conditioning remains an important aspect
 of future work. In the following paragraphs, we discuss some
@@
 While we have focused on the IPM, other constrained optimization
-paradigms, such as penalty methods and augmented Lagrangian methods
+paradigms, such as penalty methods and augmented Lagrangian methods,
 exist, and similar strategies based on condensed linear systems can be
 developed. It would be valuable to investigate which algorithm would
-be the right paradigm for constrained large-scale optimization on GPU
+be the right paradigm for constrained large-scale optimization on GPUs
 that can best handle the ill-conditioning issue of the condensed KKT
-system, and in turn, achieve the highest degree of accuracy.
-
-\paragraph*{Security-Constrained, Multi-Period, Distribution OPFs}
-Though the proposed method has demonstrated significant computational
-advantages for transmission AC OPF problems, our results can also
-interpreted that efficient CPUs can still handle these problems
+system and, in turn, achieve the highest degree of accuracy.
+
+\paragraph*{Security-Constrained, Multiperiod, Distribution OPFs}
+Although the proposed method has demonstrated significant computational
+advantages for transmission ACOPF problems, our results can also
+be interpreted that efficient CPUs can still handle these problems
 reasonably well. We anticipate that there will be more substantial
 performance gains for larger-scale optimization problems, such as
-security-constrained and multi-period OPFs or joint optimization
+security-constrained and multiperiod OPFs or joint optimization
 problems involving transmission, distribution, and gas network
@@
 for better performance due to lower computational complexity and the
-ability to reveal the inertia of the KKT system. In the future, we are
+ability to reveal the inertia of the KKT system. We are
 interested in exploring other linear solvers options, such as CHOLMOD
@@
 \paragraph*{Portability}
-Our implementation is currently only tested on NVIDIA GPUs, but our
+Our implementation is currently  tested only on NVIDIA GPUs, but our
 GPU implementation is largely based on array programming and
@@
 supporting a broader class of GPU accelerators in the future.
+
+\section*{Acknowledgment}
+This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research (ASCR) under Contract DE-AC02-06CH11347.
 

### 2023-10-27 slides/AIChE2023/main.tex
coin-or and AIChE talk

\input{preamble}
\bibliographystyle{habbrv}

\title{{\bfseries Nonlinear Optimization on Graphics Processing Units }}

\newcommand{\mytitlepage}{
  \begin{tikzpicture}[remember picture,overlay,align=center]
    \node[align=left, font=\footnotesize] at (current page.center) { 
      {\bf\Large\color{emph} Nonlinear Optimization on Graphics Processing Units}\\[3em]
      {\normalsize {\bf Sungho Shin}$^{*,1}$, {\bf Fran\c{c}ois Pacaud}$^{2}$, {\bf Mihai Anitescu}$^{1,3}$, and {\bf Exanauts}}\\[1em]
      $^*$\url{sshin@anl.gov}
      \\[1em]
      $^1$Mathematics and Computer Science Division, Argonne National Laboratory\\
      $^2$Centre Automatique et Systèmes, Mines Paris - PSL\\
      $^3$Department of Statistics, University of Chicago\\[1em]
      {\small 2023 AIChE Meeting, Orlando, Florida} \\[1em]
      \includegraphics[width=2cm]{shin.jpg}
      \includegraphics[width=2cm]{pacaud.jpg}
      \includegraphics[width=2cm]{anitescu.jpg}
    }
    \node[xshift=210,yshift=-10,anchor=east] at (current page.center) {
      \includegraphics[width=4cm]{a100.png}
    }
    \node[xshift=210,yshift=-70,anchor=east] at (current page.center) {
      \includegraphics[width=6cm]{aurora.jpg}
    } ;
  \end{tikzpicture}
}

\graphicspath{}

\author[Sungho Shin]{
  Sungho Shin$^{1}$, Fran\c{c}ois Pacaud$^{2}$, and Mihai Anitescu$^{1,3}$\\
  {\normalfont\footnotesize\url{sshin@anl.gov}}
}
\subtitle{}
\institute{
  $^1$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^2$Centre Automatique et Systèmes, Mines Paris - PSL\\
  $^3$Department of Statistics, University of Chicago
}
\date{
  2023 AIChE Annual Meeting\\
  Orlando, Florida
}

\begin{document}

\begin{frame}[noframenumbering,plain]
  \mytitlepage
\end{frame}

\begin{frame}{Accelerated Computing on GPUs}
  \begin{itemize}
  \item Accelerated computing has \textbf{driven the success of AI} (e.g., GPT models have $10^{12}$ pars).
  \item<2-> Accelerated computing \textbf{empowers scientific computing} (e.g., fluid, climate, bioinformatics).
  \item<3-> We're entering \textbf{exascale computing era} ($10^{18}$ floating point operations per second).
    \begin{center}
      \begin{tikzpicture}[align=center]
        \node[font=\small] at (0,0){
          \textbf{Aurora Supercomputer} @ Argonne (2024)\\[.5em]
          \includegraphics[height=1in]{aurora.jpg}
        };
        \node at (5.5,0){$=\quad$1 million $\times\;$};
        \node[font=\small] at (8,0){
          \textbf{iPhone 14 Pro} (2023)\\[.5em]
          \includegraphics[height=1in]{iphone-14-pro.png} 
        };
      \end{tikzpicture}
    \end{center}
  \end{itemize}
  \visible<4->{
    \begin{center}
      \tikz{
        \node[rounded corners, fill=black!10,text width=.85\textwidth,align=center,inner sep=5] at (0,0) {
          Can we harness these capabilities in the realm of \textbf{classical optimization}?\\
          (e.g., energy systems, optimal control, operations research)
        };
      }
    \end{center}
  }
\end{frame}

\begin{frame}{How Do GPUs Work? {\normalsize (or, how are they different from CPUs?)}}
  \begin{itemize}
  \item \visible<1->{Single Instruction, Multiple Data ({\bf SIMD}) parallelism,} \visible<2->{on {\bf dedicated memory}.} 
    \begin{center}
      \scalebox{.7}{
        \visible<1->{
          \begin{tikzpicture}[remember picture, scale=.9, font=\small]
            \node[draw,align=left] (I) at (0,2.75) {
              $\bullet$ Instruction $g(\cdot,\cdot)$\\
              $\bullet$ Pointer to array data
            };
            \node[draw] at (-4,0) {
              \tikz{
                \foreach\x in {1,2,3,4,5,7}{
                  \node (A\x) at (0,-\x/2) {$s_{\x}$};
                }
                \node at (0,-2.9) {$\vdots$};
              }
            };
            \node[draw] at (0,0) {
              \tikz{
                \foreach\x in {1,2,3,4,5,7}{
                  \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
                }
                \node at (0,-2.9) {$\vdots$};
              }
            };
            \node[draw] at (4,0) {
              \tikz{
                \foreach\x in {1,2,3,4,5,7}{
                  \node (C\x) at (0,-\x/2) {$y_{\x}$};
                }
                \node at (0,-2.9) {$\vdots$};
              }
            };
            \foreach\x in {1,2,3,4,5,7}{
              \draw[->] (A\x.east) -- (B\x.west);
              \draw[->] (B\x.east) -- (C\x.west);
              \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
            }
            \node at (-4,-2.5) {Input Data};
            \node[align=center] at (0,-2.5) {GPUs};
            \node at (4,-2.5) {Output Data};
          \end{tikzpicture}
        }
      }
      \scalebox{.75}{
        \visible<2->{
          \begin{tikzpicture}[remember picture, scale=.9, font=\small]

            \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
            \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
            \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};

            \node[align=center] at (1.25, 3.75) {Data};
            \node[align=center] at (6.75, 3.75) {Data};

            % Host Memory
            \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};

            % Device Memory
            \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};

            % Arrow with Dashed Line
            \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
            \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
            \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;

            \def\rows{10}
            \def\cols{10}
            \def\elementwidth{2}
            \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
              % Draw Grid
              \foreach \i in {0,...,\rows} {
                \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
              }
              \foreach \i in {0,...,\cols} {
                \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
              }
            }
          \end{tikzpicture}
        }
      }
    \end{center}
    % \item Particularly efficient for dense matrix-matrix multiplication.
  \item<3-> To achieve optimal performance, all data should {\bf reside exclusively on device memory}.
  \item<4-> Heterogeneous architectures due to {\bf various hardware vendors} (NVIDIA, AMD, and Intel).
  \end{itemize}
  \visible<5>{
    \begin{center}
      \tikz{
        \node[rounded corners, fill=black!10,text width=.95\textwidth,align=center,inner sep=5] at (0,0) {
          Adapting CPU algorithms into GPU algorithms is \textbf{not merely a matter of software engineering} and often requires a \textbf{complete redesign of the algorithm}.
        };
      }
    \end{center}
  }
\end{frame}

\begin{frame}{Exascale Computing Project}
  \begin{itemize}
  \item {\bf Mission}: Harness exascale computing capabilities to tackle {\bf real-world challenges}.
    \begin{center}
      \begin{tikzpicture}[font=\scriptsize,align=center]
        \node at (-4,0) {\includegraphics[width=2.5in]{national-labs.png}};
        \node at (2.25,.9) {
          {\bf Frontier} @Oak Ridge (AMD GPUs)\\
          \includegraphics[height=.6in]{frontier.png}
        };
        \node at (2.25,-.9) {
          {\bf Aurora} @Argonne (Intel GPUs)\\
          \includegraphics[height=.6in]{aurora.jpg}
        };
      \end{tikzpicture}
    \end{center}
  \item<2-> {\bf Challenge}: There were no implementations of {\bf sparse automatic differentiation, nonlinear optimization solver, nor sparse indefinite solvers} on GPUs.
  \item<3-> {\bf Goal}: Build a {\bf comprehensive software infrastructure} for nonlinear optimization on GPUs.
    \begin{itemize}
    \item {\bf Performance}: at least an order of magnitude speedup.
    \item {\bf Portability}: compatibility with NVIDIA, AMD, and Intel.
    \item {\bf Application}: energy infrastructure problems.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Language of Choice: Julia}
  \begin{itemize}
  \item Julia resolves the ``{\bf two-language problem}''; we're experiencing 10x faster development.
  \item<2-> {\bf Multiple dispatch}: High-level abstraction while specializing for specific data types.
  \item<3-> {\bf Portable programming}: Compatibility across various architectures (NVIDIA, AMD, Intel, etc.)
  \item<4-> {\bf JIT compilation}: No performance compromises, running as fast as C/C++/Fortran.
  \item<5-> The initial delay, often referred to as ``{\bf time-to-first-call},'' has been significantly reduced.
  \end{itemize}
  \begin{center} 
    \visible<4->{
      \includegraphics[height=1.75in]{speed-comparison.png}
    }
    \visible<5->{
      \includegraphics[height=1.75in]{TTFP.pdf}
    }
  \end{center}
\end{frame}

\begin{frame}{Summary of Exanauts' Journey}
  \begin{center}
    \begin{tikzpicture}[looseness=1,overlay, remember picture, xshift=-50]
      \draw[-latex,dashed] (-4,0) to (-4.5,-2) to (-1,-4) to (.5,2) to (-.5,0) to (-1.5,-3) to (-.5,2) to (-2.5,1) to (-3,0) to (0,0);
      \draw[-latex] (-4,0) -- (0,0);

      \draw[-latex] (-5,-3.6) -- ++ (.5,0) node[right,anchor=west,font=\scriptsize] {Ideal Research Progres};
      \draw[-latex,dashed] (-5,-4) -- ++(.5,0) node[right,anchor=west,font=\scriptsize] {Our Journey};
      \visible<2->{
        \node[align=left, text width=.5\textwidth] at (5.5,-.5) {
          \begin{itemize}
          \item {\bf Nonlinear optimization/linear algebra}:\\
            distributed Jacobi, Lagrangian decomposition, ADMM, batched TRON solver, domain decomposition preconditioners, 
            \only<2>{reduced-space interior point method, condensed interior-point methods w/ inequality relaxation}\only<3>{{\color{blue}reduced-space interior point method, condensed interior-point methods w/ inequality relaxation}}, ...
          \item {\bf Modeling/automatic differentiation}:\\
            batched second-order adjoint sensitivity,
            \only<2>{SIMD abstraction}\only<3>{{\color{blue}SIMD abstraction}}, ...
          \item {\bf Packages developed}:\\
            ExaPf, Argos, ExaADMM, BlockPowerFlow, ProxAL, ExaTron, \only<2>{MadNLP, ExaModels}\only<3>{{\color{blue}MadNLP, ExaModels}}, ...
          \end{itemize}
        };
      }
    \end{tikzpicture} 
  \end{center}
  \reference{
    M. Anitescu et. al. {\it Targeting Exascale with Julia on GPUs for multiperiod optimization with scenario constraints}. SIAG/OPT Views and News (2021).
  }
\end{frame}

\begin{frame}{Two Award-Winning Open-Source Projects {\small (2023 COIN-OR Cup)}}
  \begin{center}
    \begin{tikzpicture}[remember picture,overlay,align=center] 
      \node at (-3.75,1) {
        \includegraphics[width=2.75in]{examodels.pdf}
      };
      \node at (3.75,1) {
        \includegraphics[width=2.75in]{madnlp.pdf}
      };
      \node at (-4,-.5) {
        \footnotesize\url{https://github.com/sshin23/ExaModels.jl}
      }; 
      \node at (4,-.5) {
        \footnotesize\url{https://github.com/MadNLP/MadNLP.jl}
      };
      \node at (0, -2.5) {
        \includegraphics[height=2cm]{shin.jpg}
        \includegraphics[height=2cm]{pacaud.jpg}
        \includegraphics[height=2cm]{anitescu.jpg}
        \includegraphics[height=2cm]{coin-or.png} 
      };
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}{How \{AMPL, IPOPT, Ma27\} Have Succeeded on CPUs}
  \centering
  \begin{tikzpicture}[align=center,font=\footnotesize,overlay, remember picture] 
    \node[rounded corners, fill=black!10, text width=.225\textwidth, minimum height=.2\textheight] (A1) at (-5.7,2) {      
      \begin{align*}
        \min_{x\geq 0}\;& f(x)\\
        \st\;& c(x) = 0
      \end{align*}      
    };
    \node[rounded corners, fill=black!10, text width=.225\textwidth, minimum height=.2\textheight] (A2) at (-1.9,2) {      
      \begin{align*}
        \min_{x}\;& f(x) - \mu \log(x)\\
        \st\;& c(x) = 0
      \end{align*}

    };
    \node[rounded corners, fill=black!10, text width=.225\textwidth, minimum height=.2\textheight,font=\scriptsize] (A3) at (1.9,2) {      
      \begin{align*}
        \hspace{-.5em}
        \setlength\arraycolsep{2pt}
        \underbrace{\begin{bmatrix}
          W + \Sigma +\delta_w I& A^\top\\
          A & \delta_c I
        \end{bmatrix}
              \hspace{-.5em}
              \begin{bmatrix}
                \Delta x\\ \Delta \lambda
              \end{bmatrix}
        \hspace{-.25em} 
        =
        \hspace{-.25em}
        \begin{bmatrix}
          p^x\\p^\lambda
        \end{bmatrix}}_{\text{\scriptsize "KKT System"}}
      \end{align*}      
    };
    \node[rounded corners, fill=black!10, text width=.225\textwidth, minimum height=.2\textheight] (A4) at (5.7,2) {
      \begin{align*}
        x^{(k+1)}  = x^{(k)} + \alpha \Delta x\\
        \lambda^{(k+1)}  = \lambda^{(k)} + \alpha \Delta \lambda 
      \end{align*}      
    };
    \node[font=\footnotesize\bf] at (-5.7,2.65) {Modeling};
    \node[font=\footnotesize\bf] at (-1.9,2.65) {Barrier Reformulation};
    \node[font=\footnotesize\bf] at (1.9,2.65) {Newton's Step Comp.};
    \node[font=\footnotesize\bf] at (5.7,2.65) {Line Search};

    \only<2->{
      \node[font=\footnotesize,text width=.3\textwidth,fill=red!10,rounded corners] (B1) at (-5,-.5) {
        {\bf Algebraic Modeling Systems}\\[.5em]
        AMPL, JuMP, CasADi, ...
      };
    }
    \only<3->{
      \node[font=\footnotesize,text width=.3\textwidth,fill=blue!10,rounded corners] (B2) at (0,-.5) {
        {\bf Nonlinear Optimization Solvers}\\[.5em]
        Ipopt, Knitro, Pynumero, ...
      };
    }
    \only<4->{
      \node[font=\footnotesize,text width=.3\textwidth,fill=green!10,rounded corners] (B3) at (5,-.5) {
        {\bf Sparse Linear Solvers}\\[.5em]
        HSL (ma27, ma57, ...), Pardiso, ...
      }; 
    }
    \only<2->{
      \draw[red,latex-] (A1.south) -- (B1.north);
      \draw[red,latex-] (A3.south) -- (B1.north);
    }
    \only<3->{
      \draw[blue,latex-] (A2.south) -- (B2.north);
      \draw[blue,latex-] (A3.south) -- (B2.north);
      \draw[blue,latex-] (A4.south) -- (B2.north);
    }
    \only<4->{
      \draw[ForestGreen,latex-] (A3.south) -- (B3.north);
    }
    \node[text width=\textwidth,font=\normalsize] at (0,-2.5) {
      \begin{itemize}
      \item<2-> Algebraic modeling systems have enabled efficient {\bf sparse automatic differentiation}.
      \item<3-> Interior-point solvers have removed the combinatorial complexities of inequality constraints; however, we pay the price by solving {\bf extremely ill-conditioned linear systems}.
      \item<4-> Sparse indefinite solvers have enabled {\bf the solution of ill-conditioned linear systems}. 
      \end{itemize}
    };
  \end{tikzpicture}
\end{frame}

\begin{frame}{Why it is Challenging: Sparse Automatic Differentiation} 
  \begin{itemize}
  \item {\bf Automatic differentiation} is {\bf more efficient} than {\it finite difference} or {\it symbolic differentiation}, and is {\bf less prone to error} than {\it hand-written derivative codes}.
    \begin{center}
      \vspace{.2in}
      \begin{tikzpicture}[overlay, remember picture] 
        \node[rounded corners,fill=red!10,minimum width=1.5in] (1) at (-4,0) {\tt f(x)\{ ... \}};
        \node[rounded corners,fill=red!10,minimum width=1.5in] (2) at (4,0) {\tt df(x)\{ ... \}};
        \node[rounded corners,fill=blue!10,minimum width=1.5in] (3) at (-4,-2) {$y=f(x)$};
        \node[rounded corners,fill=blue!10,minimum width=1.5in] (4) at (4,-2) {$y'=f'(x)$};
        \only<5->{
          \draw[-latex,font=\footnotesize\it] (1) -- node[above] {Automatic Differentiation} node[below] {(e.g., AMPL, JuMP, Torch)} (2);
        }
        \only<3->{
          \draw[-latex,font=\footnotesize\it, dashed] (3) -- node[above] {Symbolic Differentiation} node [below] {(e.g., Mathematica, Maple)} (4);
          \draw[-latex,font=\footnotesize\it, dashed] (1) -- (3);
          \draw[latex-,font=\footnotesize\it, dashed] (2) -- (4);
        }
        \only<2->{
          \draw[latex-,font=\footnotesize\it] (1.south) -- ++ (-.75,-.75) node[anchor=east] {Finite Difference} ;
        }
        \only<4->{
          \draw[latex-,font=\footnotesize\it] (2.south) -- ++ (-.75,-.75) node[anchor=east] {Hand-Written Derivative Code};
        }
      \end{tikzpicture}
      \vspace{1.1in}
    \end{center}
  \item<6-> Existing {\bf algebraic modeling systems} (e.g., AMPL and JuMP) are {\bf designed for CPUs}.
  \item<7-> {\bf Machine learning frameworks} (e.g., Torch) are not effective for {\bf sparse problems}.
  \end{itemize}
\end{frame}

\begin{frame}{Solution: SIMD Abstraction}
  \begin{itemize}
  \item<1-> Large-scale optimization problems {\bf almost always have repetitive patterns}.\\
    e.g., optimal power flow w/ millions of variables can be expressed with a handful of patterns.
    \vspace{-.15in}
    \begin{columns}[t]
      \footnotesize
      \begin{column}{0\textwidth}
      \end{column}
      \begin{column}{.3\textwidth}
        \begin{itemize}
        \item referencebus voltage angle
          \vspace{-.05in}
        \item active and reactive power flow
        \end{itemize}
      \end{column}
      \begin{column}{.3\textwidth}        
        \begin{itemize}
        \item voltage angle difference limits
          \vspace{-.05in}
        \item apparent power flow limits
        \end{itemize}
      \end{column}
      \begin{column}{.3\textwidth}        
        \begin{itemize}
        \item generation cost
          \vspace{-.05in}
        \item power balance equations
        \end{itemize}
      \end{column}
    \end{columns}
    \visible<2->{
      \begin{align*}
        \min_{x^\flat\leq x \leq x^\sharp}
        & \sum_{l\in[L]}\sum_{i\in [I_l]} f^{(l)}(x; p^{(l)}_i)\tag{\bf SIMD abstraction}\\
        \st\;
        &
          \left[g^{(m)}(x; q_j)\right]_{j\in [J_m]} +\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) =0, \quad \forall m\in[M]
      \end{align*}
    }
    \vspace{-.1in} 
  \item<3-> Repeated patterns are made available by always specifying the models as {\bf iterable objects}.
    \begin{center}
      \begin{tikzpicture}
        \node[rounded corners, fill=black!90, font=\color{white}\small] at (0,0) {\tt
          constraint(c, 3 * x[i+1]\^{}3 + 2 * sin(x[i+2]) for i = 1:N-2)
        };
      \end{tikzpicture}
    \end{center} 
  \item<4-> {\bf For each repeatitive pattern}, the derivative evaluation kernel is constructed \& compiled,\\ and it is {\bf executed in parallel over multiple data}.
  \end{itemize}
  \reference{
    S. Shin, F. Pacaud, and M. Anitescu. {\it Accelerating optimal power flow with GPUs: SIMD abstraction of nonlinear programs and condensed-space interior-point methods}, arXiv:2307.16830.
  }
\end{frame}

\begin{frame}{Why it is Challenging: Sparse Linear Solvers}
  \begin{itemize}
  \item<1-> Solving KKT systems on CPUs has traditionally relied on {\bf direct LBL$^\top$ factorization}.
    \begin{center}
      \tikz{
        \node[inner sep=0pt] (augmented) at (-4,0) {\includegraphics[width=1.25in]{augmented.pdf}};
        \node[inner sep=0pt] (facL) at (0,0) {\includegraphics[width=1.25in]{lu_L.pdf}};
        \node[inner sep=0pt] (facU) at (6,0) {\includegraphics[width=1.25in]{lu_U.pdf}};
        \node[inner sep=0pt] at ($(facL)!.5!(augmented)$) {$=$};
        \node[inner sep=0pt] at ($(facL)!.35!(facU)$) {$\times$};
        \node[inner sep=0pt] at ($(facL)!.65!(facU)$) {$\times$};
        \node[align=center,inner sep=0pt,yshift=-12] at ($(facL)!.5!(facU)$) {$B$\\\footnotesize (1x1 or 2x2\\[-.25em]\footnotesize block-diagonal)};
      } 
    \end{center}
  \item<2-> Achieving numerical stability necessitates {\bf numerical pivoting}, a process that is notoriously {\bf challenging to parallelize}.
  \item<3-> Due to the ill-conditioning of the KKT system, {\bf iterative methods} are generally impractical, although there have been some promising results with {\bf specialized preconditioners}.
  \end{itemize}
  \visible<3->{
    \reference{
      Curtis, Huber, Schenk, Waechter. {\it A note on the implementation of an interior-point algorithm for nonlinear optimization with inexact step computations}. MathProg. (2012)\\
      Cao, Seth, Laird. {\it An augmented Lagrangian interior-point approach for large-scale NLP problems on graphics processing units}. CACE. (2016)\\
      Gondzio, Matrix-free interior point method, Comput Optim Appl (2012)
    }
  }
\end{frame}

\begin{frame}{Solution \#1: Condensed-Space Interior Point Method} 
  \begin{itemize}
  \item<1-> We aim to transform the KKT systems into a {\bf sparse positive definite system}.
  \item<2-> Can be achieved by (i) converting the {\bf equalities into inequalities}:
    \begin{align*}
      g(x) = 0 \quad\Longrightarrow\quad g(x)- s = 0,\quad s^{\flat}\leq s\leq  s^\sharp,
    \end{align*}
    \visible<3->{
      (ii) {\bf eliminating the slack variables} (so-called {\it condensation}):
      \begin{align*}\scriptsize
        \begin{bmatrix}
          W^{(\ell)}  + \delta^{(\ell)}_w I & & A^{(\ell)\top}& -I & I &  \\
          & \delta^{(\ell)}_w I & -I&&&-I & I\\
          A^{(\ell)}& -I & -\delta^{(\ell)}_c I\\
          Z_x^{(\ell)\flat}&&&X^{(\ell)}-X^\flat\\
          -Z_x^{(\ell)\sharp}&&&&X^\sharp-X^{(\ell)}\\
          &Z_s^{(\ell)\flat}&&&&S^{(\ell)}-S^\flat\\
          &-Z_s^{(\ell)\sharp}&&&&&S^\sharp-S^{(\ell)}\\
        \end{bmatrix}
        \begin{bmatrix}
          \Delta x \\
          \Delta s \\
          \Delta y \\
          \Delta z_x^\flat \\
          \Delta z_x^\sharp \\
          \Delta z_s^\flat \\
          \Delta z_s^\sharp \\
        \end{bmatrix} =
        \begin{bmatrix}
          p^{(\ell)}_{x }\\
          p^{(\ell)}_{s }\\
          p^{(\ell)}_{y }\\
          p^{(\ell)}_{z_x^\flat }\\
          p^{(\ell)}_{z_x^\sharp }\\
          p^{(\ell)}_{z_s^\flat }\\
          p^{(\ell)}_{z_s^\sharp }\\
        \end{bmatrix}\\
        \Longrightarrow\quad
        \color{blue}{(W + \delta_wI + \Sigma_x + A^{\top} D A ) \Delta x = q_x + A^\top (C q_s +  Dq_y )}
      \end{align*}
    }
    \vspace{-.2in}
  \item<4-> These systems can be solved {\bf without numerical pivoting} (routines available in CUDA).
  \end{itemize}
  \reference{
    S. Shin, F. Pacaud, and M. Anitescu. {\it Accelerating optimal power flow with GPUs: SIMD abstraction of nonlinear programs and condensed-space interior-point methods}, arXiv:2307.16830.
  }
\end{frame}

\begin{frame}{Solution \#2: Reduction \& Dense Reformulation} 
  \begin{itemize}
  \item<1-> Oftentimes, the decision variables can be {\bf splitted into state} and {\bf control variables}.
  \item<2-> We {\bf reduce the KKT systems} by {\bf elimintating the state variables} [Biegler, 1995].
  \item<3-> This {\bf reduction} transforms the KKT system into a compact, dense, positive definite system.
    \begin{center}
      \begin{tikzpicture}[font=\footnotesize\bf]
        \node at (0,0) {\includegraphics[width=.7\textwidth]{reduction.png}};
        \node at (-3.7,2) {Original};
        \node at (0,2) {Condensed};
        \node at (3.7,2) {Condensed \& Reduced};
      \end{tikzpicture}      
    \end{center}
  \item<4-> {\bf The elimination is a bottleneck}, but can be accelerated by exploiting structures\\
    (e.g., Riccati recursion for optimal control).
  \end{itemize}
  \reference{
    L. Biegler, J. Nocedal, C. Schmid. {\it A reduced Hessian method for large-scale constrained optimization}. SIOPT (1995).\\
    F. Pacaud, S. Shin, M. Schanen, D. A. Maldonado, and M. Anitescu. {\it Accelerating condensed interior-point methods on SIMD/GPU architectures}. JOTA (2023). \\
    D. Cole, S. Shin, F. Pacaud, V. M. Zavala, and M. Anitescu. {\it Exploiting GPU/SIMD architectures for solving linear-quadratic MPC problems}. ACC (2023).
  }
\end{frame}

\begin{frame}{Highlight: AC Optimal Power Flow {\normalsize (single GPU)}}
  \begin{itemize}
  \item<1-> Standard form polar form AC optimal power flow (AC OPF) problems.
  \item<1-> The condensed-space interior point method is employed.
  \item<1-> Solved up to $10^{-4}$ precision (note that by default, Ipopt solves up to $10^{-8}$ precision).
  \item<2-> For large-scale cases, GPU becomes {\bf significantly faster than CPU } \visible<3->{{\color{blue}(up to $\times 10$ speedup).}}
    \vspace{.1in}
    \begin{center}
      \scriptsize
      \input{./include/table.tex}\\
      AD = Automatic Differentiation\\
      $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.\\
      $^*$NVIDIA Quadro GV100  $^{**}$Intel Xeon Gold 6140
    \end{center}
  \item<4-> Automatic differentiation is $\times 300$ faster, and the rest of computation is $\times 8$ faster. 
  \end{itemize}
  \reference{
    S. Shin, F. Pacaud, and M. Anitescu. {\it Accelerating optimal power flow with GPUs: SIMD abstraction of nonlinear programs and condensed-space interior-point methods}, arXiv:2307.16830.
  }
\end{frame}

\begin{frame}{Highlight: Security-Constrained AC OPF {\normalsize (multi GPU)}}
  \begin{itemize}
  \item<1-> Standard security-constrained AC optimal power flow (SC AC OPF) problems.
  \item<1-> The {\bf reduction \& dense formulation} and the {\bf classical Schur-complement method for block-structured problems} are employed.
  \item<2-> Leveraging {\bf multiple GPUs} allows us to {\bf overcome the limitations} of a single GPU.
    \vspace{.1in}
    \begin{center}
      \scriptsize
      \input{./include/multi.tex}\\
      AD = Automatic Differentiation, LA = Linear Algebra (KKT System Solver)\\
      $^{*}$AMD CPU Epyc ”Milan” $^{**}$NVIDIA A100 (4 GPUs per node) 
    \end{center}
  \end{itemize}
  \reference{
    F. Pacaud, M. Schanen, S. Shin, D. A. Maldonado, and M. Anitescu. {\it Parallel interior-point solver for block-structured nonlinear programs on SIMD/GPU architectures}. arXiv:2301.04869.
  }
\end{frame}

\begin{frame}{Highlight: Security-Constrained AC OPF {\normalsize (multi GPU)}}
  \vspace{.1in} 
  \begin{itemize}
  \item<1-> Our approach is $\times 6$ faster than commercial multi-CPU solver.
    \begin{center}
      \includegraphics[height=.45\textheight]{kkt_timings.pdf}
    \end{center}
  \item<2-> Single-node, multi-GPU parallelism is more efficient than multi-node, multi-GPU.
    \begin{center}
      \includegraphics[height=.3\textheight]{nvlink.png}
    \end{center}
  \end{itemize}
  \reference{
    F. Pacaud, M. Schanen, S. Shin, D. A. Maldonado, and M. Anitescu. {\it Parallel interior-point solver for block-structured nonlinear programs on SIMD/GPU architectures}. arXiv:2301.04869.
  }
\end{frame}

% \begin{frame}{On-going Works}
%   \begin{itemize}
%   \item Other projects: HiOp (HPC nonlinear optimization solver) and HyKKT (KKT system solver)
%   \item Running nonlinear optimization on Frontier (AMD GPUs) and Aurora (Intel GPUs).
%   \item Preconditioned iterative methods for solving KKT systems.
%   \item Specialzied solvers for nonlinear optimal control.
%   \item Have an idea? Happy to chat!
%   \end{itemize}
% \end{frame}

\begin{frame}{Conclusions}
\begin{itemize}
  \item Modern GPU hardware offers significant potential for accelerating large-scale optimization.
  \item Implementing algorithms on GPUs can be challenging and often {\bf requires reevaluating key aspects of existing algorithms}.
  \item We have achieved promising results: up to \textbf{10x faster solutions with moderate accuracy}, though challenges remain for high-accuracy solutions.
  \item GPU-based optimization opens up opportunities to address previously intractable problems:
    \begin{itemize}
    \item {\bf Extremely large-scale} problems (multi-stage, multiscale).
    \item Problems involving {\bf expensive surrogate models} (neural nets, simulations).
    \end{itemize}
  \end{itemize}
  \visible<2->{
    \begin{center}
      \tikz{
        \node[rounded corners, fill=red!10,text width=.85\textwidth,align=center,inner sep=5] at (0,0) {
          {\bf Shin Group @ MIT is hiring postdocs}! Find more details here:\\ 
          \url{https://shin.mit.edu}
        };
      }
    \end{center}
  }
\end{frame}

\begin{frame}{Acknowledgement}
  \nocite{shin2022accelerating}
  \nocite{pacaud2023parallel} 
  \nocite{pacaud2023accelerating}
  \nocite{anitescu2021targeting}
  \begin{center}
    \tikz[align=center,font=\footnotesize]{
      \node at (0,0) {
        \bf U.S. Department of Energy, Office of Science, \\
        \bf Advanced Scientific Computing Research\\
        (DE-AC02-06CH11347)
      };
      \node at (7,0) {
        \bf U.S. Department of Energy, Office of Science\\
        \bf and National Nuclear Security Administration,\\
        {\bf Exascale Computing Porject} (17-SC-20-SC)
      };
      \node at (0,-1.5) {
        \includegraphics[height=.2\textheight]{doe.png}
      };
      \node at (7,-1.5) {
        \includegraphics[height=.2\textheight]{ecp.png}
      };
    }
  \end{center}
  {
    \footnotesize
    \bibliography{main}
  }
\end{frame}

\begin{frame}[noframenumbering,plain]
  \mytitlepage
\end{frame}

\end{document}

### 2024-02-19 paper/main.tex
Addressing comments

@@
 fixed pivot sequences.
+
 
@@
 specialized AD implementation based on the SIMD abstraction of
-NLPs. This abstraction allows us to preserve the parallelizable structure
-within the model equations, facilitating efficient derivative
-evaluations on the GPU. The AC power flow model is particularly
-well suited for this abstraction because it involves repetitive expressions
-for each component type (e.g., buses, lines, generators) and the
-the number of computational patterns does not increase with the network's
-size. Numerical results reported in this paper demonstrate that our
-proposed strategies can achieve over 20 times speedup by running on
-the GPU. Compared with general AD implementations on CPUs (such as AMPL
-and JuMP.jl), our GPU-based differentiation method can be
-approximately 500 times faster.
+NLPs. This abstraction allows us to preserve the parallelizable
+structure within the model equations, facilitating efficient
+derivative evaluations on the GPU. The AC power flow model is
+particularly well suited for this abstraction because it involves
+repetitive expressions for each component type (e.g., buses, lines,
+generators) and the the number of computational patterns does not
+increase with the network's size. {\color{blue} In other words, the
+objective and constraints can be expressed in the form of iterators.}
+Numerical results reported in this paper (Table \ref{tbl:results})
+demonstrate that our proposed {\color{blue} AD strategy} can achieve
+over 20 times speedup by running on the GPU. Compared with general AD
+implementations on CPUs (such as AMPL and JuMP.jl), our GPU-based
+differentiation method can be approximately 500 times faster.
 
@@
 \ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
-system. The initial symbolic factorization is performed on the CPU by using the KLU
-package \cite{davis2010algorithm}, and the subsequent numerical
+system via Cholesky factorization. The numerical
 factorization and triangular solves are performed by cuSOLVER with the
-fixed pivot sequence from KLU.  Software and hardware details of each
+fixed pivot sequence obtained with an approximate minimum degree ordering algorithm \cite{amestoyApproximateMinimumDegree1996}, implemented in AMD.jl \cite{montoisonAMDJlJulia2020}.
+Software and hardware details of each
 configuration are illustrated in Table \ref{tbl:settings}. The ACOPF
@@
 count is nearly the same as that of Ipopt for achieving the same level
-of accuracy in the final solution (see Table \ref{tbl:quality}). This
+of accuracy in the final solution (see Table \ref{tbl:quality}).
+{\color{blue} Note that
+the constraint violation is not strictly less than the tolerance in all configurations
+as relative constraint violation is used as convergence criteria.}
+This
 suggests that running mathematically equivalent operations on GPUs (by
@@
 in the model. When comparing ExaModels.jl running on CPUs and GPUs, we
-observe a further speedup on the GPU of up to almost 20x for large
-instances (e.g., case24465\_goc); remarkably, this is 500 times faster
+observe a further speedup on the GPU of up to 10x for large
+instances (e.g., case30000\_goc); remarkably, this is 300 times faster
 than JuMP.jl. This demonstrates that adopting SIMD abstraction and
@@
 linear solvers is only moderate, this has a high impact on the overall
-speedup because the linear solver time constitutes a significant portion of
-the total solution time. For large instances, approximately 4x speedup
-can be achieved, although this can vary depending on the instances. For
-example, for case24464\_goc, the GPU linear solver was slower than the
-CPUs. The investigation of under which circumstances cuSOLVER is
-more effective warrants further research.
+speedup because the linear solver time constitutes a significant
+portion of the total solution time. {\color{blue} This can be observed from Figure
+\ref{fig:speedup}; linear algebra is the only bottleneck in
+large-scale instances.  For large instances, GPUs can only achieve
+approximately 5x speedup.  These speedup ratios can vary depending on
+the instances.  For instance, for case24464\_goc, the GPU linear
+solver performance was close to that of the CPU solver.  The
+investigation of under which circumstances cuSOLVER is more effective
+warrants further research.}
 
@@
 large-scale instances, particularly when the number of variables is
-greater than 20,000. Notably, for the largest instance,
+greater than 20,000. {\color{blue} This is because the benefit of parallelization
+is greater for operations over large data.}
+Notably, for the largest instance,
 case30000\_goc, our GPU implementation is 4 times faster than our CPU
@@
 framework that can run entirely on GPUs. Our method has demonstrated
-promising results, achieving a 4x speedup when compared with CPU
+promising results, achieving a 5x speedup when compared with CPU
 implementations for large-scale ACOPF problems. Notably, our approach
 outperforms one of the state-of-the-art CPU-based implementations by
-a factor of 10. These results, along with our packages
+a factor of 12. These results, along with our packages
 MadNLP.jl and ExaModels.jl, showcase a significant advancement in our

### 2024-02-19 paper/main.tex
highlight added

@@
 The adoption of graphics processing units (GPUs) in the mathematical programming community has remained
-limited. Notably, nonlinear programming (NLP) remains dependent on
+limited due to the challenges associated with parallelizing the optimization algorithms. Notably, nonlinear programming (NLP) remains dependent on
 algorithms developed in the 1990s offering limited room for parallelism.
@@
 fixed pivot sequences.
-
 
@@
 demonstrate that our proposed {\color{blue} AD strategy} can achieve
-over 20 times speedup by running on the GPU. Compared with general AD
+over {\color{blue} 10} times speedup by running {\color{blue} the AD} on the GPU. Compared with general AD
 implementations on CPUs (such as AMPL and JuMP.jl), our GPU-based
@@
 with JuMP.jl) by an order of magnitude. This finding underscores the
-importance of harnessing the power of GPUs for tackling
+importance of harnessing the power of GPUs to tackle
 the computational challenges in power systems.
@@
 the treatment of the KKT systems of NLPs without numerical pivoting,
-thus allowing the solution of sparse, large-scale NLPs (with a
-prominent example being ACOPFs) on GPUs.
+thus allowing the solution of sparse, large-scale NLPs (with a prominent example being ACOPFs) on GPUs.
 
@@
 augmented Lagrangian outer iteration and the treatment of linear
-systems using a preconditioned conjugate gradient method. Prior to the
+systems using a preconditioned conjugate gradient method. Before the
 introduction of the sparse condensed-space IPM with an inequality relaxation
@@
 applying the determined step size and direction. This process is
-repeated until  satisfaction of the convergence criteria (typically based on the residual
+repeated until the satisfaction of the convergence criteria (typically based on the residual
 to the first-order optimality conditions).
@@
 Implementing kernel functions for the above computational patterns in
-the Julia language is straightforward, since Julia provides excellent
+the Julia language is straightforward since Julia provides excellent
 high-level interfaces for array and kernel programming for GPU
@@
 \subsection{Sparsity Analysis}
-The sparsity analysis is needed for determining the sparsity pattern
+The sparsity analysis is needed to determine the sparsity pattern
 of the evaluated derivatives. In the case of large-scale sparse
@@
 be expensive, since the sparsity should be analyzed for potentially
-millions of objective and constraints terms. Often, however, 
+millions of objective and constraint terms. Often, however, 
 these analyses are applied to the same computational patterns, and
@@
 \ref{config-4} are used as benchmarks. \ref{config-1} and
-\ref{config-2} share significant amount of code, especially the
+\ref{config-2} share a significant amount of code, especially the
 high-level abstractions, but they differ in how they handle the KKT
@@
 the constraint violation is not strictly less than the tolerance in all configurations
-as relative constraint violation is used as convergence criteria.}
+as relative (scaled) constraint violation is used for the convergence criteria.}
 This
@@
 speedup because the linear solver time constitutes a significant
-portion of the total solution time. {\color{blue} This can be observed from Figure
-\ref{fig:speedup}; linear algebra is the only bottleneck in
-large-scale instances.  For large instances, GPUs can only achieve
-approximately 5x speedup.  These speedup ratios can vary depending on
-the instances.  For instance, for case24464\_goc, the GPU linear
-solver performance was close to that of the CPU solver.  The
-investigation of under which circumstances cuSOLVER is more effective
-warrants further research.}
+portion of the total solution time. {\color{blue} Figure
+\ref{fig:speedup} reveals that linear algebra is the only
+computational bottleneck for large-scale instances. While derivative
+evaluation and solver internal computation could achieve a 10x speedup,
+the linear algebra part can only achieve a 5x speedup even for the
+largest case. For case24464\_goc, the GPU linear solver performance
+was close to that of the CPU solver.  The investigation of under which
+circumstances cuSOLVER is more effective warrants further research.}
 
@@
 ability to reveal the inertia of the KKT system. We are
-interested in exploring other linear solvers options, such as CHOLMOD
+interested in exploring other linear solver options, such as CHOLMOD
 \cite{chen2008algorithm}, Baspacho \cite{pineda2022theseus}, and HyKKT
